---
title: "Welcome to the mizer blog"
author: "Gustav Delius"
date: "8/3/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


There is now a sizeable and diverse community of mizer users. This blog, I hope,
will help keep us informed of what we are all up to and to share our experiences
and our best practices.

I will use this blog from time to time to explain new or less well known
features of mizer. But I would like to encourage you to contribute to this blog
too. Introduce us to your model or to your research. Or discuss ways in which
you have extended mizer or would like to extend mizer. Or show us new ways of
using mizer. All you need to do is create an R Markdown document containing your
text and your code and either email it to me at <mizer@sizespectrum.org> or, if
you are more of a GitHub wizard, create a pull request directly to the
repository at (https://github.com/sizespectrum/mizerBlog).

This blog is using [distill for R Markdown](https://rstudio.github.io/distill/),
which gives it some nice features. The one that
I like the most is that any code blocks that you include in your post will 
automatically be highlighted and linked to documentation pages. For example
if your R Markdown document contains a code block with

<pre>
library(mizer)
plotlySpectra(NS_params)
</pre>

then what will be displayed on the blog is

```{r}
library(mizer)
plotlySpectra(NS_params)
```

Note how the function names and package names have become links. Click on them
to see what they do. If you are using packages or functions that your readers
are not familiar with yet, this autolinking allows them to look them up very
conveniently.

Note that code highlighting and autolinking currently does not work for
inline code. This is due to a 
[know bug in distill](https://github.com/rstudio/distill/issues/186)
and will hopefully be fixed soon.

The output of the `plotlySpectra()` function is an interactive plot 
and this got embedded into the blog automatically. Play around
with it a bit. It behaves just as it would if you viewed it in your 
RStudio viewer. 

As the blog grows, we can start using categories to organise the posts.
The distill framework provides all such standard blog features. One
feature I have enabled already are comments, using the Disqus service.

