---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/"
)
```

# mizer
[![CRAN Status](https://www.r-pkg.org/badges/version/mizer)](https://cran.r-project.org/package=mizer)
[![Travis-CI Build Status](https://travis-ci.org/sizespectrum/mizer.svg?branch=master)](https://travis-ci.org/sizespectrum/mizer)
[![Coverage status](https://codecov.io/gh/sizespectrum/mizer/branch/master/graph/badge.svg)](https://codecov.io/github/sizespectrum/mizer?branch=master)
[![CRAN Downloads](http://cranlogs.r-pkg.org/badges/grand-total/mizer)](https://cran.r-project.org/package=mizer)
[![Rdoc](http://www.rdocumentation.org/badges/version/mizer)](http://www.rdocumentation.org/packages/mizer)

Mizer is an R package to run multi-species size-spectrum models of fish
communities. The package has been developed to model marine ecosystems that are
subject to fishing. However, it may also be appropriate for other ecosystems.

The package contains routines and functions to allow users to set up an ecosystem 
model, and then project it through time under different fishing strategies.
Methods are included to explore the results, including plots and calculation of
community indicators such as the slope of the size spectrum. Size-based models
can be complicated so mizer contains many default options that can be easily
changed by the user.

<!-- Mizer can also be used to create web apps that allow users to explore models -->
<!-- without the need to install R. An [example of such an -->
<!-- app](https://mizer.shinyapps.io/selectivity/) investigates the effect of -->
<!-- switching to a gear with a T90 extension net to reduce the catches of undersize -->
<!-- hake and red mullet -->

Mizer has been supporting research in marine ecology and fisheries science 
since 2014 
([see publications](https://sizespectrum.org/mizer/articles/publications.html). 
It is still under active
development. The new version 2.0 has increased the user-friendliness and the
flexibility of the framework. Contributions from the user community are very
welcome. There is a sister package called 
[mizerExperimental](https://sizespectrum.org/mizerExperimental) where user
contributions can be checked out and receive feedback from the community. 
Example mizer models can be contributed to 
[mizerExamples](https://sizespectrum.org/mizerExamples).

Does your project or publication use mizer? If so, we would love to know. You can
also join our Google Discussion group here:
https://groups.google.com/forum/#!forum/size-spectrum-models

Recent work on mizer was funded by the European
Commission Horizon 2020 Research and Innovation Programme under Grant Agreement
No 634495 for the project MINOUW (http://minouw-project.eu/) and the Australian
Research Council Discovery Project [Rewiring Marine Food Webs](http://marinesocioecology.org/rewiring-marine-food-webs/).



## Installation

The package is on [CRAN](https://cran.r-project.org/package=mizer) and therefore
available from R's built-in package manager.
```{r, eval = FALSE}
# Install release version from CRAN
install.packages("mizer")

# Alternatively, install development version from GitHub
devtools::install_github("sizespectrum/mizer")
```

## Example
The following code loads the mizer package, loads some information about species
in the North Sea that comes as an example with the package, sets up the 
parameters for the mizer model, and runs a simulation for 10 years.
```{r, message = FALSE}
library(mizer)
params <- newMultispeciesParams(NS_species_params, inter)
sim <- project(params, t_max = 10, effort = 0)
```
The results of the simulation can then be analysed, for example via plots:
```{r}
plot(sim)
```

See the accompanying
[Get started](https://sizespectrum.org/mizer/articles/mizer.html) page
for more details on how the package works, including detailed examples.
