---
title: "Estimating feeding kernel parameters from stomach data"
output: html_notebook
---

We want to use stomach data to estimate the feeding preference
function for a size-spectrum model. To make a connection we 
assume that the abundance of a prey size in a predator's stomach
is proportional to the rate at which the predator feeds on prey of 
that size. This is a dubious assumption because it neglects that
prey of different sizes might stay identifiable in the stomach for
different lengths of time.

For a predator of size $w'$ the rate at which it predates on
prey of size $w$ is proportional to the product of the abundance
$N(w)$ of prey of size $w$ and the preference function $\phi(w/w')$.
The proportionality constant (the so-called search volume) depends 
only on the known predator weight and is independent the prey weight
and therefore does not concern us here.

We assume that the prey abundance is given by a power law 
$$N(w)\propto w^{-\lambda}.$$
For the preference function we use the Gaussian
$$\phi(w/w') = \exp\left(-\frac{(\log(w'/w)-\log(\beta))^2}
  {2\sigma^2}\right).$$
We are following the notation used by mizer here. Note the natural log
around $\beta$ but the lack of a log around $\sigma$.

The probability density for the weight of a random prey individual
in the stomach of a predator of weight $w'$ is therefore
$$ f_w(w|w') \propto w^{-\lambda}
\exp\left(-\frac{(\log(w'/w)-\log(\beta))^2}
  {2\sigma^2}\right).$$
Mariella likes to work in terms of the predator/prey mass ratio $r=w'/w$.
The probability density for this ration can be obtained by observing that
$$ f_r(r|w') = f_w(w|w')\left|\frac{dw}{dr}\right|=\frac{w'}{r^2} f_w(w|w')$$
and thus
$$ f(r|w') \propto r^{\lambda-2}
\exp\left(-\frac{(\log(r)-\log(\beta))^2}
  {2\sigma^2}\right).$$
The probability density function
for the log of the predator/prey mass ratio $l = \log(r)$ is then obtained by observing that
$$ f_l(l|w') = f_r(r|w')\frac{dr}{dl}=r f_r(r|w').$$
This gives
$$ f_{l}(l|w') \propto e^{l(\lambda-1)}
\exp\left(-\frac{(l-\log(\beta))^2}
  {2\sigma^2}\right).$$

By combining the exponentials and completing squares and then 
normalising, the density becomes
$$ f_{l}(l|w') = 
\frac{1}{\sqrt{2\pi}\sigma}
\exp\left(-\frac{(l-(\log(\beta)+\sigma^2(\lambda-1))^2}
  {2\sigma^2}\right),$$

Thus the logarithm of the predator/prey mass ratio is normally distributed with
mean $\log(\beta)+\sigma^2(\lambda-1)$ and variance $\sigma^2$.


By estimating the mean $\mu$ and standard deviation $s$ of the observed
logarithms of predator/prey mass ratios we
obtain estimates for $\beta$ and $\sigma$ by the method of moments as
$$\beta = \exp(\mu-s^2(\lambda - 1)),~~~~ \sigma^2 = s^2.$$

We load Mariella's data and choose $\lambda$.
```{r message=FALSE, warning=FALSE}
library(readr)
library(dplyr)
params_data <- read_csv("speciesNCME_Mariella.csv")
lambda <- 2
```
Mariella told me that what she calls sigma in this file is $e^s$,
the exponential of the standard deviation in the log of the observed 
predator/prey ratios. So we obtain mizer's sigma as
```{r}
params_data$sigma_mizer <- log(params_data$sigma)
```
Mariella also told me that what she calls beta in this file is $e^\mu$,
the exponential of the mean of the log of the observed predator/prey ratios.
We convert that to mizer's beta as
```{r}
params_data$beta_mizer <- params_data$beta * 
    exp(-params_data$sigma_mizer^2 * (lambda - 1))
```
This gives
```{r}
select(params_data, c("species", "beta_mizer", "sigma_mizer", "beta", "sigma"))
```
We notice that for species that have a very large standard deviation in the log 
of the predator/prey mass ratio, namely jack mackerel, mackerel, sardine and 
anchovy, the values for $\beta$ are ridiculously small and even less than 1. 
Clearly that makes no sense. We have to give up either the assumption of
a Gaussian feeding kernel or the assumption that feeding rate is proportional
to stomach content.

One proposal would be to assume that the time for which a prey item stays
identifiable in the stomach scales allometrically with prey size. One could
for example argue that the rate at which digestion takes place on an
individual scales as its surface area, i.e., as $w^{2/3}$. Then the time
until a certain percentage of the individual has been digested scales as
$w/w^{2/3} = w^{1/3}$. Thus the abundance in the stomach of identifiable
individuals of size $w$ would be proportional to $w^{1/3}$ times the
rate at which the predator feeds on these prey items. This would modify
the probability density for the size of identifiable prey items to
$$ f_w(w|w') \propto w^{-\lambda+1/3}
\exp\left(-\frac{(\log(w'/w)-\log(\beta))^2}
  {2\sigma^2}\right).$$
The log of the preadator/prey mass ratios for identifiable prey items would 
then be normally distributed with mean $\log(\beta)+\sigma^2(\lambda-4/3)$
and the parameter $\beta$ would be estimated as
$$\beta = \exp(\mu-s^2(\lambda - 4/3)).$$
```{r}
params_data$beta_mizer <- params_data$beta * 
    exp(-params_data$sigma_mizer^2 * (lambda - 4/3))
```
This gives
```{r}
select(params_data, c("species", "beta_mizer", "sigma_mizer", "beta", "sigma"))
```
This improves things a bit but not enough. If we are willing to increase the exponent for the time till
digestion to $2/3$ we get:
```{r}
params_data$beta_mizer <- params_data$beta * 
    exp(-params_data$sigma_mizer^2 * (lambda - 5/3))
select(params_data, c("species", "beta_mizer", "sigma_mizer", "beta", "sigma"))
```

Next we analyse the stomach data for swordfish.
```{r}
ppmr <- read.csv("Swordfish_dataPPMR.csv")
```


```{r}
mu <- mean(log(ppmr$Predator.mass / ppmr$Prey.mass))
s <- sd(log(ppmr$Predator.mass / ppmr$Prey.mass))
library(ggplot2)
```
```{r}
ggplot(ppmr) +
    geom_density(aes(log(Predator.mass / Prey.mass))) +
    stat_function(fun = dnorm, args = list(mean = mu, sd = s), color = "blue")
```
```{r}
ggplot(ppmr) +
    geom_density(aes(log(Predator.mass / Prey.mass))) +
    stat_function(fun = dnorm, args = list(mean = mu, sd = s), color = "blue")
```


Next we analyse the stomach data for anchovy and sardine directly.
```{r message=FALSE}
ppmr <- read_csv("Data_PPMR.csv") %>%
    group_by(Species)
est <- ppmr %>%
    summarise(totalNprey = sum(Nprey),
              mean = sum(log(wpredator/wprey) * Nprey) / sum(Nprey), 
              var = sum((log(wpredator/wprey) - mean)^2 * Nprey) / (sum(Nprey) - 1),
              sigma = sqrt(var),
              log_beta = mean - var * (lambda - 1),
              beta = exp(log_beta),
              log10_beta = log10(beta)
    )
est
```
