---
title: "Cheatsheet: Changing Model Parameters"
output:
  html_document:
    toc: yes
    toc_float: yes
    fig_width: 5
    fig_height: 4
---

```{r include=FALSE}
library(mizer)
params <- NS_params
```

Mizer offers several ways to change a model, and it is easy to be unsure which
one to reach for. This cheatsheet explains the distinctions and gives a rule for
each. The guiding principle:

> **Change the model at the highest level that expresses your intent, and let
> mizer propagate the change downwards.** Only drop to a lower level when you
> deliberately want to override mizer's calculations.

---

## The levels of a mizer model

A mizer model is built in layers. Almost every change you make is really a choice
of *which layer* to reach into:

| Level | What it is | Change it with |
|---|---|---|
| **1. Size-independent parameters** | the high-level inputs: per-species scalars (`w_inf`, `beta`, `gamma`, `h`, `erepro`, …), fishing gears, resource scalars, and interactions | [`given_species_params(params) <-`](../reference/species_params.html), [`gear_params(params) <-`](../reference/gear_params.html), [`resource_params(params) <-`](../reference/resource_params.html), [`setInteraction()`](../reference/setInteraction.html) |
| **2. Size-dependent parameters** | arrays over size that mizer **calculates** from the parameters (search volume, metabolic rate, predation kernel, resource capacity, selectivity, …) | the [`set…()`](../reference/setParams.html) functions |
| **3. Rate functions** | the functions mizer calls to compute the rates during a simulation | [`setRateFunction()`](../reference/setRateFunction.html) |

**Levels 1 and 2 cover almost all everyday work.** The parameters at level 1 are
the highest level: most of them are used to **calculate** the size-dependent rate
arrays at level 2. Changing a parameter therefore normally updates the
corresponding array automatically — you rarely need to touch the arrays yourself.
This is what the guiding principle above means by propagating a change downwards.
Species parameters, gear parameters and resource parameters all sit together at
the top level: each is a set of scalar inputs from which mizer builds the arrays
below.

**Levels 3** is for going beyond mizer's built-in
behaviour. With [`setRateFunction()`](../reference/setRateFunction.html) you
replace the function mizer uses to compute a rate (for example a bespoke mortality
or reproduction rate), and with [`setComponent()`](../reference/setComponent.html)
you add a whole new ecosystem component — a detritus pool, a carrion pool, an
extra predator — with its own dynamics. These are advanced and are documented on
their own reference pages rather than in this cheatsheet.

---

## Level 1: Size-independent parameters

A size-structured model like mizer potentially has a huge number of parameters
because for example mortality rates or predation rates need to be specified
at every size. Luckily, mizer by default assumes some sensible functional
forms for such size-dependent parameters, so that you only need to specify
a small number of parameters that are then used to calculate the size-dependent
parameters.

There are four types of size-independent parameters: species parameters, gear
parameters, resource parameters and the interaction matrix. For a full
discussion of how mizer uses size-independent rates to set up the size-dependent
rates you can read the help page of [`setRates()`](../reference/setRates.html)

### Species parameters

There are many species-specific parameters in mizer. Luckily, mizer sets or
calculates defaults for those that you do not specify explicitly. 

#### `given` vs `calculated` vs all

Mizer
keeps track of which species parameter values you **gave** and which ones it
**calculated** (from your values or from defaults).

| Accessor | Returns |
|---|---|
| [`given_species_params(params)`](../reference/species_params.html) | only the parameters you supplied explicitly |
| [`calculated_species_params(params)`](../reference/species_params.html) | the parameters mizer derived or defaulted |
| [`species_params(params)`](../reference/species_params.html) | everything (given, with calculated filling the gaps) |

There are two replacement functions, and the difference matters:

- **`given_species_params(params) <- …`** — the one to use. It records your value
  as *given* and **triggers recalculation** of everything that depends on it
  (both the calculated scalar parameters and the size-dependent rate arrays).

    ```{r eval=FALSE}
    given_species_params(params)$beta <- 150     # updates the predation kernel too
    ```

- **`species_params(params) <- …`** — writes the value into the combined table
  **without** recalculating the derived scalar parameters, and your value may be
  **silently overwritten** the next time a recalculation is triggered. Avoid it
  unless you know you want this.

**Rule:** to change a species parameter, use `given_species_params(params) <-`.

Access columns with `$` (they come back as named vectors):

```{r eval=FALSE}
species_params(params)$w_mat        # named by species
given_species_params(params)$gamma  # only if you gave it (NA or NULL otherwise)
```

---

#### How a change propagates

Many species parameters exist only to *set up* a size-dependent rate. When you
change one with `given_species_params(params) <-`, mizer re-runs the relevant
setter for you:

| Species parameter(s) | Sets up | via |
|---|---|---|
| `gamma`, `q` | search volume | [`setSearchVolume()`](../reference/setSearchVolume.html) |
| `h`, `n` | maximum intake rate | [`setMaxIntakeRate()`](../reference/setMaxIntakeRate.html) |
| `k`, `ks`, `p` | metabolic rate | [`setMetabolicRate()`](../reference/setMetabolicRate.html) |
| `z0`, `z_ext`, `d` | external mortality | [`setExtMort()`](../reference/setExtMort.html) |
| `beta`, `sigma`, `pred_kernel_type` | predation kernel | [`setPredKernel()`](../reference/setPredKernel.html) |
| `w_mat`, `w_mat25`, `w_repro_max`, `m` | reproduction allocation | [`setReproduction()`](../reference/setReproduction.html) |

Some species parameters are instead used **directly** by the model and set up no
array — changing them just changes the model: `alpha` (assimilation), `w_min`
(egg size), `erepro` and `R_max` (reproduction), `interaction_resource`, and the
length–weight parameters `a`, `b`.

For a fuller description of the species parameters see the help page
of [`species_params()`](../reference/species_params.html).

---

### Fishing gear parameters

Fishing gears, selectivity, and catchability live in a separate table,
[`gear_params()`](../reference/gear_params.html), one row per gear–species pair.
Assigning to it recomputes the fishing arrays with
[`setFishing()`](../reference/setFishing.html). See the **Fishing** cheatsheet for
details.

```{r eval=FALSE}
gp <- gear_params(params)
gp["Cod, Otter", "catchability"] <- 0.8
gear_params(params) <- gp
```

The related setter [`setFishing()`](../reference/setFishing.html) is for supplying
selectivity/catchability **arrays** directly or setting the baseline effort.

---

### Resource parameters

The resource works just like the species parameters.
[`resource_params()`](../reference/resource_params.html) returns a named list of
scalars (`kappa`, `lambda`, `r_pp`, `n`, `w_pp_cutoff`) that set up the resource
size-spectrum arrays, and — like `given_species_params<-` — assigning to it
**rebuilds those arrays** by calling [`setResource()`](../reference/setResource.html)

```{r eval=FALSE}
resource_params(params)$kappa  <- 1e11   # rescales the carrying capacity
resource_params(params)$lambda <- 2.05   # changes the slope of the capacity
resource_params(params)$r_pp   <- 10     # rebuilds the replenishment rate
```

| Scalar(s) | Rebuilds |
|---|---|
| `kappa`, `lambda`, `w_pp_cutoff` | resource carrying capacity (`cc_pp`) |
| `r_pp`, `n` | resource replenishment rate (`rr_pp`) |

You can also set the size-resolved arrays directly. As with the species rate
arrays, doing so **freezes** them ("set manually"), so a later scalar change no
longer touches them:

| Function | Sets (and freezes) |
|---|---|
| [`resource_capacity(params) <-`](../reference/setResource.html) | the carrying capacity over size |
| [`resource_rate(params) <-`](../reference/setResource.html) | the replenishment (regeneration) rate over size |
| [`resource_level(params) <-`](../reference/setResource.html) | the resource level (fraction of capacity) |
| [`setResource(params, …)`](../reference/setResource.html) | any of the above, plus `lambda`, `n`, `w_pp_cutoff`, `resource_dynamics` |

```{r eval=FALSE}
resource_capacity(params) <- my_capacity          # array over size; now frozen
resource_params(params)$kappa <- 1e11             # ignored: capacity is frozen
params <- setResource(params, reset = TRUE)       # unfreeze: recompute from the scalars
```

This mirrors the species-parameter behaviour exactly: scalars are the "given"
inputs, the arrays are calculated from them unless you override an array by hand.

---

### The interaction matrix

The species × species interaction matrix is set with
[`setInteraction()`](../reference/setInteraction.html) and read with
`getInteraction()`. (The resource interaction is in the `interaction_resource`
species-parameter column instead.)

```{r eval=FALSE}
inter <- getInteraction(params)
inter["Cod", "Herring"] <- 0.5
params <- setInteraction(params, inter)
```

---

## Level 2: Size-dependent rates

Everything above changes level-1 parameters and lets mizer rebuild the level-2
arrays for you. This section is about reaching *into* level 2 directly if you
need a different size-dependence from the one mizer assumes by default.

Each size-dependent parameter is stored as an array and has its own **setter
function** — `metab(params) <-`, `search_vol(params) <-`, and so on — that
writes an array straight into the model. (The matching getter, e.g.
`metab(params)`, reads it back.) This is equivalent to passing the array to the
corresponding `set…()` function:

```{r eval=FALSE}
metab(params) <- my_array                             # direct setter
params <- setMetabolicRate(params, metab = my_array)  # create new model
```

Note how the `set...()` functions return a new `MizerParams` object while the
direct setter functions like `metab()<-` make the change in place, i.e.,
modify the passed `MizerParams` object.

Either way you **freeze** that array: mizer marks it as "set manually" and will
**no longer update it** when you change the species parameters that would
otherwise feed it. This is intentional — it lets you impose a bespoke rate — but
it means, for example, that after setting `search_vol` by hand, changing `gamma`
has no effect on the search volume. To hand control back to mizer, recompute the
array from the parameters:

```{r eval=FALSE}
params <- setSearchVolume(params, reset = TRUE)   # drop the manual override, recompute
```

(Calling `setSearchVolume(params)` *without* `reset = TRUE` will **not** recompute
a frozen array — mizer leaves a manually set array in place until you explicitly
reset it. Each of the size-dependent rate setters below takes this `reset`
argument, as do `setResource()` and `setFishing()`.)

The full set of size-dependent rate arrays, each with its direct setter/getter
and the `set…()` function that recomputes it:

| Rate array | Direct setter / getter | Recompute with |
|---|---|---|
| search volume | `search_vol(params) <-` | [`setSearchVolume()`](../reference/setSearchVolume.html) |
| maximum intake rate | `intake_max(params) <-` | [`setMaxIntakeRate()`](../reference/setMaxIntakeRate.html) |
| metabolic rate | `metab(params) <-` | [`setMetabolicRate()`](../reference/setMetabolicRate.html) |
| external mortality | `ext_mort(params) <-` | [`setExtMort()`](../reference/setExtMort.html) |
| external encounter rate | `ext_encounter(params) <-` | [`setExtEncounter()`](../reference/setExtEncounter.html) |
| external diffusion | `ext_diffusion(params) <-` | [`setExtDiffusion()`](../reference/setExtDiffusion.html) |
| predation kernel | `pred_kernel(params) <-` | [`setPredKernel()`](../reference/setPredKernel.html) |
| maturity ogive | `maturity(params) <-` | [`setReproduction()`](../reference/setReproduction.html) |
| reproduction allocation | `repro_prop(params) <-` | [`setReproduction()`](../reference/setReproduction.html) |

The fishing arrays (selectivity, catchability) and the resource arrays behave the
same way but are covered in the [Fishing](#fishing-parameters) and
[Resource](#resource-parameters) sections above.

---

## Level 3: Replacing a rate function

Levels 1 and 2 change the *numbers* mizer uses; level 3 changes the *code*. At
every time step of a [`project()`](../reference/project.html) run, mizer calls a
set of functions to compute the instantaneous rates (encounter, feeding level,
mortality, reproduction, …). [`setRateFunction()`](../reference/setRateFunction.html)
swaps any one of these for a function of your own, without touching the species
parameters or the rate arrays:

```{r eval=FALSE}
params <- setRateFunction(params, "Mort", "myMort")   # use myMort() for total mortality
getRateFunction(params)                               # list the current rate functions
```

The first argument names the rate to replace. The available names are the
components mizer computes internally: `Encounter`, `FeedingLevel`,
`PredRate`, `PredMort`, `FMort`, `Mort`, `EReproAndGrowth`, `ERepro`,
`EGrowth`, `Diffusion`, `ResourceMort`, `RDI`, `RDD` — and `Rates` itself, if
you need to replace the whole bundle. Each has a default `mizer…()`
implementation (e.g. [`mizerMort()`](../reference/mizerMort.html)) that you can
call, wrap, or ignore.

Your function must accept the same arguments as the mizer function it replaces
and **return an array of the same shape**. The signature always includes the
current simulation time `t`:

```{r eval=FALSE}
myRate <- function(params, n, n_pp, n_other, t, ...) { … }
```

### Time-dependent rates

This is the key reason to reach for `setRateFunction()`: **species parameters and
rate arrays are fixed for the whole simulation, but a rate *function* receives
the current time `t` and can therefore change as the simulation runs.** That lets
you express things the parameters cannot — seasonal forcing, a warming trend, a
management measure that switches on in a given year, and so on.

For example, to give total mortality an annual cycle (with `t` measured in
years), wrap the default `mizerMort()` and scale its result:

```{r eval=FALSE}
seasonalMort <- function(params, t, ...) {
    mizerMort(params, t = t, ...) * (1 + 0.3 * sin(2 * pi * t))
}
params <- setRateFunction(params, "Mort", "seasonalMort")
```

or to step fishing-independent mortality up permanently from year 30 onwards:

```{r eval=FALSE}
regimeShift <- function(params, t, ...) {
    factor <- if (t >= 30) 1.5 else 1
    mizerMort(params, t = t, ...) * factor
}
params <- setRateFunction(params, "Mort", "regimeShift")
```

Two practical points:

- **Extra parameters** your function needs go in
  [`other_params(params)`](../reference/setRateFunction.html), e.g.
  `other_params(params)$warming_rate <- 0.02`, and are then reachable inside the
  function as `params@other_params$warming_rate`.
- Your functions must be defined in the **global environment or in a package** —
  mizer cannot find a function defined inside another function.

For the full mechanism, including adding new ecosystem components with
[`setComponent()`](../reference/setComponent.html), see
`vignette("extending-mizer")`.

---

## Which should I use?

| I want to change… | Use |
|---|---|
| a per-species value (`beta`, `w_mat`, `h`, `erepro`, …) | `given_species_params(params) <- …` |
| fishing gears / selectivity / catchability | `gear_params(params) <- …` |
| the resource (kappa, lambda, r_pp, …) | `resource_params(params) <- …` |
| species interactions | `setInteraction(params, …)` |
| a size-dependent rate, keeping it tied to the parameters | change the underlying parameter (above) |
| a size-dependent rate to a bespoke array (freezing it) | the direct setter `metab(params) <- …` (or the matching `set…()` with the array argument) |
| a size-dependent rate back to default form | call matching `set...()` with `reset = TRUE`
| how a rate is *computed* (e.g. to make it time-dependent) | `setRateFunction(params, …)` |

---

## Quick reference

```{r eval=FALSE}
# ── Species parameters (preferred: triggers recalculation) ────────────────────
given_species_params(params)$beta <- 150
species_params(params)                     # view all (given + calculated)
calculated_species_params(params)          # only what mizer derived

# ── Size-dependent rate arrays ────────────────────────────────────────────────
params <- setMetabolicRate(params)         # recompute from species params
metab(params) <- my_array                  # direct setter; freezes the array
params <- setSearchVolume(params, search_vol = my_array)  # same, via set...()
params <- setSearchVolume(params, reset = TRUE)  # unfreeze: recompute again
params <- setParams(params)                # rebuild ALL rate arrays

# ── Fishing ───────────────────────────────────────────────────────────────────
gear_params(params) <- gp
params <- setFishing(params, initial_effort = c(Otter = 1))

# ── Resource (scalars rebuild the arrays, like given_species_params) ──────────
resource_params(params)$kappa <- 1e11      # rescales the carrying capacity
resource_capacity(params) <- my_capacity   # bespoke array; freezes it
params <- setResource(params, reset = TRUE)  # unfreeze: recompute from scalars

# ── Interaction ───────────────────────────────────────────────────────────────
params <- setInteraction(params, inter)

# ── Rate functions (level 3: replace how a rate is computed) ──────────────────
seasonalMort <- function(params, t, ...) {
    mizerMort(params, t = t, ...) * (1 + 0.3 * sin(2 * pi * t))   # t in years
}
params <- setRateFunction(params, "Mort", "seasonalMort")  # now time-dependent
other_params(params)$my_param <- 42                        # params for your function
```
