---
title: "Cheatsheet: Changing Model Parameters"
output:
  html_document:
    toc: yes
    toc_float: yes
    fig_width: 5
    fig_height: 4
---

```{r include=FALSE}
library(mizer)
params <- NS_params
```

Mizer offers several ways to change a model, and it is easy to be unsure which
one to reach for. This cheatsheet explains the distinctions and gives a rule for
each. The guiding principle:

> **Change the model at the highest level that expresses your intent, and let
> mizer propagate the change downwards.** Only drop to a lower level when you
> deliberately want to override mizer's calculations.

Every setter returns a **new** `MizerParams` object, so always reassign the
result (`params <- setResource(params, ...)`).

---

## The three levels of parameters

| Level | What it holds | Change it with |
|---|---|---|
| **Species parameters** | per-species scalars (`w_inf`, `beta`, `gamma`, `h`, `erepro`, …) | [`given_species_params(params) <-`](../reference/species_params.html) |
| **Size-dependent rates** | arrays over size (search volume, metabolic rate, predation kernel, …) | the [`set…()`](../reference/setParams.html) functions |
| **Other groups** | fishing, resource, interaction | [`gear_params()`](../reference/gear_params.html), [`resource_params()`](../reference/resource_params.html), [`setInteraction()`](../reference/setInteraction.html) |

Species parameters are the highest level: most of them are used to **calculate**
the size-dependent rate arrays. Changing a species parameter therefore normally
updates the corresponding rate array automatically — you rarely need to touch the
arrays yourself.

---

## Species parameters: `given` vs `calculated` vs all

Mizer separates the species parameters you **gave** from the ones it
**calculated** (from your values or from defaults).

| Accessor | Returns |
|---|---|
| [`given_species_params(params)`](../reference/species_params.html) | only the parameters you supplied explicitly |
| [`calculated_species_params(params)`](../reference/species_params.html) | the parameters mizer derived or defaulted |
| [`species_params(params)`](../reference/species_params.html) | everything (given, with calculated filling the gaps) |

There are two replacement functions, and the difference matters:

- **`given_species_params(params) <- …`** — the one to use. It records your value
  as *given* and **triggers recalculation** of everything that depends on it
  (both the calculated scalar parameters and the size-dependent rate arrays).

    ```{r eval=FALSE}
    given_species_params(params)$beta <- 150     # updates the predation kernel too
    ```

- **`species_params(params) <- …`** — writes the value into the combined table
  **without** recalculating the derived scalar parameters, and your value may be
  **silently overwritten** the next time a recalculation is triggered. Avoid it
  unless you know you want this.

**Rule:** to change a species parameter, use `given_species_params(params) <-`.

Access columns with `$` (they come back as named vectors):

```{r eval=FALSE}
species_params(params)$w_mat        # named by species
given_species_params(params)$gamma  # only where you set it (NA elsewhere)
```

---

## How a change propagates

Many species parameters exist only to *set up* a size-dependent rate. When you
change one with `given_species_params(params) <-`, mizer re-runs the relevant
setter for you:

| Species parameter(s) | Sets up | via |
|---|---|---|
| `gamma`, `q` | search volume | [`setSearchVolume()`](../reference/setSearchVolume.html) |
| `h`, `n` | maximum intake rate | [`setMaxIntakeRate()`](../reference/setMaxIntakeRate.html) |
| `k`, `ks`, `p` | metabolic rate | [`setMetabolicRate()`](../reference/setMetabolicRate.html) |
| `z0`, `z_ext`, `d` | external mortality | [`setExtMort()`](../reference/setExtMort.html) |
| `beta`, `sigma`, `pred_kernel_type` | predation kernel | [`setPredKernel()`](../reference/setPredKernel.html) |
| `w_mat`, `w_mat25`, `w_repro_max`, `m` | reproduction allocation | [`setReproduction()`](../reference/setReproduction.html) |

Some species parameters are instead used **directly** by the model and set up no
array — changing them just changes the model: `alpha` (assimilation), `w_min`
(egg size), `erepro` and `R_max` (reproduction), `interaction_resource`, and the
length–weight parameters `a`, `b`.

---

## Size-dependent rates: the `set…()` functions

Each size-dependent rate array has its own setter. Called with only a `params`
object, a setter **recomputes** the array from the current species parameters:

```{r eval=FALSE}
params <- setMetabolicRate(params)   # recompute from k, ks, p
```

[`setParams(params)`](../reference/setParams.html) runs *all* of these setters at
once (interaction, predation kernel, intake, metabolism, mortality, encounter,
diffusion, search volume, reproduction). Use it to rebuild everything after
several changes.

### Overriding an array by hand (and the consequence)

You can pass an array directly to a setter to set a rate **manually**:

```{r eval=FALSE}
params <- setSearchVolume(params, search_vol = my_array)  # species × size
```

Doing so **freezes** that array: mizer marks it as "set manually" and will
**no longer update it** when you change the species parameters that would
otherwise feed it. This is intentional — it lets you impose a bespoke rate — but
it means, for example, that after setting `search_vol` by hand, changing `gamma`
has no effect on the search volume. To hand control back to mizer, recompute the
array from the parameters:

```{r eval=FALSE}
params <- setSearchVolume(params)    # drop the manual override, recompute
```

---

## Fishing parameters

Fishing gears, selectivity, and catchability live in a separate table,
[`gear_params()`](../reference/gear_params.html), one row per gear–species pair.
Assigning to it recomputes the fishing arrays. See the **Fishing** cheatsheet for
details.

```{r eval=FALSE}
gp <- gear_params(params)
gp["Cod, Otter", "catchability"] <- 0.8
gear_params(params) <- gp
```

The related setter [`setFishing()`](../reference/setFishing.html) is for supplying
selectivity/catchability **arrays** directly or setting the baseline effort.

---

## Resource parameters

The resource works just like the species parameters.
[`resource_params()`](../reference/resource_params.html) returns a named list of
scalars (`kappa`, `lambda`, `r_pp`, `n`, `w_pp_cutoff`) that set up the resource
size-spectrum arrays, and — like `given_species_params<-` — assigning to it
**rebuilds those arrays**:

```{r eval=FALSE}
resource_params(params)$kappa  <- 1e11   # rescales the carrying capacity
resource_params(params)$lambda <- 2.05   # changes the slope of the capacity
resource_params(params)$r_pp   <- 10     # rebuilds the replenishment rate
```

| Scalar(s) | Rebuilds |
|---|---|
| `kappa`, `lambda`, `w_pp_cutoff` | resource carrying capacity (`cc_pp`) |
| `r_pp`, `n` | resource replenishment rate (`rr_pp`) |

You can also set the size-resolved arrays directly. As with the species rate
arrays, doing so **freezes** them ("set manually"), so a later scalar change no
longer touches them:

| Function | Sets (and freezes) |
|---|---|
| [`resource_capacity(params) <-`](../reference/setResource.html) | the carrying capacity over size |
| [`resource_rate(params) <-`](../reference/setResource.html) | the replenishment (regeneration) rate over size |
| [`resource_level(params) <-`](../reference/setResource.html) | the resource level (fraction of capacity) |
| [`setResource(params, …)`](../reference/setResource.html) | any of the above, plus `lambda`, `n`, `w_pp_cutoff`, `resource_dynamics` |

```{r eval=FALSE}
resource_capacity(params) <- my_capacity          # array over size; now frozen
resource_params(params)$kappa <- 1e11             # ignored: capacity is frozen
params <- setResource(params, reset = TRUE)       # unfreeze: recompute from the scalars
```

This mirrors the species-parameter behaviour exactly: scalars are the "given"
inputs, the arrays are calculated from them unless you override an array by hand.

---

## The interaction matrix

The species × species interaction matrix is set with
[`setInteraction()`](../reference/setInteraction.html) and read with
`getInteraction()`. (The resource interaction is the `interaction_resource`
species-parameter column instead.)

```{r eval=FALSE}
inter <- getInteraction(params)
inter["Cod", "Herring"] <- 0.5
params <- setInteraction(params, inter)
```

---

## Which should I use?

| I want to change… | Use |
|---|---|
| a per-species value (`beta`, `w_mat`, `h`, `erepro`, …) | `given_species_params(params) <- …` |
| a size-dependent rate, keeping it tied to the parameters | change the underlying species parameter (above) |
| a size-dependent rate to a bespoke array (freezing it) | the matching `set…()` with the array argument |
| everything at once after edits | `setParams(params)` |
| fishing gears / selectivity / catchability | `gear_params(params) <- …` |
| baseline fishing effort or selectivity arrays | `setFishing(params, …)` |
| the resource (kappa, lambda, r_pp, …) | `resource_params(params) <- …` |
| the resource capacity/rate as a bespoke array (freezing it) | `resource_capacity(params) <- …` / `resource_rate(params) <- …` / `setResource(params, …)` |
| species interactions | `setInteraction(params, …)` |

---

## Quick reference

```{r eval=FALSE}
# ── Species parameters (preferred: triggers recalculation) ────────────────────
given_species_params(params)$beta <- 150
species_params(params)                     # view all (given + calculated)
calculated_species_params(params)          # only what mizer derived

# ── Size-dependent rate arrays ────────────────────────────────────────────────
params <- setMetabolicRate(params)         # recompute from species params
params <- setSearchVolume(params, search_vol = my_array)  # freeze a manual array
params <- setSearchVolume(params)          # unfreeze: recompute again
params <- setParams(params)                # rebuild ALL rate arrays

# ── Fishing ───────────────────────────────────────────────────────────────────
gear_params(params) <- gp
params <- setFishing(params, initial_effort = c(Otter = 1))

# ── Resource (scalars rebuild the arrays, like given_species_params) ──────────
resource_params(params)$kappa <- 1e11      # rescales the carrying capacity
resource_capacity(params) <- my_capacity   # bespoke array; freezes it
params <- setResource(params, reset = TRUE)  # unfreeze: recompute from scalars

# ── Interaction ───────────────────────────────────────────────────────────────
params <- setInteraction(params, inter)
```
