---
title: "Cheatsheet: Model Setup and Calibration"
output:
  html_document:
    toc: yes
    toc_float: yes
    fig_width: 5
    fig_height: 4
---

```{r include=FALSE}
library(mizer)
params <- NS_params
```

This cheatsheet gives a quick overview of the functions used to **build** a
mizer model, bring it to **steady state**, **calibrate** it to observations, and
**project** it forward. For full documentation of each function, follow the
links.

Every setter function returns a **new** `MizerParams` object, so always reassign
the result (`params <- steady(params)`). Change species parameters through
[`species_params(params) <-`](../reference/species_params.html) so that
dependent quantities are recalculated.

---

## Creating a model

Choose a constructor for the type of model you need.

| Function | Model type |
|---|---|
| [`newSingleSpeciesParams()`](../reference/newSingleSpeciesParams.html) | one species in a fixed background |
| [`newCommunityParams()`](../reference/newCommunityParams.html) | single size spectrum, no species identity |
| [`newTraitParams()`](../reference/newTraitParams.html) | several species differing only in asymptotic size |
| [`newMultispeciesParams()`](../reference/newMultispeciesParams.html) | fully general multi-species model |

Most work uses `newMultispeciesParams()`, driven by a **species parameter data
frame** — one row per species. Only two columns are required:

| Column | Meaning |
|---|---|
| `species` | species name |
| `w_inf` | von Bertalanffy asymptotic weight (g) — the required maximum-size parameter |

`w_max` (the computational grid boundary) defaults to `1.5 * w_inf`; `w_mat`,
`beta`, `sigma`, `h`, `gamma`, `alpha`, `erepro`, `R_max`, … all have defaults
or are calculated. Weights are in **grams**, lengths in **cm**, time in
**years**.

```{r eval=FALSE}
species_params <- read.csv(
    system.file("extdata", "NS_species_params.csv", package = "mizer"))
params <- newMultispeciesParams(species_params)
```

Useful optional arguments to `newMultispeciesParams()`:

| Argument | Effect |
|---|---|
| `interaction` | species × species matrix of overlaps in `[0, 1]` (default all 1) |
| `gear_params` | fishing gear definitions (see the fishing cheatsheet) |
| `no_w`, `min_w`, `max_w` | size-grid resolution and range (`no_w = 100` default) |

Inspect the result with [`summary(params)`](../reference/summary.html),
[`species_params(params)`](../reference/species_params.html),
[`getInteraction(params)`](../reference/getInteraction.html), and
[`gear_params(params)`](../reference/gear_params.html).

---

## Finding the steady state

A freshly constructed model has only a rough spectrum. Settle it onto a steady
state, which also sets the initial values used by calibration and `project()`.

| Function | Use |
|---|---|
| [`steady(params)`](../reference/steady.html) | run the dynamics to convergence with births held fixed (the default) |
| [`projectToSteady(params)`](../reference/projectToSteady.html) | version with births responding dynamically; exposes `t_max`, `return_sim` |
| [`steadySingleSpecies(params)`](../reference/steadySingleSpecies.html) | set each species to its single-species steady form with births held fixed (fast starting point) |

```{r eval=FALSE}
params <- steady(params)
```

During model setup and calibration you almost always want `steady()` or
`steadySingleSpecies()` because keeping births constant lets the dynamics
  converge reliably onto *a* steady state. Afterwards these functions re-tune the
  reproduction parameters so that density-dependent reproduction reproduces
  exactly that birth rate at the new steady state — use `preserve` to choose
  whether `reproduction_level` (default), `R_max`, or `erepro` is held fixed
  during that re-tuning.

---

## Calibrating to observations

Supply observations in the species-parameter columns `biomass_observed` and/or
`yield_observed` (optionally with `biomass_cutoff` / `yield_cutoff` size
thresholds). Then run the calibration loop, **re-running `steady()` after any
match/calibrate step**:

| Function | Adjusts | To match |
|---|---|---|
| [`calibrateBiomass(params)`](../reference/calibrateBiomass.html) | `kappa` (resource level) | total community biomass |
| [`matchBiomasses(params)`](../reference/matchBiomasses.html) | per-species abundance | each `biomass_observed` |
| [`calibrateYield(params)`](../reference/calibrateYield.html) | overall abundance scale | total community yield |
| [`matchGrowth(params)`](../reference/matchGrowth.html) | `h`, `gamma`, `ks`, `k` | growth to `w_mat` / `w_inf` |

```{r eval=FALSE}
params <- calibrateBiomass(params)   # total biomass
params <- matchBiomasses(params)     # per species
params <- matchGrowth(params)        # growth
params <- steady(params)             # re-converge
```

`matchGrowth()` and `matchBiomasses()` pull on different parameters; alternate
them, re-running `steady()` between, until both are satisfied.

---

## When calibration misbehaves

- **A species collapses to near-zero.** Its mortality exceeds the growth it can
  fund. Check its predation-kernel parameters `beta` and `sigma`, its row of the
  interaction matrix, and whether its fishing mortality is too high.
- **`steady()` will not settle.** The initial spectrum is probably far from the
  steady state. Run
  [`steadySingleSpecies(params)`](../reference/steadySingleSpecies.html) first to
  get a sensible starting spectrum, or take a smaller step in whatever parameter
  you changed and re-run.

Diagnose the fit with `plotBiomassObservedVsModel(params)` and
`plotGrowthCurves(params)` — see [Verifying the model](#verifying-the-model).

---

## Density-dependent reproduction

[`setBevertonHolt()`](../reference/setBevertonHolt.html) sets how strongly
reproduction is density-limited. `reproduction_level` is the fraction of maximum
recruitment realised at steady state (0 = density independent, closer to 1 =
strongly limited).

```{r eval=FALSE}
params <- setBevertonHolt(params, reproduction_level = 0.25)
```

You can instead pass `R_max`, `erepro`, or a per-species named vector.

---

## Projecting forward

[`project()`](../reference/project.html) runs the model and returns a
`MizerSim`. See the analysis-and-plotting cheatsheet for exploring the result.

```{r eval=FALSE}
sim <- project(params, t_max = 20, effort = 1)
```

| Argument | Meaning |
|---|---|
| `effort` | scalar, per-gear vector, or time × gear array (see fishing cheatsheet) |
| `t_max` | number of years to simulate |
| `dt` | integration time step (reduce if a run is unstable) |
| `t_save` | interval at which output is stored |
| `method` | `"euler"` (default), `"predictor_corrector"`, `"tr_bdf2"` |

Pass a `MizerSim` back to `project()` to continue from where it ended.

---

## Verifying the model

```{r eval=FALSE}
plotSpectra(params)                    # sensible, overlapping spectra?
plotGrowthCurves(params, species = "Cod")
plotBiomassObservedVsModel(params)     # points near the 1:1 line?
plotYieldObservedVsModel(params)
```

---

## Quick reference

```{r eval=FALSE}
# ── Build ─────────────────────────────────────────────────────────────────────
params <- newMultispeciesParams(species_params, interaction)
params <- newTraitParams()          # or newCommunityParams(), newSingleSpeciesParams()

# ── Steady state ──────────────────────────────────────────────────────────────
params <- steady(params)
params <- steadySingleSpecies(params)   # fast starting spectrum

# ── Calibrate to data (re-run steady() after each) ────────────────────────────
params <- calibrateBiomass(params)      # total biomass  → kappa
params <- matchBiomasses(params)        # per-species biomass
params <- calibrateYield(params)        # total yield
params <- matchGrowth(params)           # growth → h, gamma, ks, k

# ── Reproduction ──────────────────────────────────────────────────────────────
params <- setBevertonHolt(params, reproduction_level = 0.25)

# ── Project ───────────────────────────────────────────────────────────────────
sim <- project(params, t_max = 20, effort = 1)

# ── Verify ────────────────────────────────────────────────────────────────────
plotBiomassObservedVsModel(params)
plotGrowthCurves(params)
```
