---
title: "Cheatsheet: Fishing"
output:
  html_document:
    toc: yes
    toc_float: yes
    fig_width: 5
    fig_height: 4
---

```{r include=FALSE}
library(mizer)
params <- NS_params
```

This cheatsheet gives a quick overview of how fishing is set up in mizer:
**gears**, **selectivity**, **catchability**, and **effort**. For full
documentation of each function, follow the links.

Fishing mortality is
$$F_{g,i}(w) = S_{g,i}(w)\, Q_{g,i}\, E_g,$$
the product of gear $g$'s **selectivity** $S_{g,i}(w)$ for species $i$ at size $w$, its **catchability** for
species $i$, and its **effort**. Selectivity and catchability are configured through
the [`gear_params`](../reference/gear_params.html) data frame; effort is set at
run time.

---

## The gear parameter data frame

One row per **gear–species combination** (a gear catching three species has
three rows). Access it with [`gear_params(params)`](../reference/gear_params.html).
Required columns:

| Column | Meaning |
|---|---|
| `gear` | gear name |
| `species` | species this row applies to |
| `sel_func` | name of the selectivity function |
| `catchability` | scales `F` for this gear–species pair (default 1) |

Plus one column per parameter of the chosen `sel_func` (see below). Row names
follow the pattern `"species, gear"`.

**Editing an existing gear table.** Pull it out, change what you need, and assign
it back — the assignment triggers recalculation of the selectivity and
catchability arrays:

```{r eval=FALSE}
gp <- gear_params(params)
gp["Cod, Otter", "catchability"] <- 0.8
gear_params(params) <- gp                 # assignment triggers recalculation
```

**Setting one up from scratch.** Assign a fresh data frame with one row per
gear–species combination. Only `species` is strictly required: `gear` defaults
to the species name, `sel_func` to `knife_edge`, `catchability` to 1, and the
`knife_edge` cut-off to `w_mat`. You must, however, supply the parameter columns
of whatever `sel_func` you choose. Here two gears fish cod, each with its own
length-based selectivity:

```{r eval=FALSE}
gear_params(params) <- data.frame(
    gear         = c("Otter", "Beam"),
    species      = c("Cod",   "Cod"),
    sel_func     = "sigmoid_length",       # recycled to both rows
    l50          = c(25, 20),              # 50% selected at this length (cm)
    l25          = c(20, 15),              # 25% selected at this length (cm)
    catchability = 1
)
```

This replaces the whole gear table; mizer generates the `"species, gear"` row
names for you.

If each species is caught by only one gear, the gear columns may instead be
supplied within `species_params` when building the model; mizer copies them into
`gear_params`. Later edits to those `species_params` columns do **not**
propagate — edit `gear_params` after construction.

---

## Selectivity functions

Each selectivity function takes `w` as its first argument and returns a value in
`[0, 1]` at each size. Its other arguments must appear as columns in
`gear_params`.

| `sel_func` | Parameter column(s) | Shape |
|---|---|---|
| [`knife_edge`](../reference/knife_edge.html) (default) | `knife_edge_size` | step from 0 to 1 (default size `w_mat`) |
| [`sigmoid_length`](../reference/sigmoid_length.html) | `l50`, `l25` | smooth; lengths (cm) at 50% and 25% selection |
| [`double_sigmoid_length`](../reference/double_sigmoid_length.html) | `l50`, `l25`, `l50_right`, `l25_right` | dome-shaped (selects a length band) |
| [`sigmoid_weight`](../reference/sigmoid_weight.html) | `sigmoidal_weight`, `sigmoidal_sigma` | smooth transition in weight |

`sigmoid_length` is the most commonly used. You can also supply your own
function (first argument `w`, returns selectivity at size) and name it in
`sel_func`.

```{r eval=FALSE}
gp <- gear_params(params)
gp$sel_func <- "sigmoid_length"
gp$l50 <- 25      # 50% selected at 25 cm
gp$l25 <- 20      # 25% selected at 20 cm
gear_params(params) <- gp
```

---

## The selectivity and catchability arrays

Behind the scenes mizer turns the `gear_params` table into two numeric arrays,
the ones that enter the fishing-mortality formula directly. You can read them,
and — when a `sel_func` cannot express the shape you need — set them by hand.

| Function | Returns | Dimensions |
|---|---|---|
| [`catchability(params)`](../reference/setFishing.html) / `getCatchability(params)` | $Q_{g,i}$ | gear × species |
| [`selectivity(params)`](../reference/setFishing.html) / `getSelectivity(params)` | $S_{g,i}(w)$ | gear × species × size |

The bare and `get`-prefixed names are equivalent; use whichever reads better.
Each has a matching setter that pushes an array straight into the model (this
routes through `setFishing()`, so validation still runs):

```{r eval=FALSE}
catchability(params)                       # gear × species matrix of Q
selectivity(params)["Otter", "Cod", ]      # the S curve for one gear–species pair

# Assign a custom selectivity curve that no sel_func produces
sel <- getSelectivity(params)
sel["Otter", "Cod", ] <- my_curve          # length = number of size bins, in [0, 1]
selectivity(params) <- sel                 # triggers recalculation via setFishing()
```

A direct assignment here is for shapes you cannot obtain from a selectivity
function. Once you set an array by hand, mizer marks it as manual and stops
recalculating it from `gear_params` — so later edits to the gear table leave your
array untouched (you'll see a message saying so). To discard the hand-set array
and rebuild from `gear_params`, call `setFishing(params, reset = TRUE)`.

---

## Fishing effort

The model stores a **baseline effort** per gear, used when `project()` is called
without an explicit `effort` argument.

| Function | Use |
|---|---|
| [`initial_effort(params)`](../reference/initial_effort.html) | read baseline effort (a named vector) |
| [`initial_effort(params) <-`](../reference/initial_effort.html) | set baseline effort |
| [`getEffort(sim)`](../reference/getEffort.html) | effort actually used over time in a simulation |

```{r eval=FALSE}
initial_effort(params) <- c(Industrial = 0, Pelagic = 1, Beam = 0.5, Otter = 0.5)
```

At run time, [`project()`](../reference/project.html) accepts `effort` in four
forms:

```{r eval=FALSE}
project(params, effort = 1)                        # scalar: all gears, constant
project(params, effort = c(Otter = 0.5, Beam = 1)) # named vector: per gear, constant
project(params, effort = c(0.5, 1, 0, 0.5))        # vector in gear order, constant
project(params, effort = effort_array)             # time × gear array: through time
```

For a time-varying scenario, build a `time × gear` array with numeric,
increasing row names and gear column names:

```{r eval=FALSE}
gears <- names(getInitialEffort(params))
years <- 2010:2030
effort_array <- array(1, dim = c(length(years), length(gears)),
                      dimnames = list(time = years, gear = gears))
effort_array[as.character(2020:2030), "Otter"] <- 1.5   # ramp one gear from 2020
sim <- project(params, effort = effort_array)
```

---

## Inspecting the fishing setup

```{r eval=FALSE}
gear_params(params)        # the gear table
catchability(params)       # Q array (gear × species)
selectivity(params)        # S array (gear × species × size)
initial_effort(params)     # baseline effort per gear
plotFMort(params)          # realised fishing mortality at size
getFMort(params)           # F by species × size
getFMortGear(params)       # F by gear × species × size
getYieldGear(sim)          # yield by gear (from a MizerSim)
```

---

## Quick reference

```{r eval=FALSE}
# ── Gears and selectivity ─────────────────────────────────────────────────────
gp <- gear_params(params)
gp$sel_func <- "sigmoid_length"
gp$l50 <- 25; gp$l25 <- 20
gp$catchability <- 1
gear_params(params) <- gp

# ── Effort ────────────────────────────────────────────────────────────────────
initial_effort(params) <- c(Otter = 0.5, Beam = 1)   # baseline
sim <- project(params, effort = 1)                   # constant during run
sim <- project(params, effort = effort_array)        # time × gear array

# ── Inspect ───────────────────────────────────────────────────────────────────
initial_effort(params)
plotFMort(params)
getFMortGear(params)
```
