---
title: "Cheatsheet: Analysis and Plotting"
output:
  html_document:
    toc: yes
    toc_float: yes
    fig_width: 5
    fig_height: 4
---

```{r include=FALSE}
library(mizer)
sim <- NS_sim
params <- NS_params
```

This cheatsheet gives a quick overview of the functions available in mizer for
analysing the results of simulations and creating plots. For full documentation
of each function, follow the links.

Most functions accept either a `MizerSim` object (returning a **time series**)
or a `MizerParams` object (returning a **single value** from the initial state).

---

## Accessing simulation arrays

These functions extract raw arrays from a `MizerSim` object.

| Function | Returns | Dimensions |
|---|---|---|
| [`N(sim)`](../reference/N.html) | species abundance density | time × species × size |
| [`NResource(sim)`](../reference/NResource.html) | resource abundance density | time × size |
| [`finalN(sim)`](../reference/finalN.html) | species abundance at last time | species × size |
| [`finalNResource(sim)`](../reference/finalNResource.html) | resource abundance at last time | size |
| [`getEffort(sim)`](../reference/getEffort.html) | fishing effort | time × gear |
| [`getTimes(sim)`](../reference/getTimes.html) | saved time steps | vector |

**Example:** Pull out Cod abundance:
```{r eval=FALSE}
N(sim)[, "Cod", ]          # time × size for Cod
N(sim)["2010", "Cod", ]    # size vector for Cod in year 2010
finalN(sim)["Cod", ]       # size vector for Cod at the final time step
```

---

## Summary functions

These compute derived quantities from abundances. All accept `MizerSim` or
`MizerParams`. See [`?summary_functions`](../reference/summary_functions.html)
for the full list.

| Function | Returns | Dimensions |
|---|---|---|
| [`getBiomass(sim, min_w, max_w)`](../reference/getBiomass.html) | total biomass | time × species |
| [`getSSB(sim)`](../reference/getSSB.html) | spawning stock biomass | time × species |
| [`getN(sim, min_w, max_w)`](../reference/getN.html) | total abundance | time × species |
| [`getYield(sim)`](../reference/getYield.html) | total yield across gears | time × species |
| [`getYieldGear(sim)`](../reference/getYieldGear.html) | yield by gear | time × gear × species |
| [`getFeedingLevel(sim)`](../reference/getFeedingLevel.html) | feeding level at size | time × species × size |
| [`getPredMort(sim)`](../reference/getPredMort.html) | predation mortality at size | time × species × size |
| [`getFMort(sim)`](../reference/getFMort.html) | fishing mortality at size | time × species × size |
| [`getFMortGear(sim)`](../reference/getFMortGear.html) | fishing mortality by gear | time × gear × species × size |
| [`getDiet(params)`](../reference/getDiet.html) | diet resolved by prey at size | predator × size × prey |
| [`getTrophicLevel(params)`](../reference/getTrophicLevel.html) | trophic level at size | species × size |
| [`getTrophicLevelBySpecies(params)`](../reference/getTrophicLevelBySpecies.html) | mean trophic level per species | species |

**Size range:** `getBiomass()` and `getN()` accept `min_w`, `max_w`, `min_l`,
`max_l` to restrict the calculation to a size range.

**Example:**
```{r eval=FALSE}
getSSB(sim)                              # SSB of all species over time
getBiomass(sim, min_w = 10, max_w = 1e4) # biomass of 10g–10kg fish
getYield(sim)["2010", ]                  # yield in year 2010
```

The result is an `ArrayTimeBySpecies` (time × species) or `ArraySpeciesBySize`
(species × size), which can be plotted directly with `plot()` — see
[Plotting any array directly](#plotting-any-array-directly-with-plot) below.

---

## Indicator functions

These compute community-level indicators. All accept `MizerSim` (time series)
or `MizerParams` (single value from initial state). See
[`?indicator_functions`](../reference/indicator_functions.html).

| Function | Key arguments | Returns |
|---|---|---|
| [`getProportionOfLargeFish(sim)`](../reference/getProportionOfLargeFish.html) | `threshold_w = 100`, `biomass_proportion` | proportion of large fish through time |
| [`getMeanWeight(sim)`](../reference/getMeanWeight.html) | `min_w`, `max_w`, `species` | mean community weight through time |
| [`getMeanMaxWeight(sim)`](../reference/getMeanMaxWeight.html) | `measure = "both"/"numbers"/"biomass"` | mean asymptotic weight through time |
| [`getCommunitySlope(sim)`](../reference/getCommunitySlope.html) | `min_w`, `max_w`, `species` | slope, intercept, R² through time |

**Example:**
```{r eval=FALSE}
lfi <- getProportionOfLargeFish(sim, min_w = 10, max_w = 5000, threshold_w = 500)
lfi[c("1972", "2010")]

slope <- getCommunitySlope(sim, min_w = 10, max_w = 5000)
head(slope)
```

---

## Plotting functions

All plotting functions return a `ggplot2` object that you can customise further
(see [Working with ggplot2](#working-with-ggplot2)). See
[`?plotting_functions`](../reference/plotting_functions.html).

The rest of this section starts with the general mechanism — calling `plot()`
directly on any array, and the arguments that control it — and only then
describes the dedicated `plot...()` functions, which are mostly shortcuts for it.

### Plotting any array directly with `plot()`

The arrays returned by the summary and rate functions carry a mizer array class
and have their own `plot()` method, so you can plot any quantity without a
dedicated plot function.

| Class | Typical source | `plot()` shows |
|---|---|---|
| [`ArrayTimeBySpecies`](../reference/ArrayTimeBySpecies.html) | `getBiomass(sim)`, `getSSB(sim)`, `getYield(sim)`, `getN(sim)` | value vs time, one line per species |
| [`ArraySpeciesBySize`](../reference/ArraySpeciesBySize.html) | `getFeedingLevel(params)`, `getPredMort(params)`, `getEncounter(params)` | value vs size, one line per species |
| [`ArrayTimeBySpeciesBySize`](../reference/ArrayTimeBySpeciesBySize.html) | `getFMort(sim)`, `getPredMort(sim)` | one time slice vs size (set with `time`) |
| [`ArrayResourceBySize`](../reference/ArrayResourceBySize.html) | `NResource(params)`, `getResourceMort(params)`, `resource_rate(params)`, `resource_capacity(params)` | resource quantity vs size |

```{r eval=FALSE}
plot(getBiomass(sim))          # value vs time, one line per species
plot(getSSB(sim))
plot(getFeedingLevel(params))  # value vs size, one line per species
plot(getResourceMort(params))  # plankton resource mortality vs size
```

The array plots come with a small toolkit for combining and comparing them:

```{r eval=FALSE}
# Add another compatible array as extra lines on an existing plot
p <- plot(getBiomass(sim), species = "Cod")
addPlot(p, getBiomass(sim), species = "Herring", linetype = "dashed")

# Compare two compatible arrays
plot2(getFMort(params), getFMort(params2), "Before", "After")
plotRelative(getEGrowth(params), getEGrowth(params2))  # relative difference

# Interactive (hover) version of any array plot
plotHover(getBiomass(sim))
```

| Function | What it does |
|---|---|
| [`addPlot()`](../reference/addPlot.html) | adds a compatible array as extra lines on an existing plot |
| [`plot2()`](../reference/plot2.html) | compares two compatible arrays (colour = species, linetype = which object) |
| [`plotRelative()`](../reference/plotRelative.html) | shows the relative difference between two compatible arrays |
| [`plotHover()`](../reference/plotHover.html) | turns any of these ggplots into a hover-enabled plotly plot |

### Common arguments

Most analysis and plotting functions — including `plot()` on an array and the
dedicated `plot...()` functions below — share these optional arguments:

| Argument | Effect |
|---|---|
| `species` | character vector — restrict to a subset of species |
| `time_range` | numeric vector — average over this time period (plots against size) |
| `tlim` | numeric vector `c(min, max)` — restrict time axis (plots against time) |
| `wlim`/`llim` | numeric vector `c(min, max)` — restrict the size (x) axis (plots against size) |
| `ylim` | numeric vector `c(min, max)` — restrict the value (y) axis |
| `highlight` | character vector — draw named species with thicker lines |
| `total` | logical — add a line for the community total |
| `log_x`, `log_y`, `log` | logical — log-scale the x or y axis |

`wlim`/`llim` (size axis) and `ylim` (value axis) only set the visible window:
data outside the range is hidden but nothing is recomputed. To change the
underlying numbers — for example the size range that a biomass is summed over —
pass `min_w`/`max_w` (or `min_l`/`max_l`) to the `get...()` function instead, e.g.
`plotBiomass(sim, min_w = 10)` restricts the calculation to fish above 10 g.

Which of these apply depends on the array's shape:

`plot(<ArrayTimeBySpecies>)` accepts: `species`, `tlim`, `total`,
`background`, `highlight`, `log_x`, `log_y`, `ylim`.

`plot(<ArraySpeciesBySize>)` accepts: `species`, `highlight`, `log_x`, `log_y`,
`wlim`, `ylim`, `all.sizes`.

### Dedicated plot functions

Each dedicated `plot...()` function is essentially `plot()` applied to the matching
`get...()` array, so `plotBiomass(sim)` is `plot(getBiomass(sim))` and
`plotFeedingLevel(sim)` is `plot(getFeedingLevel(sim))`. They accept the common
arguments above. The tables below note only where a function does something you
could **not** get by plotting the array directly.

Each also has a `plotly` counterpart (e.g. `plotlyBiomass()`) for interactive
use — the array `plot()`s use [`plotHover()`](../reference/plotHover.html) instead.

#### Against time

| Function | How it relates to plotting the array directly |
|---|---|
| [`plotBiomass(sim)`](../reference/plotBiomass.html) | same as `plot(getBiomass(sim))` |
| [`plotYield(sim)`](../reference/plotYield.html) | same as `plot(getYield(sim))` |
| [`plotYieldGear(sim)`](../reference/plotYieldGear.html) | like `plotYield()` but keeps the gear dimension, drawing one panel per fishing gear |

```{r eval=FALSE}
plotBiomass(sim, species = c("Cod", "Herring"), total = TRUE)
plotBiomass(sim, tlim = c(1980, 1990))
plotYield(sim, log_y = FALSE)
```

#### Against body size

By default these show the final time step; use `time_range` to average over a
period.

| Function | How it relates to plotting the array directly |
|---|---|
| [`plotFeedingLevel(sim)`](../reference/plotFeedingLevel.html) | same as `plot(getFeedingLevel(sim))` |
| [`plotPredMort(sim)`](../reference/plotPredMort.html) | same as `plot(getPredMort(sim))` |
| [`plotFMort(sim)`](../reference/plotFMort.html) | same as `plot(getFMort(sim))` |
| [`plotSpectra(sim)`](../reference/plotSpectra.html) | abundance/biomass spectra: additionally overlays the resource spectrum and background species, and `power` rescales the y axis |
| [`plotCDF(sim)`](../reference/plotCDF.html) | cumulative version of the spectrum (`normalise` for proportion vs total) |
| [`plotGrowthCurves(sim)`](../reference/plotGrowthCurves.html) | a distinct plot: size at age rather than a size spectrum |
| [`plotDiet(params)`](../reference/plotDiet.html) | a distinct plot: stacked diet composition by prey |

```{r eval=FALSE}
plotSpectra(sim, power = 2, time_range = 1990:2000)
plotFeedingLevel(sim, highlight = c("Cod", "Haddock"))
plotGrowthCurves(sim, species = "Cod", max_age = 20)
plotDiet(params, species = "Cod")
plotCDF(sim, power = 1)              # cumulative biomass; power = 0 for numbers
plotCDF(sim, normalise = FALSE)     # cumulative total rather than proportion
```

### Summary plot

```{r eval=FALSE}
plot(sim)      # 5-panel summary: feeding level, biomass, predation mort, fishing mort, spectra
plot(params)   # same panels for a model's steady state (no biomass-through-time panel)
```

### Comparing two simulations or models

These take two compatible `MizerSim` or `MizerParams` objects (e.g. before and
after a change) and show them together. For whole spectra use the dedicated
functions below; for any other rate array use `plot2()` and `plotRelative()`
from [Plotting any array directly](#plotting-any-array-directly-with-plot).

| Function | Shows |
|---|---|
| [`plotSpectra2(object1, object2, name1, name2)`](../reference/plotSpectra2.html) | two abundance spectra overlaid |
| [`plotSpectraRelative(object1, object2)`](../reference/plotSpectraRelative.html) | relative difference of two spectra |
| [`plotCDF2(object1, object2, name1, name2)`](../reference/plotCDF2.html) | two cumulative distributions overlaid |

```{r eval=FALSE}
plotSpectra2(params, params2, "Before", "After")
plotSpectraRelative(params, params2)         # 2 (N2 - N1) / (N1 + N2)
plotCDF2(sim, sim2, "Unfished", "Fished")
```

### Animating spectra through time

[`animate()`](../reference/animate.html) plays a spectrum or rate array through
the course of a simulation.

```{r eval=FALSE}
animate(sim)                 # abundance spectra over time
animate(getFMort(sim))       # an ArrayTimeBySpeciesBySize over time
```

---

### Working with ggplot2

All `plot...()` functions return a `ggplot2` object, so you can customise them:

```{r eval=FALSE}
library(ggplot2)
p <- plotBiomass(sim, species = c("Cod", "Herring"))
p + theme_bw() + labs(title = "Biomass through time")
p + geom_hline(aes(yintercept = 1e10), linetype = "dashed")
```

Change species colours and line types via the `MizerParams` object:

```{r eval=FALSE}
params <- setColours(params, c("Cod" = "darkblue"))
params <- setLinetypes(params, c("Cod" = "dashed"))
```

---

## Quick reference

```{r eval=FALSE}
# ── Accessing raw arrays ───────────────────────────────────────────────────────
N(sim)                  # time × species × size
NResource(sim)          # time × size
finalN(sim)             # species × size  (last time step)
finalNResource(sim)     # size            (last time step)

# ── Species biomass / abundance / yield (time × species) ──────────────────────
getBiomass(sim)         # total biomass
getSSB(sim)             # spawning stock biomass
getN(sim)               # total abundance (numbers)
getYield(sim)           # catch in weight
getYieldGear(sim)       # catch by gear (time × gear × species)

# ── Rates at size (time × species × size) ─────────────────────────────────────
getFeedingLevel(sim)    # satiation (0 = starving, 1 = full)
getPredMort(sim)        # predation mortality
getFMort(sim)           # fishing mortality
getFMortGear(sim)       # fishing mortality by gear (time × gear × species × size)

# ── Diet and trophic (species × size × …) ────────────────────────────────────
getDiet(params)                 # proportion of diet from each prey (predator × size × prey)
getTrophicLevel(params)         # trophic level at size (species × size)
getTrophicLevelBySpecies(params) # mean trophic level (species)

# ── Community indicators (time series) ────────────────────────────────────────
getProportionOfLargeFish(sim, threshold_w = 100)
getMeanWeight(sim)
getMeanMaxWeight(sim)
getCommunitySlope(sim)          # returns data.frame with slope, intercept, R²

# ── Dedicated plot functions ──────────────────────────────────────────────────
# Each plot*() is a shortcut for plot() on the matching get*() array, and each has
# an interactive plotly*() twin (plotlyBiomass(), plotlySpectra(), …).
plot(sim)               # 5-panel summary
plotBiomass(sim)        # biomass vs time
plotYield(sim)          # yield vs time
plotYieldGear(sim)      # yield vs time, faceted by gear
plotSpectra(sim)        # abundance spectra vs size (+ resource & background)
plotFeedingLevel(sim)   # feeding level vs size
plotPredMort(sim)       # predation mortality vs size
plotFMort(sim)          # fishing mortality vs size
plotGrowthCurves(sim)   # size vs age
plotDiet(params, species = "Cod")  # diet composition vs size
plotCDF(sim)            # cumulative biomass/abundance over size

# ── Plot any array directly, plus combine / compare tools ─────────────────────
plot(getResourceMort(params))   # any get*() array plots directly (resource mort vs size)
p <- plot(getBiomass(sim), species = "Cod")
addPlot(p, getBiomass(sim), species = "Herring", linetype = "dashed")  # add lines
plot2(getFMort(params), getFMort(params2), "Before", "After")  # compare arrays
plotRelative(getEGrowth(params), getEGrowth(params2))          # relative diff
plotHover(getBiomass(sim))      # interactive (hover) version of an array plot
animate(sim)                    # animate spectra through time

# ── Compare two simulations or models ─────────────────────────────────────────
plotSpectra2(params, params2, "Before", "After")
plotSpectraRelative(params, params2)      # relative difference of spectra
plotCDF2(sim, sim2, "Unfished", "Fished")
```
