---
title: "The Community Model"
output:
  html_document:
    number_sections: yes
    toc: yes
    fig_width: 5
    fig_height: 5
---

```{r include=FALSE}
library(mizer)
```
The simplest version of the size spectrum model is the community model. In the
community model, individuals are only characterized by their size and are
represented by a single group representing an across species average.
Reproduction is not considered and the recruitment flux $R$ is
set to be constant. The resource spectrum only extends to the start of the
community spectrum. Standard metabolism is turned off.

In this section we describe how a community model can be set up and projected
through time. We then use a community model to illustrate the idea of a trophic
cascade. Due to the relative simplicity of this type of model they are useful
for gently introducing some of the concepts behind the mizer package.
Consequently, this section should hopefully serve as an introduction to using
mizer.

## Setting up a community model {#sec:set_up_comm_model}

The first stage in implementing a model using `mizer` is to create an object of
class `MizerParams`. This class contains the model parameters including the
life-history parameters of the species in the model, the stock-recruitment
relationships, the fishing selectivity functions and the parameters of the
plankton spectrum.

To avoid having to make a `MizerParams` object directly, the
`set_community_model()` wrapper function, has been provided that conveniently
creates a `MizerParams` object specifically for a community model. The
documentation for the function can be seen by entering:
```{r label=help_set_community_model, eval=FALSE}
?set_community_model
```

As can be seen in the help page, the function can take many arguments. We can
ignore most of these for the moment as they almost all come with
default values.

The arguments that you should pay attention to are: `z0` (the level of
background mortality), `alpha` (the assimilation efficiency of the community),
`f0` (the average feeding level of the community which is used to calculate
$\gamma,$ as described in [the equation for
gamma](model_description.html#eq:gamma) and `recruitment` (the level of constant
recruitment).

Although default values for these parameters are provided, you are encouraged to
explore how changing the values affects the simulated community. For example,
the default value of `z0` is 0.1. Increasing this value effectively 'shortens'
the length of the community spectrum. The value of the constant recruitment
should be set so that the community spectrum is a continuation of the plankton
spectrum. This can be done with trial and error.  A reasonable value for `alpha`
is 0.2.

The `set_community_model()` function is called by passing in the arguments by
name. Any parameter that is not passed in is set to the default value. For
example, the following line sets up the parameters with `z0` = 0.1, `f0` = 0.7,
`alpha` = 0.2 and `recruitment` = 4e7. All other parameters will have their
default value:

```{r label=demo_comm_model_params}
params <- set_community_model(z0 = 0.1, f0 = 0.7, alpha = 0.2, recruitment = 4e7)
```

Calling the function creates and returns an object of type `MizerParams`. We can
check this using the `class()` function.
```{r label=class_comm_params}
class(params)
```

If you are going through this vignette for the first time, it is likely that you
have no idea what a `MizerParams` object actually is. In R-speak it is an *S4*
object, which means it is an object made up of *slots*. Slots are essentially
containers that store the object data. The names of these slots can be seen by
calling the `slotNames()` method on the object:
```{r label=slots_MP}
slotNames(params)
```

As you can see, the `params` object is made up of lots of slots. A quick
description can be found in the [`MizerParams` help
page](../reference/MizerParams-class.html) (run `class ? MizerParams`. The slots
of an object are accessed by using the `@` operator. For example, to access the
`w` slot (which contains a vector of the size bins in the model) you would use
```{r label=demo_slot_accessor, results='hide'}
params@w
```

Rather than picking through the slots to find out what is in a particular `MizerParams` object, a summary of the object can be seen by calling the `summary()` method on it:

```{r label=summary_comm_params}
summary(params)
```

In the summary you can see that the size range of the community spectrum has
been set from $`r min(params@w)`$ to $`r max(params@w)`$ and these are spread
over $`r length(params@w)`$ size bins. Similar information is available for the
plankton spectrum. Additionally, the community is made up of only one species,
called *Community*, which has an asymptotic size of 
$`r params@species_params$w_inf[1]`$ and a preferred predator prey mass ratio of 
$`r params@species_params$beta[1]`$. The `w_mat` parameter has been set to `NA` as
it is not used when running a community model. These values have all been set by
default using the `set_community_model()` function. If you want to set different
values for these, you will need to call the `set_community_model()` function and
pass in the desired argument values.

## Running the community model {#sec:running_comm_model}
By using the `set_community_model()` function we now have a `MizerParams` object
that contains all the information we need about the model community. We can use
this to perform a simulation and project the community through time. In the
`mizer` package, projections are performed using the `project()` function. You can
see the help page for `project()` for more details and it is described fully in
[the section on running a simulation.](running_a_simulation.html) We will ignore
the details for the moment and just use `project()` to run some simple
projections. The arguments for `project()` that we need to be concerned with are
`effort`, which determines the fishing effort (and therefore fishing mortality)
through time, and `t_max`, which is the length of the simulation. Initial
population abundances are set automatically by the `get_initial_n()` function.
It is possible to set your own initial abundances but we will not do this here.

To run a projection for 50 time steps, with no fishing effort (i.e. we want to
model an unexploited community) we run:
```{r label=demo_comm_run}
sim <- project(params, t_max = 50, effort = 0)
```

The resulting object, `sim`, is of type `MizerSim`. This class holds the results
of the simulation, including the community and plankton abundances at size
through time, as well as the original model parameters. It is explained in
detail in [the section on running a simulation.](running_a_simulation.html)

After running the projection, it is possible to explore the results using a
range of plots and analyses. These are described fully in [the section on
exploring the simulation results.](exploring_the_simulation_results.html)

To quickly look at the results of the projection you can call the generic
`plot()` method. This plots the feeding level, predation mortality, fishing
mortality and abundance by size in the last time step of the simulation, and the
biomass through time. Each of the plots can be show individually if desired.

```{r label=print_plot_comm_sim, fig.cap="Example plot of the community model."}
plot(sim)
```

In the above plot there are several things going on that are worth talking
about. Looking at the total biomass of the community against time, you can see
that the biomass quickly reaches a stable equilibrium. The other panels show
what is happening at the last time step in the simulation, which in this case is
when the community is at equilibrium. Fishing mortality is 0 because we set the
`effort` argument to 0 when running the simulation. The predation mortality rate
is clearly a function of size, with the smallest sizes experiencing the
highest levels of predation. The feeding level describes how satiated an
individual is, with 0 being unfed, and 1 being full satiated. The feeding level
at size will be strongly affected by the values of the `f0` and `alpha`
arguments passed to the `set_community_model()` function. The plankton and
community spectra are shown in the bottom panel of the plot (the plotted
plankton spectrum has been truncated to make for a better plot, but really
extends all the way back to $`r signif(params@w_full[1],3)`$ g). You can see
that the community spectrum forms a continuum with the plankton spectrum. This
is strongly affected by the level of fixed recruitment (the `recruitment`
argument passed to `set_community_model()`).

Note the hump in the biomass at the largest end of the community spectrum. This
is because the size spectrum model can be broadly described as 'big things
eating little things'. Given this, what is eating the very biggest things?
Without fishing pressure, the mortality of the largest individuals is only from
the background mortality (determined by the `z0` argument) and the mortality
from predation is almost 0. This is difficult to see in the plot due to the
predation mortality being so high for the smaller individuals.

We can see this more clearly by extracting the predation mortality information
from the `MizerSim` object, `sim`, that we created above. This is easily done by
using the `getPredMort()` function (see the help page for more details). There are
several functions that can be used for extracting information from a `MizerSim`
object, e.g. `getFeedingLevel()` and `getFMort()`. For more information see 
[the section on exploring the simulation results](exploring_the_simulation_results.html). Here we just call
`getPredMort()` using the `sim` object:

```{r label=get_m2_comm}
m2 <- getPredMort(sim)
```

The resulting `m2` object is an array that contains the predation mortality at
time by species by size. Here we only have one species so the species dimension
is dropped, leaving us with a two dimensional array of time by size. We
projected the model for $`r dim(m2)[1]-1`$ time steps but the length of the time
dimension is $`r dim(m2)[1]`$ as the initial population is also included as a
time step.

To pull out the predation mortality at size in the final time step we use:
```{r label=m2_comm_final_time_step, results='hide'}
m2[51,]
```

If you plot this predation mortality on a log-log scale you can see how the
predation mortality declines to almost zero for the largest sizes.

```{r label=print_plot_comm_m2, warning=FALSE, fig.cap="Predation mortality without fishing in the community model (note the log scales for both axes)."}
plot(x = sim@params@w, y = m2[51,], log = "xy", type = "l", 
     xlab = "Size", ylab = "Predation mortality")
```
 

## Example of a trophic cascade with the community model {#sec:trophic_cascade_comm_model}
It is possible to use the community model to simulate a trophic cascade.
To do this we need to perform two simulations, one with fishing and one without.

This means we need to consider how fishing is handled in `mizer`. The
`set_community_model()` function automatically sets the fishing selectivity to
have a knife-edge shape, with only individuals larger than 1 kg selected (the
size at the knife-edge can be changed by setting the `knife_edge_size`
argument). Although it is possible to change the selectivity function, here we
will use the default knife-edge selectivity. We set up the parameter object
exactly as before, but now we are explicitly setting the size at which
individuals are selected by the fishing gear.
```{r label=set_comm_fishing}
params_knife <- set_community_model(z0 = 0.1, recruitment = 4e7,
    alpha = 0.2, f0 = 0.7, knife_edge_size = 1000)
```

First we perform a simulation without fishing in the same way we did above by
setting the `effort` argument to 0:
```{r label=sim_comm_no_fish}
sim0 <- project(params_knife, effort = 0, t_max = 50)
```

Now we want to simulate again, this time with fishing. In the simulations,
fishing mortality is calculated as the product of the fishing selectivity,
effort and catchability (see [the section on fishing
gears](multispecies_model.html#sec:fishing_gear) for more details). By default
catchability is set to 1. This means that a fishing effort of 1 will result in a
fishing mortality of 1 for fully selected sizes. Here we run a simulation with
fishing effort set to 1 for the duration of the simulation:

```{r label=sim_comm_with_fish}
sim1 <- project(params_knife, effort = 1, t_max = 50)
```

You can compare the difference between these scenarios by using the `plot()` method as before. Of particular interest is the fishing mortality at size. The knife-edge selectivity at 1000 g can be clearly seen and an effort of 1 has resulted in a fishing mortality of 1 for the fully selected sizes.

```{r label=print_plot_comm_fmort, fig.cap="Summary plot for the community model when fishing with knife-edge selectivity at size = 1000 g."}
plot(sim1)
```
 
To explore the presence of a trophic cascade, we are interested in looking at the relative change in abundance when the community is fished compared to when it is not fished. To do this we need to get the abundances at size from the simulation objects. The abundances are stored in the `n` slot of the `MizerSim` objects. The `n` slot contains a three dimensional array with dimensions time x species x size. Here we have 51 time steps (50 from the simulation plus one which stores the initial population), 1 species and 100 sizes:

```{r label='dim_n'}
dim(sim0@n)
```

We want the abundances in the final time step, and we can use these to calculate the relative abundances:

```{r label='relative_comm_abundance'}
relative_abundance <- sim1@n[51,,] / sim0@n[51,,]
```

This can then be plotted using basic R plotting commands.
The sizes are stored in the `params@w` slot (a slot of a slot!).


```{r label=plot_relative_comm_abund, fig.cap="Relative abundances from the unfished (dashed line) and fished (solid line) community model."}
plot(x = sim0@params@w, y = relative_abundance, log = "x", type = "n",
    xlab = "Size (g)", ylab = "Relative abundance")
lines(x = sim0@params@w, y = relative_abundance)
lines(x = c(min(sim0@params@w), max(sim0@params@w)), y = c(1,1), lty = 2)
```


The impact of fishing on species larger than 1000g can be clearly seen.
The fishing pressure lowers the abundance of large fish (the decrease in relative abundance at 1000 g). This then relieves the predation pressure on their smaller prey (the preferred predator-prey size ratio is given by the $\beta$ parameter, which is set to 100 by default), leading to an increase in their abundance. This in turn increases the predation mortality on their smaller prey, which reduces their abundance and so on.  

## The impact of changing $\sigma$ {#sec:comm_model_change_sigma}

As described above, the $\sigma$ parameter determines the width of the predator prey size preference. Here we take a look at how changing the value of $\sigma$ can affect the dynamics of the community.
In the examples above, $\sigma$ is set in the `set_community_model()` function by default to a value of $`r params_knife@species_params$sigma[1]`$.
When projected through time, the community abundances converge to a stable equilibrium.
What happens if we reduce the value of $\sigma$, for example by setting it to 1.0? We can do this by passing in the new value of $\sigma$ into `set_community_model()`.

```{r label=comm_params_sigma1}
params_sigma1 <- set_community_model(z0 = 0.1,
     f0 = 0.7, alpha = 0.2, recruitment = 4e7, sigma = 1)
```

We want to project this new model through time using the `project()` function. Note that we have introduced a new argument: $dt$. This is the step size of the solver. It does not have anything to do with the biology in the model. It only affects the internal engine of `project()` that performs the projection. 

As you can see in the underlying model equations in [the model description section](model_description.html), the model runs in continuous time. Therefore, to project it forward, `project()` must solve the system of equations using numerical methods. The quality of these methods is strongly affected by $dt$. 
The default value of $dt$ is 0.1, which will be fine for most of the projections we run in this Vignette. Here it is necessary to reduce the value to 0.01 to avoid introducing any artefacts into the projected values. Decreasing $dt$ increases the time it takes to run a projection.
Here we project the new parameters object for 50 time steps without fishing:

```{r label=comm_project_sigma1}
sim_sigma1 <- project(params_sigma1, effort = 0, t_max = 50, dt=0.01)
```

Let's take a look at how the abundances change through time. We can do this with the `plotBiomass()` function:

```{r label=plot_comm_biomass_sigma1, fig.keep='last', fig.cap="Biomass of the community model when sigma is reduced to a value of 1.0."}
plotBiomass(sim_sigma1)
```

The plot above shows that abundances of the community no longer converge to a stable equilibrium and the dynamics appear to be chaotic. The ecological significance of the change in dynamics, and of the ability of simple community models to show chaotic behaviour, is still being debated. It can be argued that the size of the oscillations are too large to be 'true'. Additionally, when a trait-based model is implemented, the magnitude of the oscillations are much smaller. The next section is about [the trait based model.](trait_model.html)


