---
title: "The Community Model"
output:
  html_document:
    toc: yes
    fig_width: 5
    fig_height: 5
---

```{r include=FALSE}
library(mizer)
```
The simplest version of the size spectrum model is the community model. In the
community model, individuals are only characterized by their size and are
represented by a single group representing an across-species aggregate.
Reproduction is not considered; the reproduction rate $R$ is
set to be constant. The resource spectrum only extends to the start of the
community spectrum. Metabolism is turned off.

In this section we describe how a community model can be set up and projected
through time. We then use a community model to illustrate the idea of a trophic
cascade. Due to the relative simplicity of this type of model, it is useful
for gently introducing some of the concepts behind the mizer package.
Consequently, this section should hopefully serve as an introduction to using
mizer.

## Setting up a community model {#sec:set_up_comm_model}

The first stage in implementing a model using `mizer` is to create an object of
class `MizerParams`. This class contains the model parameters including the
life-history parameters of the species in the model, the fishing selectivity
functions and the parameters of the resource spectrum.

To avoid having to make a `MizerParams` object directly, the
`newCommunityParams()` wrapper function, has been provided that conveniently
creates a `MizerParams` object specifically for a community model. The
documentation for the function can be seen by clicking on the function 
name anywhere it appears on this page: `newCommunityParams()`.

As can be seen in the help page, the function can take many arguments. We can
ignore most of these for the moment as they almost all come with
default values.

The arguments that you should pay attention to are: `z0` (the external 
mortality rate), `alpha` (the assimilation efficiency of the community),
`f0` (the average feeding level of the community, which is used to calculate
$\gamma$) and `h` (the coefficient of the maximum intake rate).

Although default values for these parameters are provided, you are encouraged to
explore how changing the values affects the simulated community. For example,
the default value of `z0` is \(`r species_params(newCommunityParams())$z0`\). 
Increasing this value effectively 'shortens' the length of the community 
spectrum. 

The `newCommunityParams()` function is called by passing in the arguments by
name. Any parameter that is not passed in is set to the default value. For
example, the following line sets up the parameters with `z0` = 0.05, `f0` = 0.5. 
All other parameters will have their default value:

```{r label=demo_comm_model_params}
params <- newCommunityParams(z0 = 0.05, f0 = 0.5)
```

Calling the function creates and returns an object of type `MizerParams`. We can
check this using the `class()` function.
```{r label=class_comm_params}
class(params)
```

To work with mizer you do not have to worry about how this object is realised
internally. All you need to know is that it contains all the information needed
to run model simulations and that mizer provides you with functions to access,
use and modify this information.

A very brief description of the model contained in the MizerParams object can 
be seen by calling the `summary()` method on it:

```{r label=summary_comm_params}
summary(params)
```

In the summary you can see that the size range of the community spectrum has
been set from $`r min(w(params))`$ to $`r max(w(params))`$ and this is divided
into $`r length(w(params))`$ size bins. Similar information is available for the
resource spectrum. Additionally, the community is made up of only one species,
called *Community*, which has an asymptotic size of 
$`r species_params(params)$w_inf[1]`$ and a preferred predator prey mass ratio of 
$`r species_params(params)$beta[1]`$. The `w_mat` parameter has been set to `NA` as
it is not used when running a community model. These values have all been set by
default using the `newCommunityParams()` function. 

## Running the community model {#sec:running_comm_model}
By using the `newCommunityParams()` function we now have a `MizerParams` object
that contains all the information we need about the model community. We can use
this to perform a simulation and project the community through time. In the
`mizer` package, projections are performed using the `project()` function. You can
see the help page for `project()` for more details and it is described fully in
[the section on running a simulation.](running_a_simulation.html) We will ignore
the details for the moment and just use `project()` to run some simple
projections. The arguments for `project()` that we need to be concerned with are
`effort`, which determines the fishing effort (and therefore fishing mortality)
through time, and `t_max`, which is the number of years we want to project into the future. Initial
population abundances are set automatically by the `get_initial_n()` function.
It is possible to set your own initial abundances but we will not do this here.

To run a projection for 50 years, with no fishing effort (i.e. we want to
model an unexploited community) we run:
```{r label=demo_comm_run}
sim <- project(params, t_max = 50, effort = 0)
```

The resulting object, `sim`, is of type `MizerSim`. 
```{r}
class(sim)
```

This class holds the results
of the simulation, including the community and resource abundances at size
through time, as well as the original model parameters. It is explained in
detail in [the section on running a simulation.](running_a_simulation.html)

After running the projection, it is possible to explore the results using a
range of plots and analyses. These are described fully in 
[the section on exploring the simulation results.](exploring_the_simulation_results.html)

To quickly look at the results of the projection you can call the `plot()`
method. This plots the feeding level, predation mortality, fishing mortality and
abundance by size in the last time step of the simulation, and the total
biomass through time. Each of the plots can be shown individually if desired.

```{r label=print_plot_comm_sim}
plot(sim)
```

In the above plot there are several things going on that are worth talking
about. Looking at the total biomass of the community against time, you can see
that the biomass quickly reaches a stable equilibrium. The other panels show
what is happening at the last time step in the simulation, which in this case is
when the community is at equilibrium. Fishing mortality is 0 because we set the
`effort` argument to 0 when running the simulation. The predation mortality rate
is clearly a function of size, with the smallest sizes experiencing the
highest levels of predation. The feeding level describes how satiated an
individual is, with 0 being unfed, and 1 being full satiated. The feeding level
at size will be strongly affected by the values of the `f0` and `alpha`
arguments passed to the `newCommunityParams()` function. 

The resource and
community spectra are shown in the bottom panel of the plot (the plotted
resource spectrum has been truncated to make for a better plot, but really
extends all the way back to $`r signif(w_full(params)[1], 3)`$ g). You can see
that the community spectrum forms a continuum with the resource spectrum. This
is strongly affected by the level of fixed reproduction rate (the `reproduction`
argument passed to `newCommunityParams()`)

Note the hump in the biomass at the largest end of the community spectrum. This
is because the size spectrum model can be broadly described as 'big things
eating little things'. Given this, what is eating the very biggest things?
Without fishing pressure, the mortality of the largest individuals is only from
the external mortality (determined by the `z0` argument) and the mortality
from predation is almost 0. This is difficult to see in the plot due to the
predation mortality being so high for the smaller individuals.

We can see this more clearly by extracting the predation mortality information
from the `MizerSim` object, `sim`, that we created above. This is easily done by
using the `getPredMort()` function (see the help page for more details). There are
several functions that can be used for extracting information from a `MizerSim`
object, e.g. `getFeedingLevel()` and `getFMort()`. For more information see 
[the section on exploring the simulation results](exploring_the_simulation_results.html). 
Here we just call `getPredMort()` using the `sim` object:

```{r label=get_m2_comm}
pred_mort <- getPredMort(sim)
```

The resulting `pred_mort` object is an array that contains the predation mortality at
time by species by size. Here we only have one species so the species dimension
is dropped, leaving us with a two dimensional array of time by size. We
projected the model for $`r dim(pred_mort)[1]-1`$ time steps but the length of the time
dimension is $`r dim(pred_mort)[1]`$ as the initial population is also included as a
time step. We can get the index of the final time with
```{r}
idxFinalT(sim)
```

To pull out the predation mortality at size in the final time step we use:
```{r label=m2_comm_final_time_step, results='hide'}
pred_mort_final <- pred_mort[idxFinalT(sim), ]
```

If you plot this predation mortality on a log-log scale you can see how the
predation mortality declines to almost zero for the largest sizes.

```{r label=print_plot_comm_m2, warning=FALSE}
plot(x = w(params), y = pred_mort_final, log = "xy", type = "l", 
     xlab = "Size [g]", ylab = "Predation mortality [1/year]")
```

Note how we used `w(sim)` to get the vector of sizes to plot along the x-axis and
how we specified that we wanted to have logarithmic axes.

In the long run it is worthwhile to use the ggplot2 package for creating plots,
so we show also how you would create the above graph with `ggplot()`. For
more detail see the section on 
[using ggplot2 and plotly with mizer](plotting.html).
```{r}
library(ggplot2)
sd <- data.frame(x = w(params), y = pred_mort_final)
ggplot(sd, aes(x = x, y = y)) +
  geom_line() +
  scale_x_log10() + scale_y_log10() +
  xlab("Size [g]") + ylab("Predation mortality [1/year]")
```
 

## Example of a trophic cascade with the community model {#sec:trophic_cascade_comm_model}
It is possible to use the community model to simulate a trophic cascade.
To do this we need to perform two simulations, one with fishing and one without.

This means we need to consider how fishing is handled in `mizer`. The
`newCommunityParams()` function automatically sets the fishing selectivity to
have a knife-edge shape, with only individuals larger than 1 kg selected (the
size at the knife-edge can be changed by setting the `knife_edge_size`
argument). Although it is possible to change the selectivity function, here we
will use the default knife-edge selectivity. We set up the parameter object
with default parameters:
```{r label=set_comm_fishing}
params_knife <- newCommunityParams()
```

First we perform a simulation without fishing in the same way we did above by
setting the `effort` argument to 0:
```{r label=sim_comm_no_fish}
sim0 <- project(params_knife, effort = 0, t_max = 50)
```

Now we want to simulate again, this time with fishing. In the simulations,
fishing mortality is calculated as the product of the fishing selectivity,
effort and catchability (see 
[the section on fishing gears](multispecies_model.html#sec:fishing_gear) 
for more details). By default
catchability is set to 1. This means that a fishing effort of 1 will result in a
fishing mortality of 1/year for fully selected sizes. Here we run a simulation with
fishing effort set to 1 for the duration of the simulation:

```{r label=sim_comm_with_fish}
sim1 <- project(params_knife, effort = 1, t_max = 50)
```

You can compare the difference between these scenarios by using the `plot()`
method as before. Of particular interest is the fishing mortality at size. The
knife-edge selectivity at 1000 g can be clearly seen and an effort of 1 has
resulted in a fishing mortality of 1 for the fully selected sizes.

```{r label=print_plot_comm_fmort}
plot(sim1, power = 2)
```
 
To explore the presence of a trophic cascade, we are interested in looking at
the relative change in abundance when the community is fished compared to when
it is not fished. To do this we need to get the abundances at size from the
simulation objects. This is done with the `N()` function, which returns a three
dimensional array with dimensions time x species x size. Here we have 51 time
steps (50 from the simulation plus one which stores the initial population), 1
species and 100 sizes:

```{r label='dim_n'}
dim(N(sim0))
```

We want the abundances in the final time step, and we can use these to calculate
the relative abundances:

```{r label='relative_comm_abundance'}
relative_abundance <- N(sim1)[51, , ] / N(sim0)[51, , ]
```
For convenience, to save you from having to determine the index of the final 
time, mizer provides the function `finalN()`, so we could have done
```{r label='relative_comm_abundance_simple'}
relative_abundance <- finalN(sim1) / finalN(sim0)
```

This can then be plotted using basic R plotting commands.


```{r label=plot_relative_comm_abund}
plot(x = w(params), y = relative_abundance, log = "x", type = "n",
    xlab = "Size (g)", ylab = "Relative abundance")
lines(x = w(params), y = relative_abundance)
lines(x = c(min(w(params)), max(w(params))), y = c(1, 1), lty = 2)
```

The impact of fishing on species larger than 1000g can be clearly seen. The
fishing pressure lowers the abundance of large fish (the decrease in relative
abundance at 1000 g). This then relieves the predation pressure on their smaller
prey (the preferred predator-prey size ratio is given by the $\beta$ parameter,
which is set to 100 by default), leading to an increase in their abundance. This
in turn increases the predation mortality on their smaller prey, which reduces
their abundance and so on.

## The impact of changing $\sigma$ {#sec:comm_model_change_sigma}

As described above, the $\sigma$ parameter determines the width of the predator
prey size preference. Here we take a look at how changing the value of $\sigma$
can affect the dynamics of the community.

In the examples above, \(\sigma\) is set in the `newCommunityParams()` function 
by default to a value of \(`r species_params(params_knife)$sigma[1]`\).
We can see this by looking at the `sigma` column of the species_params data
frame that is contained in the MizerParams object:
```{r}
species_params(params)$sigma
```

When projected through time, the community abundances converge to a stable 
equilibrium. What happens if we reduce the value of $\sigma$, for example by 
setting it to 1.0? We can do this by passing in the new value of $\sigma$ into 
`newCommunityParams()`.

```{r label=comm_params_sigma1}
params_sigma1 <- newCommunityParams(sigma = 1)
```

We want to project this new model through time using the `project()` function.
Here we project the new parameter object for 50 time steps without fishing and
save at intervals of 0.1 years (`t_save = 0.1`):

```{r label=comm_project_sigma1}
sim_sigma1 <- project(params_sigma1, effort = 0, t_max = 50, 
                      dt = 0.01, t_save = 0.1)
```
Note that we have introduced a new argument: $dt$. This is the step size of the
solver. It does not have anything to do with the biology in the model. It only
affects the internal engine of `project()` that performs the projection.
As you can see in the underlying model equations in 
[the model description section](model_description.html), 
the model is formulated in continuous time. Therefore,
to project it forward, `project()` must solve the system of equations using
numerical methods. The quality of these methods is strongly affected by $dt$.
The default value of $dt$ is 0.1, which will be fine for most of the projections
we run in this document. Here it is necessary to reduce the value to 0.01 to
avoid introducing any artefacts into the projected values. Decreasing $dt$
increases the time it takes to run a projection.

Let's take a look at how the abundances change through time. We can do this with
the `plotBiomass()` function:

```{r label=plot_comm_biomass_sigma1}
plotBiomass(sim_sigma1)
```

The plot above shows that abundances of the community no longer converge to a
stable equilibrium and the dynamics appear to be chaotic. The ecological
significance of the change in dynamics, and of the ability of simple community
models to show chaotic behaviour, is still being debated. It can be argued that
the size of the oscillations are too large to be 'true'. Additionally, when a
trait-based model is implemented, the magnitude of the oscillations are much
smaller.

The next section is about [the trait based model.](trait_model.html)


