---
title: "Calculation of Default Parameter Values"
output:
  html_document:
    toc: yes
    fig_width: 5
    fig_height: 5
vignette: >
  %\VignetteIndexEntry{Calculation of Default Parameter Values}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
library(mizer)
```

## Introduction & Philosophy

A key design feature of the `mizer` package is that a multi-species size-spectrum model can be set up using only a small amount of information. This is possible because `mizer` uses **allometric scaling relations** and **size-based feeding rules** to choose sensible default values for parameters that are not explicitly provided by the user.

For example, to set up a model using the `newMultispeciesParams()` function, the user only *needs* to provide the maximum sizes ($w_{\text{max}}$) for the species in the community. Almost all other physiological, ecological, and mortality parameters can be automatically filled in using defaults.

This vignette details the philosophy, the relationships, and the mathematical derivations behind the default parameter values calculated by `mizer`. For a description of the model equations themselves, see the [vignette on the model description](model_description.html).

---

## Species Parameter Defaults Reference Table

When creating a `MizerParams` object, `validSpeciesParams()` (and subsequent setup functions like `setSearchVolume()`, `setMaxIntakeRate()`, `setMetabolicRate()`, `setExtMort()`, and `setReproduction()`) check the species parameter data frame and fill in missing parameters.

Below is a reference of the default values and the functions responsible for calculating them:

| Parameter | Description | Default Value / Formula | Function |
| :--- | :--- | :--- | :--- |
| `w_repro_max` | Size at which reproduction stops | `w_max` | `validSpeciesParams()` |
| `w_mat` | Size at maturity | `w_max / 4` | `validSpeciesParams()` |
| `w_min` | Egg size / birth size | `0.001` g | `validSpeciesParams()` |
| `alpha` | Assimilation efficiency | `0.6` | `validSpeciesParams()` |
| `interaction_resource` | Species interaction strength with resource | `1` | `validSpeciesParams()` |
| `n` | Allometric scaling exponent of intake | `3/4` | `validSpeciesParams()` |
| `p` | Allometric scaling exponent of metabolism | `n` (which is `3/4`) | `validSpeciesParams()` |
| `z_ext` | Size-dependent external mortality | `0` | `validSpeciesParams()` |
| `d` | Exponent of size-dependent external mortality | `n - 1` | `validSpeciesParams()` |
| `E_ext` | External encounter rate | `0` | `validSpeciesParams()` |
| `D_ext` | External diffusion rate | `0` | `validSpeciesParams()` |
| `is_background` | Flag indicating background species | `FALSE` | `validSpeciesParams()` |
| `h` | Maximum intake rate coefficient | Derived from growth/age at maturity (or `30`) | `get_h_default()` |
| `gamma` | Search volume coefficient | Derived from target feeding level $f_0$ | `get_gamma_default()` |
| `f0` | Target feeding level | `0.6` (or derived from `gamma` if given) | `get_f0_default()` |
| `fc` | Critical feeding level | `0.2` | `set_species_param_default()` |
| `ks` | Standard metabolic rate coefficient | Derived from critical feeding level $f_c$ | `get_ks_default()` |
| `k` | Activity/movement cost coefficient | `0` | `setMetabolicRate()` |
| `z0` | Constant external mortality rate | `z0pre * w_max^z0exp` | `setExtMort()` |
| `erepro` | Reproduction efficiency | `1` | `setReproduction()` |
| `beta` | Preferred predator/prey mass ratio | `30` | `default_pred_kernel_params()` |
| `sigma` | Width of predation kernel | `2` | `default_pred_kernel_params()` |

---

## The Dependency Chain

Calculating the default physiological parameters follows a sequential dependency chain. One parameter value determines the next, starting from basic size parameters and ending with standard metabolic rates:

```
[w_min, w_mat, age_mat] (Given or from VB parameters)
         │
         ▼
 ┌───────────────┐
 │       h       │  (Maximum intake rate coefficient)
 └───────┬───────┘
         ├──────────────────────────┐
         ▼                          ▼
 ┌───────────────┐          ┌───────────────┐
 │     gamma     │          │      ks       │  (Metabolic rate coefficient)
 └───────────────┘          └───────────────┘
(Search volume coefficient)
```

1. **Age at maturity** ($\text{age}_{\text{mat}}$) is either supplied directly, or derived from von Bertalanffy parameters via `age_mat_vB()`.
2. **Maximum intake rate** ($h$) is calculated from $\text{age}_{\text{mat}}$, egg size $w_{\text{min}}$, maturity size $w_{\text{mat}}$, and the assimilation parameters.
3. **Search volume** ($\gamma$) and **metabolism** ($k_s$) are then both computed using the calculated value of $h$.

---

## Mathematical Derivations

### 1. Maximum Intake Rate Coefficient ($h$)

The maximum intake rate coefficient $h_i$ for species $i$ is calculated by `get_h_default()`. The derivation is based on the idea that an average individual must grow from egg size $w_{\text{min}}$ to maturity size $w_{\text{mat}}$ in a given time $\text{age}_{\text{mat}}$ when feeding at a constant feeding level $f_0$ (which defaults to $0.6$).

By definition, the rate of somatic growth is:
$$\frac{dw}{dt} = g(w)$$

Prior to maturation, individuals invest nothing in reproduction ($\psi(w) = 0$), so growth is equal to the energy available for growth and reproduction $E_r(w)$:
$$g(w) = \max(0, \alpha f(w) h w^n - \text{metab}(w))$$

Under the default assumptions:
* The feeding level is constant: $f(w) = f_0$.
* The activity cost is zero: $k = 0$, so $\text{metab}(w) = k_s w^p$.
* The metabolic exponent equals the intake exponent: $p = n$.

Substituting these in gives:
$$\frac{dw}{dt} = \alpha f_0 h w^n - k_s w^n = (\alpha f_0 h - k_s) w^n$$

We also define the critical feeding level $f_c$ as the feeding level at which assimilation exactly balances metabolic costs, i.e., $g(w) = 0$. This implies:
$$\alpha f_c h w^n = k_s w^n \implies k_s = \alpha f_c h$$

Using this expression for $k_s$ in the growth equation yields:
$$\frac{dw}{dt} = (\alpha f_0 h - \alpha f_c h) w^n = \alpha h (f_0 - f_c) w^n$$

We can solve this differential equation by separating variables and integrating from birth ($t = 0$, $w = w_{\text{min}}$) to maturity ($t = \text{age}_{\text{mat}}$, $w = w_{\text{mat}}$):
$$\int_{w_{\text{min}}}^{w_{\text{mat}}} w^{-n} \, dw = \alpha h (f_0 - f_c) \int_{0}^{\text{age}_{\text{mat}}} dt$$

Evaluating the integrals:
$$\frac{w_{\text{mat}}^{1-n} - w_{\text{min}}^{1-n}}{1-n} = \alpha h (f_0 - f_c) \text{age}_{\text{mat}}$$

Solving for $h$ gives the formula implemented in `get_h_default()`:
$$h = \frac{w_{\text{mat}}^{1-n} - w_{\text{min}}^{1-n}}{\text{age}_{\text{mat}} (1-n) \alpha (f_0 - f_c)}$$

If the age at maturity is not known, `mizer` attempts to estimate it from von Bertalanffy parameters using:
$$\text{age}_{\text{mat}} = - \frac{\ln\left(1 - \left(\frac{w_{\text{mat}}}{w_{\infty}}\right)^{1/b}\right)}{k_{\text{vb}}} + t_0$$
If no growth information is provided at all, `mizer` defaults to $h = 30$.

---

### 2. Search Volume Coefficient ($\gamma$)

The search volume coefficient $\gamma_i$ is calculated by `get_gamma_default()`. It is selected such that if the prey abundance were described by a power-law resource spectrum:
$$N_R(w_p) = \kappa w_p^{-\lambda}$$
then the resulting encounter rate would lead to the target feeding level $f_0$ (default $0.6$).

The encounter rate $E_i(w)$ of predator species $i$ at size $w$ is:
$$E_i(w) = \gamma_i w^q \int \theta_{iR} N_R(w_p) \phi_i(w, w_p) w_p \, dw_p$$

Substituting the power-law resource spectrum:
$$E_i(w) = \gamma_i w^q \theta_{iR} \kappa \int w_p^{1-\lambda} \phi_i(w, w_p) \, dw_p$$

If the predation kernel depends only on the predator/prey mass ratio $x = w/w_p$, then we can substitute $w_p = w/x$ and $dw_p = -w/x^2 \, dx$:
$$\int w_p^{1-\lambda} \phi_i(w/w_p) \, dw_p = w^{2-\lambda} \int_0^\infty x^{\lambda - 2} \phi_i(x) \, dx$$

Thus, the encounter rate scales with size as:
$$E_i(w) = \gamma_i w^{q + 2 - \lambda} \theta_{iR} \kappa \int_0^\infty x^{\lambda - 2} \phi_i(x) \, dx$$

Let us define the size-independent component of the encounter rate as:
$$A_{\text{avail}, i} = \theta_{iR} \kappa \int_0^\infty x^{\lambda - 2} \phi_i(x) \, dx$$
so that $E_i(w) = \gamma_i w^{q + 2 - \lambda} A_{\text{avail}, i}$.

The feeding level is given by:
$$f_i(w) = \frac{E_i(w)}{E_i(w) + h_i w^n}$$

For the feeding level to be independent of size and equal to $f_0$, we must require the size scaling exponents of encounter and maximum intake to match:
$$q + 2 - \lambda = n \implies q = n + \lambda - 2$$
When this holds, the size $w$ cancels out:
$$f_0 = \frac{\gamma_i A_{\text{avail}, i}}{\gamma_i A_{\text{avail}, i} + h_i}$$

Solving this equation for $\gamma_i$ yields:
$$\gamma_i = \frac{h_i}{A_{\text{avail}, i}} \frac{f_0}{1 - f_0}$$

In the code, $A_{\text{avail}, i}$ is calculated numerically by setting $\gamma_i = 1$ and evaluating the encounter rate at the maximum size grid point divided by the appropriate power of $w$:
$$\text{avail_energy}_i = \frac{E_{i}|_{\gamma_i=1}(w_{\text{max}})}{w_{\text{max}}^{q + 2 - \lambda}}$$
Then, the default $\gamma$ is:
$$\gamma_i = \frac{h_i}{\text{avail_energy}_i} \frac{f_0}{1 - f_0}$$

---

### 3. Standard Metabolic Rate Coefficient ($ks$)

The standard metabolic rate coefficient $k_{s,i}$ is calculated by `get_ks_default()`. It is chosen so that the critical feeding level $f_c$ (default $0.2$) is exactly the feeding level required for assimilation to balance metabolism at maturity size $w_{\text{mat}}$.

At the critical feeding level $f_c$, the assimilation rate equals metabolic losses. At size $w_{\text{mat}}$:
$$\alpha_i f_c h_i w_{\text{mat}}^n = \text{metab}_i(w_{\text{mat}})$$

Using the default metabolic cost equation $\text{metab}_i(w) = k_{s,i} w^p$ (with activity $k = 0$):
$$\alpha_i f_c h_i w_{\text{mat}}^n = k_{s,i} w_{\text{mat}}^p$$

Solving for $k_{s,i}$ gives the formula:
$$k_{s,i} = f_c \alpha_i h_i w_{\text{mat}}^{n - p}$$

If $n = p$, this simplifies to:
$$k_{s,i} = f_c \alpha_i h_i$$

---

### 4. Target Feeding Level ($f_0$)

For models where the search volume coefficient $\gamma_i$ is provided by the user but the target feeding level $f_0$ is not, `get_f0_default()` calculates the inverse of the search volume default.

Using the relationship derived above:
$$f_{0,i} = \frac{\gamma_i A_{\text{avail}, i}}{\gamma_i A_{\text{avail}, i} + h_i}$$
Since the encounter rate calculated with the given $\gamma_i$ is $E_i(w) = \gamma_i w^{q+2-\lambda} A_{\text{avail}, i}$, we evaluate:
$$\text{avail_energy}_i = \frac{E_i(w_{\text{max}})}{w_{\text{max}}^{q+2-\lambda}}$$
And then calculate:
$$f_{0,i} = \frac{1}{h_i / \text{avail_energy}_i + 1}$$

---

## Other Defaults

* **External Mortality ($z_0$)**: Background mortality that is not due to predation or fishing. If not specified, it scales allometrically with maximum size:
  $$z_{0, i} = z_{0\text{pre}} w_{\text{max}, i}^{z_{0\text{exp}}}$$
  where $z_{0\text{pre}}$ defaults to $0.6$ and $z_{0\text{exp}}$ defaults to $n - 1$ (typically $-1/4$).
* **Predation Kernel**: By default, `mizer` uses a lognormal predation kernel $\phi_i(w, w_p)$ with preferred predator-prey mass ratio $\beta = 30$ and width $\sigma = 2$.
* **Maturity Ogive**: The proportion of individuals mature at size $w$ is a sigmoidal function:
  $$\text{maturity}(w) = \frac{1}{1 + (w/w_{\text{mat}})^{-10}}$$

---

## Defaults Editions

Sometimes improved methods for choosing defaults are added to `mizer`. To prevent older code from changing behaviour unexpectedly when the package is updated, `mizer` uses a versioning system controlled by the `defaults_edition()` function.

* **Edition 1** (Current default): Historically chosen default relations.
* **Edition 2** (Available): Catchability is set to $0.3$ (instead of $1$), initial fishing effort is $1$ (instead of $0$), initial abundance is calculated using the steady-state solver instead of a simple power law, and `psi` is determined strictly by the maturity ogive and the reproductive investment proportion.

---

## Worked Example

The following code illustrates how `mizer` automatically calculates defaults when setting up a new model with only the species name and maximum weight:

```{r}
# Create species parameter data frame with only w_max specified
simple_species <- data.frame(
  species = c("Species A", "Species B"),
  w_max = c(1000, 8000)
)

# Initialize parameters
params <- newMultispeciesParams(simple_species)

# Inspect the automatically filled species parameters
species_params(params)[, c("species", "w_mat", "w_min", "h", "gamma", "ks")]
```
