---
title: "Developer Guide"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Mizer Developer Vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

This vignette is for you if you need to extend mizer 
to meet the needs of your research project.
You will already have read the [main mizer vignette](https://cran.r-project.org/web/packages/mizer/vignettes/mizer_vignette.pdf) and thus be
familiar with what mizer can do out of the box.
You now want to implement the extension or modification of the
model required for your research, and for that you need to dive into 
the internal workings of mizer. This vignette is meant to make
that as easy as possible.

The first thing you should do, even before reading this vignette, is to
go to https://github.com/sizespectrum/mizer/issues and create a new "issue""
to share your ideas and plans with the mizer community. You may get back 
valuable feedback and advice. Another way to get in touch with the
mizer community is via the [size-spectrum modelling Google group](https://groups.google.com/forum/#!forum/size-spectrum-models).


# Setting up working environment
In this section we describe how to set up your working environment
to allow you to easily work with the mizer code. Much of it you
may already have in place, so feel free to skip ahead.

## Installing R, RStudio, devtools

Mizer is compatible with R versions 3.1 and later.
If you still need to [install R](https://cran.r-project.org/), simply install 
the latest version. This vignette was prepared with
`r R.version.string`.

This guide assumes that you will be using RStudio to work with R. There is
really no reason not to use RStudio and it makes a lot of things much easier.
RStudio develops rapidly and adds useful features all the time and so it pays 
to upgrade to the [latest version](https://www.rstudio.com/products/rstudio/download/preview/) 
frequently. This guide was written with version 1.2.1114.

To work with R packages as a developer, you will need to
[install additional tools](https://support.rstudio.com/hc/en-us/articles/200486498-Package-Development-Prerequisites). Once you have these in place, you should install the devtools package with
```{r eval=FALSE}
install.packages("devtools")
```
You are now all set to develop R packages, and 
[RStudio makes this extra easy](https://support.rstudio.com/hc/en-us/articles/200486488-Developing-Packages-with-RStudio). There is even a cheat sheet for package development accessible from the Help menu in RStudio.

## Installing git
In case git is not already installed on your system, you need to 
[install it](http://happygitwithr.com/install-git.html). You do not need to install
any GUI for git because RStudio has 
[built-in support for git](https://support.rstudio.com/hc/en-us/articles/200532077-Version-Control-with-Git-and-SVN). 
A good place to learn about using git and github is 
[this chapter](http://r-pkgs.had.co.nz/git.html) 
in the guide by Hadley on R package development.

## Forking mizer from github

Mizer is developed using the version control system _git_ and the code is hosted on github.
To work with the code you will create your own git repository with a copy of the mizer code.

Go to https://github.com/sizespectrum/mizer and fork it into your own repository. 
Then create a version control new project in Rstudio, and indicate your depository url.
On the "Git" tab select the "dev" branch. This contains the latest development
code for mizer. 

To run this code go to the "Build" tab in RStudio and click on "Install and Restart". 
This builds and loads the mizer package based on this development code.

# Finding your way around the mizer code

Mizer is organised in a modular fashion. It is separated into setup functions,
simulation functions, and analysis and plotting functions. These will
be described below after we have introduced the mizer model.

The core of mizer is the `project()` function
that runs a simulation of the size-spectrum model. It takes a specification
of the model contained in an object of type `MizerParams` and returns the
results of the simulation in an object of type `MizerSim`.

There are several different methods for setting up a `MizerParams` object for
specifying various concrete models. These setup functions make various 
simplifying assumptions about the model parameters to reduce the amount of
information that needs to be specified. This usually takes the form of
assuming allometric scaling laws. Below we will first describe the most
general size spectrum model handled by mizer before describing the
specialisations for which mizer has special setup functions.

There are many functions for analysing and plotting the results of a mizer
simulation contained in a `MizerSim` object.

The `MizerParams` and `MizerSim` objects are S4 objects,
meaning that their slots are rigorously defined and are accessed with the '@'
notation. You do not need to learn about S4 classes in order to understand the
mizer code, because the code avoids using S4 methods. In the presentation
below we assume that the `MizerParams` object is called `params` and the
`MixerSim` object is called `sim`.


## The MizerParams object
An object of class 'MizerParams' holds all the information needed for the
'project()' method to simulate a model. 

## The project methods

## The MizerSim object

## Shiny apps

# Developer guidelines

## Collaborating on Github

## Coding style
[Google's R Style Guide](https://google.github.io/styleguide/Rguide.xml)
[The tidyverse style guide](http://style.tidyverse.org/)

## Always write tests for your code
We use [testthat](http://testthat.r-lib.org/) and
[shinytest](https://rstudio.github.io/shinytest/articles/shinytest.html).



[Edit this page](https://github.com/gustavdelius/mizer/edit/master/vignettes/developer_vignette.Rmd)
