---
title: "Dynamic stability and Hopf bifurcations"
output:
  html_document:
    number_sections: yes
    toc: yes
    fig_width: 6
    fig_height: 6
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

## Introduction

In a mizer model, a steady state represents an equilibrium where the abundance of every species at every size remains constant over time. While `steady()` attempts to find such a state by projecting the dynamics forward in time until changes stop, this only works if the steady state is *dynamically stable*. 

If a steady state is unstable, small perturbations will grow rather than decay, and the system will naturally move away from the steady state, often settling into persistent oscillations (a limit cycle) or more complex dynamics.

Mizer provides the `steadyNewton()` function to find steady states irrespective of their stability by solving the algebraic equilibrium equations directly. Once a steady state is found, its stability can be analysed using `getStability()`.

This vignette explains how to perform this stability analysis, what the results mean, and how to use the linearised limit cycle approximation provided by `getLimitCycleSim()` to understand emergent oscillations around a Hopf bifurcation.

## Linear stability analysis

Mizer projects the state of the ecosystem forward in discrete time steps. We can write this discrete-time map abstractly as:

$$ N(t+\Delta t) = G(N(t)) $$

where $N(t)$ represents the full state vector (abundances of all species at all sizes) at time $t$. At a steady state $N^*$, we have $N^* = G(N^*)$.

To determine if $N^*$ is stable, we linearise the dynamics for small perturbations $\delta N(t) = N(t) - N^*$ around the steady state:

$$ \delta N(t+\Delta t) \approx L \, \delta N(t) $$

where $L$ is the Jacobian matrix evaluated at $N^*$, with elements $L_{ij} = \frac{\partial G_i}{\partial N_j}(N^*)$.

The stability of the steady state is determined by the eigenvalues $\lambda_i$ of $L$. The **spectral radius** is the maximum of the absolute values (moduli) of these eigenvalues: $\rho = \max_i |\lambda_i|$. 

## Analyzing the North Sea model

Let's load mizer and start with the standard North Sea model.

```{r setup}
suppressMessages(devtools::load_all("..", quiet = TRUE))
params <- NS_params
```

First, we find its steady state and check its stability. We can do this in one step by passing `stability = TRUE` to `steadyNewton()`.

```{r NS_stable}
# Find the exact steady state and compute its stability
params_stable <- steadyNewton(params, stability = TRUE)

# The stability results are attached as an attribute
stab <- attr(params_stable, "stability")
stab$spectral_radius
```

The spectral radius is less than 1 (around 0.8), so the steady state is dynamically stable. Small perturbations will decay and the system will return to this equilibrium.

However, the way the system returns to equilibrium is determined by the dominant eigenvalue. If the dominant eigenvalue is complex, the system will exhibit decaying oscillations as it returns to the steady state. Let's check the dominant eigenvalue:

```{r dominant_mode}
stab$eigenvalues[1]
```

The dominant eigenvalue is indeed complex (with a non-zero imaginary part). This means the leading mode of the system is an oscillation (a stable spiral).

## The Hopf bifurcation and Limit Cycles

If we were to change parameters (for instance, increasing the maximum reproduction rate `R_max` or changing feeding preferences) to make the model less stable, this complex eigenvalue could cross the unit circle (its modulus exceeding 1). When this happens, the system undergoes a **Hopf bifurcation** (or Neimark-Sacker bifurcation in discrete time). 

At a Hopf bifurcation, the system transitions from a stable steady state to an oscillating **limit cycle**. The period of this oscillation is determined by the argument (angle) of the complex eigenvalue $\lambda$:

$$ \text{Period} \approx \frac{2\pi}{|\arg(\lambda)|} \text{ time steps} $$

Mizer calculates the period of the dominant complex eigenvalue and reports it as `hopf_period`, even if the state is stable.

```{r hopf_period}
stab$hopf_period
```

The model has an oscillatory component with a period of roughly `r round(stab$hopf_period, 1)` years.

## Simulating the oscillation shape

Because the dominant mode is complex, perturbations will initially follow the shape of the corresponding complex eigenvector.

Mizer provides the `getLimitCycleSim()` function to construct a `MizerSim` object representing one period of this oscillation in the linear approximation. While the true perturbations in our stable model will decay, `getLimitCycleSim()` ignores the decay (acting as if we are exactly at the bifurcation point) to isolate and visualise the pure shape of the cohort oscillations.

```{r limit_cycle}
lcs <- getLimitCycleSim(params_stable)
```

Because `lcs` is a standard `MizerSim` object, we can use any of the standard mizer plotting functions to visualise the pattern of the cycle. For example, we can plot the biomass dynamics over one period:

```{r plot_lcs_biomass}
plotBiomass(lcs)
```

This linearised cycle provides a very useful picture of which species and sizes are driving the natural oscillations in the system, and what the cohort resonance would look like if the system were destabilised.

## Comparing with full nonlinear dynamics

We can verify our linear stability analysis by running a full dynamic projection using `project()`. Since the steady state is stable (spectral radius < 1), if we perturb it, it should oscillate with the predicted period but the amplitude will decay over time.

We will perturb the initial abundance by adding a 10% bump to the initial state, and project for 50 years.

```{r true_projection}
# Perturb the steady state
params_perturbed <- params_stable
params_perturbed@initial_n <- params_perturbed@initial_n * 1.1

# Project forward in time
sim_full <- project(params_perturbed, t_max = 50)

plotBiomass(sim_full)
```

As predicted by the spectral radius < 1, the steady state is stable and the biomasses oscillate with decaying amplitude, eventually settling back to the straight lines of the steady state. 

If you compare the shape of these decaying oscillations to the pure, un-decayed oscillations produced by `getLimitCycleSim()` above, you will see they match perfectly. The linear stability analysis successfully decomposes the complex nonlinear ecosystem dynamics into a simple period, shape, and decay rate.

