---
title: "Dynamic stability and Hopf bifurcations"
output:
  html_document:
    number_sections: yes
    toc: yes
    fig_width: 6
    fig_height: 6
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

## Introduction

In a mizer model, a steady state represents an equilibrium where the abundance of every species at every size remains constant over time. While `steady()` attempts to find such a state by projecting the dynamics forward in time until changes stop, this only works if the steady state is *dynamically stable*.

If a steady state is unstable, small perturbations will grow rather than decay, and the system will naturally move away from the steady state, often settling into persistent oscillations (a limit cycle) or more complex dynamics.

Mizer provides the `steadyNewton()` function to find steady states irrespective of their stability by solving the algebraic equilibrium equations directly. Once a steady state is found, its stability can be analysed using `getStability()`.

This vignette explains how to perform this stability analysis, what the results mean, how mizer's steady-state finders detect and characterise limit cycles, and how to use the linearised limit cycle approximation provided by `getLimitCycleSim()` to understand emergent oscillations around a Hopf bifurcation.

## Linear stability analysis

Mizer discretises the size axis but leaves time continuous. On the size grid the
model is therefore a system of ordinary differential equations,

$$ \frac{dN}{dt} = F(N), $$

where $N$ is the full state vector (abundances of all species at all sizes) and
$F$ collects the divergence of the growth flux, the mortality sink and the
reproductive influx at the egg size. A steady state $N^*$ is a state at which
$F(N^*) = 0$.

Close to $N^*$ a small perturbation $\delta N(t) = N(t) - N^*$ obeys the
linearised equation $d\,\delta N/dt = J\,\delta N$, where $J = \partial F /
\partial N$ evaluated at $N^*$ is the Jacobian matrix. Its eigenvalues
$\lambda_i$ decide everything: each mode of the perturbation evolves as
$e^{\lambda_i t}$, so the steady state is **stable** when every eigenvalue has
negative real part and **unstable** as soon as one has a positive real part.

Because abundances in a mizer model span dozens of orders of magnitude across
the size spectrum, evaluating stability in terms of absolute perturbations can
be numerically poorly-scaled. It is more natural to consider **multiplicative
(relative) perturbations**, where we perturb the steady state by a small
relative amount $x(t)$:

$$ N_i(t) = N_i^* \left(1 + x_i(t)\right). $$

The relative Jacobian $K$ that governs these relative perturbations,

$$ K_{ij} = \frac{N_j^*}{N_i^*} \frac{\partial F_i}{\partial N_j}(N^*), $$

is related to $J$ by a similarity transform $K = D^{-1} J D$, where $D$ is a
diagonal matrix of the steady-state abundances $N^*$. Because $K$ and $J$ are
similar matrices, **they have exactly the same eigenvalues**.

`getStability()` computes $J$ by perturbing each state variable in turn and
differencing the rates of change that mizer's own rate functions return, so the
eigenvalues it reports are those of the model as it is configured, including any
rate function registered with `setRateFunction()` and whichever spatial scheme
`second_order_w()` selects. **No time step enters this calculation.** The
eigenvalues are a property of the model, not of a solver, and they are what a
simulation with a small enough time step converges to.

*(Note on numerical implementation: while analysing the relative Jacobian $K$ is
conceptually natural, computing it requires dividing by $N_i^*$, which causes
floating-point overflow for structurally zero size classes (e.g., extinct species
or truncated tails). Therefore, `getStability()` numerically computes the
absolute Jacobian $J$ using finite-difference steps scaled multiplicatively to
the local density. This is mathematically equivalent to the relative analysis but
avoids dividing by zero.)*

### The stability of the model and the stability of the solver

There is a second, quite different question one can ask: not whether the *model*
is stable, but whether *mizer's numerical time step* is. `project()` advances
the state by a discrete map

$$ N(t+\Delta t) = G(N(t)), $$

and the eigenvalues $\mu_i$ of the Jacobian of $G$ at the fixed point decide
whether that map amplifies a perturbation: it does not when the **spectral
radius** $\max_i|\mu_i|$ is below 1. `getDiscreteStability()` reports those
numbers for a chosen step size `dt`.

The two questions have different answers, and the difference matters. Mizer's
step solves the transport implicitly, and implicit schemes introduce a "temporal
numerical diffusion" that artificially damps oscillations. At a large `dt` a
steady state that the model does not hold can therefore have a spectral radius
below 1 — the simulation settles down onto a state that is an artefact of the
step size. We will see exactly that below.

It is tempting to try to recover $J$ from $G$ algebraically. If mizer's step were
*fully* implicit, $N^{t+\Delta t} = N^t + \Delta t\,F(N^{t+\Delta t})$, then the
two Jacobians would be related exactly by $\mu = 1/(1 - \Delta t\,\lambda)$ and
one could simply invert. But mizer's step is not fully implicit: it evaluates all
the rates — encounter, feeding level, predation mortality, growth — at the state
at the *start* of the step, and only the transport solve that follows is
implicit. The inversion is then correct only to first order in $\Delta t$, and at
$\Delta t = 1$ it is wrong by more than an order of magnitude. This is why
`getStability()` differentiates $F$ directly instead.

## Driving the North Sea model unstable

Let's load mizer and start with the standard North Sea model.

```{r setup}
suppressMessages(devtools::load_all("..", quiet = TRUE))
params <- NS_params
```

The North Sea model at its default fishing effort is dynamically stable, so it makes a rather uneventful example: perturb it and it simply settles back down. To see something more interesting we increase the fishing pressure, setting the effort of every gear to `1.5`. This pushes the model across a **Hopf bifurcation** into a regime where the steady state is unstable.

Because the steady state is now unstable, `steady()` cannot find it — projecting the dynamics forward would diverge away from it. This is exactly the situation `steadyNewton()` is designed for: it solves the equilibrium equations directly and therefore finds the steady state regardless of its stability.

```{r find_steady}
params_f15 <- steadyNewton(params, effort = 1.5)
```

## Diagnosing the instability

Is this steady state stable? We compute its stability metrics with
`getStability()`.

```{r coupled_stability}
stab <- getStability(params_f15, effort = 1.5)
stab$max_real_part
stab$stable
```

The largest real part among the eigenvalues, $\text{Re}(\lambda) =
`r round(stab$max_real_part, 3)` > 0$, so the steady state is unstable:
a perturbation grows at that exponential rate, roughly a factor $e$ every
`r round(1 / stab$max_real_part)` years.

### What mizer's time step makes of it

The instability is a property of the model. Whether you *see* it in a simulation
depends on the time step, and `getDiscreteStability()` says by how much. It
reports the spectral radius $\max_i|\mu_i|$ of the one-step map; converting that
into a growth rate per year, $\log(\max_i|\mu_i|)/\Delta t$, puts it on the same
footing as `max_real_part`:

```{r discrete_stability}
growth_rate <- function(dt) {
    rho <- getDiscreteStability(params_f15, effort = 1.5, dt = dt)$spectral_radius
    log(rho) / dt
}
sapply(c(0.1, 0.5, 1), growth_rate)
```

Against the model's own `r round(stab$max_real_part, 3)` per year, mizer's
default `euler` step gets none of these right, and not in a tidy direction
either. At $\Delta t = 0.1$ the implicit transport solve damps the growing
oscillation almost exactly as fast as the model grows it, so the scheme sits at
the edge of stability and a century-long run stays flat — a steady state that
looks perfectly settled and is an artefact of the step size. At
$\Delta t = 0.5$ the map is firmly stable; at $\Delta t = 1$ it is unstable once
more, but oscillating with a period near 3.4 years that belongs to no mode of
the model.

This is the practical reason to keep an eye on `dt`, to prefer
`method = "tr_bdf2"` when the dynamics matter, and the reason `getStability()`
does not go through the one-step map at all.

To see *how* it is unstable, we look at the dominant eigenvalue (the one with the largest real part).

```{r dominant_mode}
stab$eigenvalues[1]
```

It is **complex** with a positive real part ($\lambda \approx `r round(Re(stab$eigenvalues[1]), 3)` + `r round(Im(stab$eigenvalues[1]), 3)`i$). A purely real, positive eigenvalue would signal monotone growth, but a complex pair with a positive real part means the perturbation grows *while oscillating* — a Hopf bifurcation. The period of this oscillation is set by the imaginary part of the eigenvalue,

$$ \text{Period} = \frac{2\pi}{|\text{Im}(\lambda)|} \text{ years}, $$

which mizer reports as `dominant_period`:

```{r dominant_period}
stab$dominant_period
```

So the linear analysis predicts a growing oscillation with a period of roughly `r round(stab$dominant_period, 2)` years.

## Watching the limit cycle emerge with `projectToSteady()`

We can confirm these predictions by projecting the dynamics with `projectToSteady()`.

Because the steady state is unstable, running the dynamics forward will never "converge" to a fixed point in the usual sense. `projectToSteady()` recognises this: it attaches a `"convergence"` attribute to its result that reports whether the trajectory settled on a stable fixed point or a limit cycle, along with its period and relative amplitude. By passing `return_sim = TRUE`, it returns a full `MizerSim` object covering the run.

```{r pts_trbdf2}
sim_cycle <- projectToSteady(params, effort = 1.5, t_max = 200, t_per = 0.2,
                             return_sim = TRUE, method = "tr_bdf2")
attr(sim_cycle, "convergence")
```

`projectToSteady()` correctly identifies the attractor as a limit cycle with a period of about `r round(attr(sim_cycle, "convergence")$period, 1)` years and a peak-to-trough biomass amplitude of about `r round(100 * attr(sim_cycle, "convergence")$amplitude)`% of the mean. Applied to a model with a genuine stable steady state (such as the default `params` at effort 0), the function reports `type = "below_tolerance"` instead, meaning the state stopped moving on the scale of the distance function:

```{r pts_steady}
attr(projectToSteady(params, t_max = 100), "convergence")$type
```

Because `return_sim = TRUE` returns a standard `MizerSim` object, we can plot the emergence of the limit cycle directly:

```{r plot_lcs_biomass2}
plotBiomass(sim_cycle)
```

Notice how well the detected nonlinear period (about `r round(attr(sim_cycle, "convergence")$period, 1)` years) agrees with the `dominant_period` of about `r round(stab$dominant_period, 2)` years from the linear analysis, even though the cycle we are looking at is anything but small: its relative amplitude is about `r round(100 * attr(sim_cycle, "convergence")$amplitude)`% of the mean. The linear analysis is a prediction about infinitesimal perturbations of the steady state, so the remaining difference of under six percent is what the nonlinearity does to the shape of the large-amplitude cycle. Close to the bifurcation point the two agree even more closely.

## Visualising the oscillation shape

Finally, `getLimitCycleSim()` constructs a `MizerSim` object covering one period of the oscillation *in the linear approximation*, using the eigenvector of the dominant oscillatory mode. This isolates the pure **shape** of the cycle — which species and sizes participate and with what phase — without the growth or the nonlinear distortion of the full run. Fish and resource are driven by the same eigenvector and the same scale factor, so the resource carries the phase the mode gives it: on this model it leads the total fish biomass by about 0.3 years of the 5.12-year period.

```{r limit_cycle}
lcs <- getLimitCycleSim(params_f15, amplitude = 0.1)
```

Because `lcs` is an ordinary `MizerSim` object, any of the standard mizer plotting functions can be used to explore it. For example, the biomass over one period:

```{r plot_lcs_biomass}
plotBiomass(lcs)
```

This linearised cycle gives a clear picture of the cohort-resonance pattern driving the oscillation — the mode shape that the full nonlinear limit cycle grows out of.

## A caveat: the analysis assumes a smooth model

Everything in this vignette rests on the one-step map being differentiable at
the steady state, so that the finite-difference Jacobian means something. That
holds for models built from mizer's own rate functions, but not necessarily for
a model carrying a custom rate registered with `setRateFunction()`. If such a
rate jumps as a function of the abundances, `steadyNewton()` may stall and
`getStability()` can return a plausible-looking number that describes neither
branch of the dynamics.

The cheapest check is to re-run `getStability()` with a different `h`: on a
smooth model `max_real_part` is unchanged to several figures. See
[Discontinuous rate functions](discontinuous_rates.html).
