---
title: "Dynamic stability and Hopf bifurcations"
output:
  html_document:
    number_sections: yes
    toc: yes
    fig_width: 6
    fig_height: 6
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

## Introduction

In a mizer model, a steady state represents an equilibrium where the abundance of every species at every size remains constant over time. While `steady()` attempts to find such a state by projecting the dynamics forward in time until changes stop, this only works if the steady state is *dynamically stable*.

If a steady state is unstable, small perturbations will grow rather than decay, and the system will naturally move away from the steady state, often settling into persistent oscillations (a limit cycle) or more complex dynamics.

Mizer provides the `steadyNewton()` function to find steady states irrespective of their stability by solving the algebraic equilibrium equations directly. Once a steady state is found, its stability can be analysed using `getStability()`.

This vignette explains how to perform this stability analysis, what the results mean, how mizer's steady-state finders detect and characterise limit cycles, and how to use the linearised limit cycle approximation provided by `getLimitCycleSim()` to understand emergent oscillations around a Hopf bifurcation.

## Linear stability analysis

Mizer projects the state of the ecosystem forward in discrete time steps. We can write this discrete-time map abstractly as:

$$ N(t+\Delta t) = G(N(t)) $$

where $N(t)$ represents the full state vector (abundances of all species at all sizes) at time $t$. At a steady state $N^*$, we have $N^* = G(N^*)$.

Because abundances in a mizer model span dozens of orders of magnitude across the size spectrum, evaluating stability in terms of absolute perturbations (e.g., $N(t) = N^* + \delta N(t)$) can be numerically poorly-scaled. Instead, it is more natural to consider **multiplicative (relative) perturbations**, where we perturb the steady state by a small relative amount $x(t)$:

$$ N_i(t) = N_i^* \left(1 + x_i(t)\right) $$

Linearising the dynamics for these small relative perturbations $x(t)$ gives:

$$ x(t+\Delta t) \approx K \, x(t) $$

where $K$ is the relative Jacobian matrix. The elements of $K$ describe the proportional change in species $i$ resulting from a proportional change in species $j$:

$$ K_{ij} = \frac{\partial \log G_i}{\partial \log N_j}(N^*) = \frac{N_j^*}{N_i^*} \frac{\partial G_i}{\partial N_j}(N^*) $$

Notice that $K$ is related to the standard absolute Jacobian $L$ (with elements $L_{ij} = \frac{\partial G_i}{\partial N_j}$) by a similarity transform $K = D^{-1} L D$, where $D$ is a diagonal matrix of the steady-state abundances $N^*$. Because $K$ and $L$ are similar matrices, **they have exactly the same eigenvalues** $\mu_i$. 

While a pure discrete-time stability analysis would check if the spectral radius $\max_i |\mu_i| < 1$, doing so evaluates the stability of the numerical scheme rather than the underlying mathematical model. Implicit numerical schemes like the one mizer uses introduce severe "temporal numerical diffusion" that artificially damps oscillations. To evaluate the true continuous-time stability, `getStability()` maps these discrete eigenvalues back to the exact continuous-time eigenvalues $\lambda_i$ using the algebraic relation $\lambda_i = (1 - 1/\mu_i) / \Delta t$.

**Why does this exact mapping work?**
Mizer solves the continuous PDEs of the size spectrum using a fully implicit backward Euler step:
$$ N(t+\Delta t) = N(t) + \Delta t \, \frac{dN}{dt}(t+\Delta t) $$
If we define the continuous Jacobian as $J = \partial(dN/dt)/\partial N$ and the discrete Jacobian of the one-step map as $L = \partial N(t+\Delta t)/\partial N(t)$, differentiating the implicit step yields:
$$ L = I + \Delta t \, J \, L \quad \implies \quad (I - \Delta t \, J) L = I \quad \implies \quad J = \frac{1}{\Delta t}(I - L^{-1}) $$
This relates their eigenvalues exactly by $\lambda_i = (1 - 1/\mu_i) / \Delta t$. By applying this mapping, we analytically strip away the temporal numerical diffusion of the implicit solver and recover the exact continuous-time stability.

The stability of the continuous-time mathematical steady state is determined by the real parts of these continuous eigenvalues. If the maximum real part is negative ($\max_i \text{Re}(\lambda_i) < 0$), every perturbation decays and the steady state is stable; if any real part is positive, some perturbation grows and it is unstable.

If you specifically want to study the numerical stability of mizer's Euler method for a particular time step, `getStability()` optionally accepts a `dt` argument. The discrete eigenvalues $\mu_i$ will then reflect the one-step map under that time step size, and the returned **spectral radius** $\max_i |\mu_i|$ determines numerical stability (stable if $<1$).

*(Note on numerical implementation: While analysing the relative Jacobian $K$ is conceptually natural, computing it requires dividing by $N_i^*$, which causes floating-point overflow for structurally zero size classes (e.g., extinct species or truncated tails). Therefore, `getStability()` numerically computes the absolute Jacobian $L$ using finite-difference steps scaled multiplicatively to the local density. This is mathematically equivalent to the relative analysis but avoids dividing by zero, ensuring robust eigenvalue computation. Fast decaying modes map to $\mu \approx 0$, which can amplify numerical noise during the $1/\mu$ inversion, so mizer safely filters these out to isolate the dominant continuous eigenvalues.)*

## Driving the North Sea model unstable

Let's load mizer and start with the standard North Sea model.

```{r setup}
suppressMessages(devtools::load_all("..", quiet = TRUE))
params <- NS_params
```

The North Sea model at its default fishing effort is dynamically stable, so it makes a rather uneventful example: perturb it and it simply settles back down. To see something more interesting we increase the fishing pressure, setting the effort of every gear to `1.5`. This pushes the model across a **Hopf bifurcation** into a regime where the steady state is unstable.

Because the steady state is now unstable, `steady()` cannot find it — projecting the dynamics forward would diverge away from it. This is exactly the situation `steadyNewton()` is designed for: it solves the equilibrium equations directly and therefore finds the steady state regardless of its stability.

```{r find_steady}
params_f15 <- steadyNewton(params, effort = 1.5)
```

## Diagnosing the instability

Is this steady state stable? We compute its stability metrics with `getStability()`.

```{r default_stability}
stab_default <- getStability(params_f15, effort = 1.5)
stab_default$spectral_radius
stab_default$max_real_part
```

Notice the distinction between discrete numerical stability and continuous physical stability:
- The **spectral radius** $\max |\mu| = `r round(stab_default$spectral_radius, 2)` < 1$, which means the numerical one-step map of the default backward Euler solver at $\Delta t = 1$ is discretely stable. The temporal numerical diffusion of backward Euler artificially damps the oscillation.
- Mapping back to continuous time reveals the true physical PDE stability with **`max_real_part`** $\text{Re}(\lambda) = `r round(stab_default$max_real_part, 2)` > 0$, showing that the continuous system is physically unstable.

To convert discrete eigenvalues $\mu$ of the one-step map to continuous eigenvalues $\lambda$, `getStability()` uses the Backward Euler transformation:
$$ \lambda = \frac{1 - 1/\mu}{\Delta t} $$
To ensure mathematical consistency across all state variables and prevent fast-decaying modes from producing spurious eigenvalue artifacts due to near-zero matrix inversion noise, `getStability()` evaluates the resource step using the same implicit Backward Euler scheme:
$$ N_R(t+\Delta t) = \frac{N_R(t) + \Delta t \, r_R(w) c_R(w)}{1 + \Delta t \, (r_R(w) + \mu_R(w))} $$

Although `project()` uses the exact analytical resource solution ($N_R(t+\Delta t) = N_R^* + (N_R(t) - N_R^*) e^{-(r_R + \mu_R)\Delta t}$) at each step, the discrete eigenvalues $\mu$ from `getStability()` accurately reflect the numerical step in `project(method = "euler")`. For fast resource modes ($r_R + \mu_R \gg 1$), both exact decay ($\mu_{exact} = e^{-(r_R + \mu_R)\Delta t} \approx 0$) and Backward Euler ($\mu_{BE} = \frac{1}{1 + (r_R + \mu_R)\Delta t} \approx 0$) damp out perturbations within a single timestep. For slow, coupled modes ($|\mu| \approx 1$) governing system stability, Taylor expanding around $\Delta t = 0$ shows that $\mu_{BE} = 1 - k\Delta t + \mathcal{O}(\Delta t^2)$ matches $\mu_{exact} = 1 - k\Delta t + \mathcal{O}(\Delta t^2)$ to first order.

Furthermore, by default `getStability()` treats the resource spectrum as a *quasi-static* fast variable that instantly equilibrates to the fish. Analysing the full coupled fish–resource system with `include_resource = TRUE` includes the resource dynamics:

```{r coupled_stability}
stab <- getStability(params_f15, effort = 1.5, include_resource = TRUE)
stab$spectral_radius
stab$max_real_part
stab$stable
```

With the full coupled fish–resource dynamics included, both the discrete numerical solver ($\max |\mu| = `r round(stab$spectral_radius, 2)` > 1$) and the continuous PDE ($\text{Re}(\lambda) = `r round(stab$max_real_part, 2)` > 0$) are unstable.

To see *how* it is unstable, we look at the dominant eigenvalue (the one with the largest real part).

```{r dominant_mode}
stab$eigenvalues[1]
```

It is **complex** with a positive real part ($\lambda \approx `r round(Re(stab$eigenvalues[1]), 2)` + `r round(Im(stab$eigenvalues[1]), 2)`i$). A purely real, positive eigenvalue would signal monotone growth, but a complex pair with a positive real part means the perturbation grows *while oscillating* — a Hopf (or Neimark–Sacker) bifurcation. The period of this oscillation is set by the imaginary part of the eigenvalue,

$$ \text{Period} = \frac{2\pi}{|\text{Im}(\lambda)|} \text{ years}, $$

which mizer reports as `dominant_period`:

```{r dominant_period}
stab$dominant_period
```

So the linear analysis predicts a growing oscillation with a period of roughly `r round(stab$dominant_period, 1)` years.

## Watching the limit cycle emerge with `projectToSteady()`

We can confirm these predictions by projecting the dynamics with `projectToSteady()`.

Because the steady state is unstable, running the dynamics forward will never "converge" to a fixed point in the usual sense. `projectToSteady()` recognises this: it attaches a `"convergence"` attribute to its result that reports whether the trajectory settled on a stable fixed point or a limit cycle, along with its period and relative amplitude. By passing `return_sim = TRUE`, it returns a full `MizerSim` object covering the run.

```{r pts_trbdf2}
sim_cycle <- projectToSteady(params, effort = 1.5, t_max = 200, t_per = 0.2,
                             return_sim = TRUE, method = "tr_bdf2")
attr(sim_cycle, "convergence")
```

`projectToSteady()` correctly identifies the attractor as a limit cycle with a period of about `r round(attr(sim_cycle, "convergence")$period, 1)` years and a peak-to-trough biomass amplitude of about `r round(100 * attr(sim_cycle, "convergence")$amplitude)`% of the mean. Applied to a model with a genuine stable steady state (such as the default `params` at effort 0), the function reports `type = "below_tolerance"` instead, meaning the state stopped moving on the scale of the distance function:

```{r pts_steady}
attr(projectToSteady(params, t_max = 100), "convergence")$type
```

Because `return_sim = TRUE` returns a standard `MizerSim` object, we can plot the emergence of the limit cycle directly:

```{r plot_lcs_biomass2}
plotBiomass(sim_cycle)
```

Notice that the detected nonlinear period (about `r round(attr(sim_cycle, "convergence")$period, 1)` years) is slightly shorter than the `dominant_period` of about `r round(stab$dominant_period, 1)` years from the linear PDE analysis. This is expected: at effort 1.5 we are beyond the bifurcation where nonlinearities reshape the large-amplitude cycle. Close to the bifurcation point the two agree closely; further from it the linear analysis remains an excellent qualitative guide, correctly predicting an oscillatory instability.

## Visualising the oscillation shape

Finally, `getLimitCycleSim()` constructs a `MizerSim` object covering one period of the oscillation *in the linear approximation*, using the leading complex eigenvector. This isolates the pure **shape** of the cycle — which species and sizes participate and with what phase — without the growth or the nonlinear distortion of the full run.

```{r limit_cycle}
lcs <- getLimitCycleSim(params_f15, include_resource = TRUE, amplitude = 1)
```

Because `lcs` is an ordinary `MizerSim` object, any of the standard mizer plotting functions can be used to explore it. For example, the biomass over one period:

```{r plot_lcs_biomass}
plotBiomass(lcs)
```

This linearised cycle gives a clear picture of the cohort-resonance pattern driving the oscillation — the mode shape that the full nonlinear limit cycle grows out of.

## A caveat: the analysis assumes a smooth model

Everything in this vignette rests on the one-step map being differentiable at
the steady state, so that the finite-difference Jacobian means something. That
holds for models built from mizer's own rate functions, but not necessarily for
a model carrying a custom rate registered with `setRateFunction()`. If such a
rate jumps as a function of the abundances, `steadyNewton()` may stall and
`getStability()` can return a plausible-looking number that describes neither
branch of the dynamics.

The cheapest check is to re-run `getStability()` with a different `h`: on a
smooth model the spectral radius is unchanged to several figures. See
[Discontinuous rate functions](discontinuous_rates.html).
