---
title: "Editing the Mizer Website"
output:
  html_document:
    number_sections: yes
    toc: yes
    fig_width: 5
    fig_height: 5
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


## Website built from package documentation
The mizer website at [https://sizespectrum.org/mizer/] is created
programmatically from the documentation included in the mizer GitHub
repository. Thus to edit the website just means editing this documentation.
The website creation is performed by the 
[`pkgdown`](https://pkgdown.r-lib.org/index.html) package. You can trigger
a build in your local repository by issuing the command
```{r eval=FALSE}
pkgdown::build_site()
```
This will convert all existing documentation into html files and put these
into the `docs` subdirectory of your mizer directory. In detail:

1. All .Rmd files in the `vignettes` subdirectory (and its subdirectories, if
   any) will produce html files of the same name in `docs/articles`.

2. All .Rd documentation files in `man` will produce html files of the same
   name in `docs/reference`.
   
4. The `NEWS` file will be converted to `docs/news/index.html`

5. The README.md file will be converted to `docs/index.html`.

Then the package will use the information in the `_pkgdown.yml` file to
create a navbar and index pages. The navbar is configured with
```
navbar:
  title: "mizer"
  left:
    - icon: fa-home fa-lg
      href: index.html
    - text: "Get Started"
      href: articles/mizer.html
    - text: "Articles"
      href: articles/index.html
    - text: "Reference"
      href: reference/index.html
    - text: "Publications"
      href: articles/publications.html
    - text: "News"
      href: news/index.html
  right:
    - icon: fa-github fa-lg
      href: https://github.com/sizespectrum/mizer
```
This is quite self-explanatory, specifying the order of the navbar entries,
their text and link destination, and whether they are on the left or right side.
It is clearly easy to add additional navbar entries.
It would be possible to also include drop-down items in navbar. See the
[pkgdown documentation](https://pkgdown.r-lib.org/articles/pkgdown.html#navigation-bar)
for more details.

The file `docs/articles/index.html` is created according to the article: 
section in the `_pkgdown_yml` file. At an early stage it looked as follows:
```
articles:
- title: User Guides
  contents:
  - model_description
- title: Developer Guides
  contents:
  - developer_vignette
  - mathematical_details
```
By now it has more entries to link to more vignettes. It
is not necessary to use this index file. The menu entry could just as well
point to a more elaborate file created from an .Rmd file in the vignette
subdirectory, which might become useful in the future.

The file `docs/reference/index.html` is created according to the reference:
section in the `_pkgdown_yml` file, with a similar syntax as the article:
section.

## Changing page content
At the top of every page on the website, except the homepage, there is
a link to the page source. So it is easy to know which file to edit to
make changes to a page. If you want to rebuild the website after making
a change to a single page you may not want to re-build everything. You can
rebuild just a single vignette. For example to rebuild this page after 
editing the vignette `vignette/editing_website.Rmd` you can call
```{r eval=FALSE}
pkgdown::build_article("editing_website")
```
To rebuild the reference pages after changing the roxygen code doc you can
run
```{r eval=FALSE}
pkgdown::build_reference()
```

To change the homepage you edit index.md. This file is very similar to 
README.md that is used on the mizer homepage on GitHub, but can contain
extra content like embedded videos that the GitHub homepage does not
support. After changes to index.md you can update your local copy of
the website with
```{r eval=FALSE}
pkgdown::build_home()
```

Of course these changes will affect only your local copy of the website.
To get those changes reflected on the online site you will need to make
a pull request for your changed files against the master branch of the upstream 
repository. 


## GitHub pages
The website is hosted with [GitHubPages](https://pages.github.com/). In the
settings pages of the mizer repository on GitHub the source is set to
"master branch/docs folder". This means that only the master branch controls
the website.


## Search bar
The search bar on the website is powered by Docsearch from Algolia.
This is a free service for open-source projects. The set-up is
explained in the 
[pkgdown documentation](https://pkgdown.r-lib.org/articles/pkgdown.html#search).


## Embedding videos
It is easy to embed videos into a webpage by putting an iframe into the
Rmarkdown source file. For example putting the line
```
<iframe width="560" height="315" src="https://www.youtube.com/embed/0RlXqLbFbWc" frameborder="0" allow="accelerometer; autoplay; encrypted-media; gyroscope; picture-in-picture" allowfullscreen></iframe>
```
into the source of this page produced the following embedded video:

<iframe width="560" height="315" src="https://www.youtube.com/embed/0RlXqLbFbWc" frameborder="0" allow="accelerometer; autoplay; encrypted-media; gyroscope; picture-in-picture" allowfullscreen></iframe>

The GitHub site itself (as opposed to GitHub Pages) does not support
embedded videos. 
