---
title: "Exploring The Simulation Results"
output: html_document
---

```{r label=lib, include=FALSE}
library(mizer)
params_data <- read.csv("NS_species_params.csv")
inter <- read.csv("inter.csv", row.names=1)
inter <- as(inter, "matrix")
params_data_gears <- params_data
params_data_gears$gear <- c("Industrial","Industrial","Industrial",
		      "Pelagic","Beam","Otter",
		      "Beam","Otter","Beam",
		      "Otter","Otter","Otter")
params_gears <- MizerParams(params_data_gears, interaction = inter)
gear_names <- c("Industrial","Pelagic","Beam","Otter")
times <- seq(from = 1, to = 10, by = 1)
effort_array <- array(NA, dim = c(length(times), length(gear_names)),
    dimnames = list(time = times, gear = gear_names))
effort_array[,"Industrial"] <- 0.5
effort_array[,"Pelagic"] <- seq(from = 1, to = 2, length = length(times))
effort_array[,"Beam"] <- seq(from = 1, to = 0, length = length(times))
effort_array[,"Otter"] <- seq(from = 1, to = 0.5, length = length(times))
```


# Exploring the simulation results {#sec:exploring}

In the previous sections we saw how to set up a model and project it forward through time under our desired fishing scenario. The result of running a projection is an object of class `MizerSim`. What do we then do?  How can we explore the results of the simulation? In this section we introduce a range of summaries, plots and indicators that can be easily produced using methods included in `mizer`.

We will use the same `MizerSim` object for these examples:

```{r label = exploration_simulation}
sim <- project(params_gears,effort=effort_array, dt=0.1, t_save = 1)
```

## Directly accessing the slots of `MizerSim objects`

We have seen in previous sections that a `MizerSim` object is made up of various *slots* that contain the original parameters (`species_params`), the population abundances (`n`), the abundance of the plankton spectrum (`n_pp`) and the fishing effort (`effort`).

The projected species abundances at size through time can be seen in the `n` slot.  This is a three-dimensional array (time x species x size).
Consequently, this array can get very big so inspecting it can be difficult.
In the example we have just run the time dimension of the `n` slot has `r dim(sim@n)[1]` rows (one for the initial population and then one for each of the saved time steps). There are also 12 species each with 100 sizes. We can check this by running the `dim()` function and looking at the dimensions of the `n` array:

```{r label=show_dim_n}
dim(sim@n)
```

To pull out the abundances of a particular species through time at size you can subset the array. For example to look at Cod through time you can use:

```{r results='hide'}
sim@n[,"Cod",]
```

This returns a two-dimensional array: time x size, containing the cod abundances. The time dimension depends on the value of the argument `t_save` when `project()` was run. You can see that even though we specified `dt` to be 0.1 when we called `project()`, the `t_save` argument had means that the output is only saved every time step.

The `n_pp` slot can be accessed in the same way.

## Summary methods for `MizerSim objects` {#sec:summaries}

As well as the `summary()` methods that are available for both `MizerParams` and `MizerSim` objects, there are some useful summary methods to pull information out of a `MizerSim` object (see Table~\@ref(tab:summary_methods)). All of these methods have help files to explain how they are used.
(It is also possible to use most of these methods with a `MizerParams` object if you also supply the population abundance as an argument. This can be useful for exploring how changes in parameter value or abundance can affect summary statistics and indicators. We won't explore this here but you can see their help files for more details.)

<!-- table was here -->

<!--

\begin{table}[ht]
    \begin{center}
	\begin{tabularx}{\linewidth}{r|X|X}
	\hline
	Method & Returns & Description \\
	\hline
	getSSB() & Two dimensional array (time x species) & Total Spawning Stock Biomass (SSB) of each species through time where SSB is calculated as the sum of weight of all mature individuals.\\
	getBiomass() & Two dimensional array (time x species) & Total biomass of each species through time.\\
	getN() & Two dimensional array (time x species) & Total abundance of each species through time.\\
	%getPhiPrey() & & \\
	getFeedingLevel() & Three dimensional array (time x species x size) & Feeding level of each species by size through time. \\
	%getPredRate() & &\\
	getM2() & Three dimensional array (time x species x size) & The predation mortality imposed on each species by size through time.\\
	getFMort() & Three dimensional array (time x species x size) & Total fishing mortality on each species by size through time.\\
	getFMortGear() & Four dimensional array (time x gear x species x size) & Fishing mortality on each species by each gear at size through time.\\
	%getZ() & & \\
	%getRDI & & \\
	%getRDD & & \\ % Need to add these
	getYieldGear() & Three dimensional array (time x gear x species) & Total yield by gear and species through time. \\
	getYield() & Two dimensional array (time x species) & Total yield of each species across all gears through time. \\
	\hline
	\end{tabularx}
	\caption{Summary methods for `MizerSim` objects.}
	 {#tab:summary_methods}
    \end{center}
\end{table}

-->

The methods `getBiomass()` and `getN()` have additional arguments that allow the user to set the size range over which to calculate the summary statistic.  This is done by passing in a combination of the arguments `min_l`, `min_w`, `max_l` and `max_w` for the minimum and maximum length or weight.

If `min_l` is specified there is no need to specify `min_w` and so on. However, if a length is specified (minimum or maximum) then it is necessary for the species parameter data.frame (see Section~\@ref(sec:species_parameters_dataframe)) to include the parameters `a` and `b` for length-weight conversion.
It is possible to mix length and weight constraints, e.g. by supplying a minimum weight and a maximum length. The default values are the minimum and maximum weights of the spectrum, i.e. the full range of the size spectrum is used.


### Examples of using the summary methods {#sec:summaries_example}

Here we show a simple demonstration of using a summary method using the `sim` object we created earlier.
Here, we use `getSSB()` to calculate the SSB of each species through time (note the use of the `head()` function to only display the first few rows).

```{r label = getSSB_demo}
ssb <- getSSB(sim)
dim(ssb)
head(ssb)
```

As mentioned above, we can specify the size range for the `getBiomass()` and `getN()` methods. For example, here we calculate the total biomass of each species but only include individuals that are larger than 10 g and smaller than 1000 g.

```{r label = getBiomass_demo}
biomass <- getBiomass(sim, min_w = 10, max_w = 1000)
head(biomass)
```

## Methods for calculating indicators

Methods are available to calculate a range of indicators from a `MizerSim` object after a projection.
These can be seen in Table~\@ref(tab:indicator_methods).
You can read the help pages for each of the methods for full instructions on how to use them, along with examples.

With all of the methods in the table it is possible to specify the size range of the community to be used in the calculation (e.g. to exclude very small or very large individuals) so that the calculated metrics can be compared to empirical data. This is used in the same way that we saw with the method `getBiomass()` in Section~\@ref(sec:summaries).
It is also possible to specify which species to include in the calculation.  See the help files for more details.

<!-- table was here -->

<!--

\begin{table}[ht]
    \begin{center}
	\begin{tabularx}{\linewidth}{r|X|X}
	\hline
	Method & Returns & Description \\
	\hline
	getProportionOfLargeFish() & A vector with values at each time step. & Calculates the proportion of large fish through time. The threshold value can be specified. It is possible to calculation the proportion of large fish based on either length or weight. \\
	getMeanWeight() & A vector with values at each saved time step.  & The mean weight of the community through time. This is calculated as the total biomass of the community divided by the total abundance.\\
	getMeanMaxWeight() & Depends on the `measure` argument. If `measure` = ``both'' then you get a matrix with two columns, one with values by numbers, the other with values by biomass at each saved time step. If `measure` = ``numbers'' or ``biomass'' you get a vector of the respective values at each saved time step. & The mean maximum weight of the community through time. This can be calculated by numbers or by biomass. See the help file for more details.\\
	getCommunitySlope() & A data.frame with four columns: time step, slope, intercept and $R^2$ value. & Calculates the slope of the community abundance spectrum through time by performing a linear regression on the logged total numerical abundance and logged body size. \\
	\hline
	\end{tabularx}
	\caption{Indicator methods for `MizerSim` objects.}
	 {#tab:indicator_methods}
    \end{center}
\end{table}

-->

### Examples of calculating indicators {#sec:indicators_example}

For these examples we use the `sim` object we created earlier.

The slope of the community can be calculated using the `getCommunitySlope()` method. Initially we include all species and all sizes in the calculation (only the first five rows are shown):

```{r label=getCommunitySlope_demo}
slope <- getCommunitySlope(sim)
head(slope)
```

This gives the slope, intercept and $R^2$ value through time (see the help file for `getCommunitySlope` for more details).

We can include only the species we want with the `species` argument. Here we only include demersal species.  We also restrict the size range of the community that is used in the calculation to between 10 g and 5 kg.
The `species` is a character vector of the names of the species that we want to include in the calculation.

```{r label = getCommunitySlope_with_args}
dem_species <- c("Dab","Whiting","Sole","Gurnard","Plaice","Haddock",
    "Cod","Saithe")
slope <- getCommunitySlope(sim, min_w = 10, max_w = 5000, 
    species = dem_species)
head(slope)
```

## Plotting the results

R is very powerful when it comes to exploring data through plots.  A useful package for plotting is `ggplot2`.  `ggplot2` uses data.frames for input data. Many of the summary methods and slots of the `mizer` classes are arrays or matrices. Fortunately it is straightforward to turn arrays and matrices into data.frames using the command `melt` which is in the `reshape2` package.
Although `mizer` does include some dedicated plots, it is definitely worth your time getting to grips with these `ggplot2` and other plotting packages. This will make it possible for you to make your own plots.

Included in `mizer` are several dedicated plots that use `MizerSim` objects as inputs (see Table~\@ref(tab:plots)).
As well as displaying the plots, these methods all return objects of type `ggplot` from the `ggplot2` package meaning that they can be further modified by the user (e.g. by changing the plotting theme).
See the help page of the individual plot methods for more details.
The generic `plot()` method has also been overloaded for `MizerSim` objects. This produces several plots in the same window to provide a snapshot of the results of the simulation.

Some of the plots plot values by size (for example `plotFeedingLevel()` and `plotSpectra()`).
For these plots, the default is to use the data at the final time step of the projection.
With these plotting methods, it is also possible to specify a different time step or a time range to average the values over before plotting. 

<!-- table was here -->

<!--

\begin{table}[ht]
    \begin{center}
	\begin{tabularx}{\linewidth}{r|X}
	\hline
	Plot & Description \\
	\hline
	plotBiomass() & Plots the total biomass of each species through time. A time range to be plotted can be specified. The size range of the community can be specified in the same way as the method `getBiomass()`.\\
	plotSpectra() & Plots the abundance (biomass or numbers) spectra of each species and the plankton community. It is possible to specify a minimum size which is useful for truncating the plot. \\
	plotFeedingLevel() & Plots the feeding level of each species against size.\\
	plotM2() & Plots the predation mortality of each species against size.\\
	plotFMort() & Plots the total fishing mortality of each species against size.\\
	plotYield() & Plots the total yield of each species across all fishing gears against time.\\
	plotYieldGear() & Plots the total yield of each species by gear against time.\\
	plot() & Produces 5 plots (`plotFeedingLevel()`, `plotBiomass()`, `plotM2()`, `plotFMort()` and `plotSpectra()`) in the same window as a summary.\\
	\hline
	\end{tabularx}
	\caption{Plot methods for `MizerSim` objects.}
	 {#tab:plots}
    \end{center}
\end{table}

-->

### Plotting examples {#sec:plot_examples}

Using the plotting methods is straightforward. For example, to plot the total biomass of each species against time you use the `plotBiomass()` method:

```{r label=plotBiomass, fig.keep='last', fig.cap="An example of using the plotBiomass() method."}
plotBiomass(sim)
```

As mentioned above, some of the plot methods plot values against size at a point in time (or averaged over a time period).
For these plots it is possible to specify the time step to plot, or the time period to average the values over. The default is to use the final time step.
Here we plot the abundance spectra (biomass), averaged over time = 5 to 10 (see Figure~\@ref(fig:plotSpectra_example)):

```{r label=plotSpectra_example, fig.keep='last', fig.cap="An example of using the plotSpectra() method, plotting values averaged over the period t = 5 to 10."}
plotSpectra(sim, time_range = 5:10, biomass=TRUE)
```


As mentioned above, and as we have seen several times in this vignette, the generic `plot()` method has also been overloaded.  This produces 5 plots in the same window (`plotFeedingLevel()`, `plotBiomass()`, `plotM2()`, `plotFMort()` and `plotSpectra()`, see Figure~\@ref(fig:plot_example)). It is possible to pass in the same arguments that these individual plots use, e.g. arguments to change the time period over which the data is averaged.

```{r label=demo_summary_plot, fig.cap="Example output from using the summary plot() method."}
plot(sim)
```

