---
title: "Exploring the Simulation Results"
output:
  html_document:
    toc: yes
    fig_width: 5
    fig_height: 5
---

```{r include=FALSE}
library(mizer)
gear_names <- c("Industrial","Pelagic","Beam","Otter")
times <- seq(from = 1, to = 10, by = 1)
effort_array <- array(NA, dim = c(length(times), length(gear_names)),
    dimnames = list(time = times, gear = gear_names))
effort_array[,"Industrial"] <- 0.5
effort_array[,"Pelagic"] <- seq(from = 1, to = 2, length = length(times))
effort_array[,"Beam"] <- seq(from = 1, to = 0, length = length(times))
effort_array[,"Otter"] <- seq(from = 1, to = 0.5, length = length(times))
```


## Introduction {#sec:exploring}

In the sections on [the multispecies model](multispecies_model.html) and on
[running a simulation](running_a_simulation.html) we saw how to set up a model
and project it forward through time under our desired fishing scenario. The
result of running a projection is an object of class `MizerSim`. What do we then
do?  How can we explore the results of the simulation? In this section we
introduce a range of summaries, plots and indicators that can be easily produced
using functions included in mizer.

We will use the following `MizerSim` object for these examples, where the
effort array is the one we created in the 
[previous section on running a simulation](running_a_simulation.html):

```{r label = exploration_simulation}
sim <- project(NS_params, effort = effort_array, dt = 0.1, t_save = 1)
```

## Accessing the simulation results

The projected species abundances at size through time can be obtained with 
`N(sim)`.  This returns a three-dimensional array (time x species x size). Consequently,
this array can get very big so inspecting it can be difficult. In the example we
have just run, the time dimension of `n` has `r dim(N(sim))[1]` rows
(one for the initial population and then one for each of the saved time steps).
There are also 12 species each with 100 sizes. We can check this by running the
`dim()` function and looking at the dimensions of the `n` array:

```{r label=show_dim_n}
dim(N(sim))
```

To pull out the abundances of a particular species through time at size you can
subset the array. For example to look at Cod through time you can use:

```{r results='hide'}
N(sim)[, "Cod", ]
```

This returns a two-dimensional array: time x size, containing the cod
abundances. The time dimension depends on the value of the argument `t_save`
when `project()` was run. You can see that even though we specified `dt` to be
0.1 when we called `project()`, the `t_save = 1` argument has meant that the
output is only saved every year.

Often we are particularly interested in the results at the final time-step.
These we can access with
```{r results='hide'}
finalN(sim)
```
which is a two dimensional array (species x size).

The projected resource abundances can be accesses similarly with
```{r results='hide'}
NResource(sim)
```
This returns a two-dimensional array (time x size). And if we are only
interested in the final time step
```{r results='hide'}
finalNResource(sim)
```
returns a vector with one entry for each size class.

## Summary functions{#sec:summaries}

As well as the `summary()` methods that are available for both `MizerParams` and
`MizerSim` objects, there are other useful summary functions to pull information
out of a `MizerSim` object. A description of the different summary functions
available is given in 
[the summary functions help page.](../reference/summary_functions.html)

All of these functions have help files to explain how they are used. (It is also
possible to use most of these functions with a `MizerParams` object if you also
supply the population abundance as an argument. This can be useful for exploring
how changes in parameter value or abundance can affect summary statistics and
indicators. We won't explore this here but you can see their help files for more
details.)

The functions `getBiomass()` and `getN()` have additional arguments that allow the
user to set the size range over which to calculate the summary statistic.  This
is done by passing in a combination of the arguments `min_l`, `min_w`, `max_l`
and `max_w` for the minimum and maximum length or weight.
If `min_l` is specified there is no need to specify `min_w` and so on. However,
if a length is specified (minimum or maximum) then it is necessary for the
species parameter data.frame (see 
[the species parameters section](multispecies_model.html#sec:species_parameters_dataframe)) 
to include
the parameters `a` and `b` for length-weight conversion. It is possible to mix
length and weight constraints, e.g. by supplying a minimum weight and a maximum
length. The default values are the minimum and maximum weights of the spectrum,
i.e. the full range of the size spectrum is used.


### Examples of using the summary functions {#sec:summaries_example}

Here we show a simple demonstration of using a summary function using the `sim`
object we created earlier. Here, we use `getSSB()` to calculate the SSB of each
species through time (note the use of the `head()` function to only display the
first few rows).

```{r label = getSSB_demo}
ssb <- getSSB(sim)
dim(ssb)
head(ssb)
```

As mentioned above, we can specify the size range for the `getsummaryBiomass()` and
`getN()` functions. For example, here we calculate the total biomass of each
species but only include individuals that are larger than 10 g and smaller than
1000 g.

```{r label = getBiomass_demo}
biomass <- getBiomass(sim, min_w = 10, max_w = 1000)
head(biomass)
```


## Functions for calculating indicators

Functions are available to calculate a range of indicators from a `MizerSim`
object after a projection. A description of the different indicator functions
available is given in [the indicator functions help page.](../reference/indicator_functions.html).
You can read the help pages for each
of the functions for full instructions on how to use them, along with examples.

With all of the functions in the table it is possible to specify the size range of
the community to be used in the calculation (e.g. to exclude very small or very
large individuals) so that the calculated metrics can be compared to empirical
data. This is used in the same way that we saw with the function `getBiomass()` in
[the section on summary functions for MizerSim objects.](exploring_the_simulation_results.html#sec:summaries).
It is also
possible to specify which species to include in the calculation.  See the help
files for more details.


### Examples of calculating indicators {#sec:indicators_example}

For these examples we use the `sim` object we created earlier.

The slope of the community can be calculated using the `getCommunitySlope()`
function. Initially we include all species and all sizes in the calculation (only
the first five rows are shown):

```{r label=getCommunitySlope_demo}
slope <- getCommunitySlope(sim)
head(slope)
```

This gives the slope, intercept and $R^2$ value through time (see the help file
for `getCommunitySlope` for more details).

We can include only the species we want with the `species` argument. Below we
only include demersal species.  We also restrict the size range of the community
that is used in the calculation to between 10 g and 5 kg. The `species` argument is a
character vector of the names of the species that we want to include in the
calculation.

```{r label = getCommunitySlope_with_args}
dem_species <- c("Dab", "Whiting", "Sole", "Gurnard", "Plaice", "Haddock",
    "Cod", "Saithe")
slope <- getCommunitySlope(sim, min_w = 10, max_w = 5000, 
    species = dem_species)
head(slope)
```

## Plotting the results

R is very powerful when it comes to exploring data through plots.  Two useful
packages for plotting are `ggplot2` and `plotly`.  These use data.frames for input data
whereas many of the mizer functions return arrays or
matrices. Fortunately it is straightforward to turn arrays and matrices into
data.frames using the `melt()` function from the `reshape2` package that mizer
makes available to you.
Although `mizer` does include some dedicated plots, it is definitely worth your
time getting to grips with these other plotting packages. This
will make it possible for you to make your own plots. We provide some details 
in the section on [using ggplot2 and plotly with mizer](plotting.html).

Included in `mizer` are several dedicated plots that use `MizerSim` objects as
inputs (see [the plots help page.](../reference/plotting_functions.html)). As
well as displaying the plots, these functions all return objects of type `ggplot`
from the `ggplot2` package, meaning that they can be further modified by the user
(e.g. by changing the plotting theme). See the help page of the individual plot
functions for more details. The generic `plot()` method has also been overloaded
for `MizerSim` objects. This produces several plots in the same window to
provide a snapshot of the results of the simulation.

Some of the plots plot values by size (for example `plotFeedingLevel()` and
`plotSpectra()`). For these plots, the default is to use the data at the final
time step of the projection. With these plotting functions, it is also possible
to specify a different time, or a time range to average the values over before
plotting.



### Plotting examples {#sec:plot_examples}

Using the plotting functions is straightforward. For example, to plot the total
biomass of each species against time you use the `plotBiomass()` function:

```{r label=plotBiomass, fig.keep='last'}
plotBiomass(sim)
```

As mentioned above, some of the plot functions plot values against size at a
point in time (or averaged over a time period). For these plots it is possible
to specify the time step to plot, or the time period to average the values over.
The default is to use the final time step. Here we plot the abundance spectra
(biomass), averaged over time = 5 to 10:

```{r label=plotSpectra_example, fig.keep='last'}
plotSpectra(sim, time_range = 5:10)
```


As mentioned above, and as we have seen several times in this guide, the generic
`plot()` method has also been overloaded.  This produces 5 plots in the same
window (`plotFeedingLevel()`, `plotBiomass()`, `plotPredMort()`, `plotFMort()`
and `plotSpectra()`). It is possible to pass in the same arguments that these
individual plots use, e.g. arguments to change the time period over which the
data is averaged.

```{r label=demo_summary_plot}
plot(sim)
```

The next section describes how to use what we have learned to 
[model the North Sea.](a_multispecies_model_of_the_north_sea.html) 
