---
title: "Extending mizer"
output:
  html_document:
    toc: yes
    fig_width: 5
    fig_height: 5
vignette: >
  %\VignetteIndexEntry{Extending mizer}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(mizer)
```

# Overview

This vignette is for users who want to change how mizer works without editing
the package source code. There are three extension routes:

1. Use `setRateFunction()` when you want to replace one of the standard rate
   calculations such as encounter, growth, mortality, or recruitment.
2. Use `setComponent()` when you want to add a new ecosystem component such as
   detritus, carrion, or an extra resource pool.
3. Use `customFunction()` only as a last resort when neither of the first two
   mechanisms is sufficient.

The safest workflow is to start with the smallest change that can express your
idea. In particular, if you only need to change one rate, prefer
`setRateFunction()` over replacing `mizerRates()` or patching internal mizer
functions.

# Choosing an extension mechanism

The following table is a quick guide.

```{r mechanism-table, echo = FALSE}
mechanisms <- data.frame(
    Goal = c(
        "Change one built-in rate calculation",
        "Add a new state variable or ecosystem pool",
        "Store parameters needed by your custom code",
        "Replace arbitrary internal mizer code"
    ),
    Use = c(
        "`setRateFunction()`",
        "`setComponent()`",
        "`other_params()` or `component_params`",
        "`customFunction()`"
    ),
    Notes = c(
        "Best option for time-dependent or alternative rate formulations.",
        "Lets you add component dynamics and optional encounter or mortality contributions.",
        "Use `other_params()` for model-wide parameters and `component_params` for one component.",
        "Experimental and fragile. Use only if the standard mechanisms cannot express your change."
    ),
    check.names = FALSE
)
knitr::kable(mechanisms)
```

## `setRateFunction()`

Use `setRateFunction()` when your model still fits mizer's standard flow of
rate calculations, but one step should be done differently. Examples include:

- making encounter or mortality explicitly time-dependent,
- using an alternative growth or reproduction formulation,
- swapping in a different recruitment density-dependence,
- replacing the whole `mizerRates()` pipeline with a custom version.

Your function is registered by name:

```{r eval = FALSE}
params <- setRateFunction(params, "Mort", "myMort")
```

The function must be available by name in the global environment or in an
installed package. mizer stores the function name, not the function object.

## `setComponent()`

Use `setComponent()` when your model needs an additional dynamical quantity that
is not already represented in `MizerParams`, for example detritus, carrion, an
oxygen pool, or a second resource spectrum.

A component can contribute to the model in up to three ways:

- `dynamics_fun` updates the component itself during projection,
- `encounter_fun` adds to `getEncounter()`,
- `mort_fun` adds to `getMort()`.

The component state can be any R object. For example it can be a scalar, a
vector on the resource size grid, or a list with several fields.

## `customFunction()`

Use `customFunction()` only if you have confirmed that your goal cannot be
expressed with `setRateFunction()`, `setComponent()`, `resource_dynamics()<-`,
or `setReproduction()`. It replaces an internal mizer function in the package
namespace and can easily break the package if your replacement is not fully
compatible.

# How custom rate functions are called

Each overridable rate has its own expected function signature and return shape.
The most important point is that custom rate functions should return plain
numeric objects, not `ArraySpeciesBySize` or `ArraySpeciesByTime` objects. The
`get*()` wrappers add those classes afterwards where appropriate.

The table below summarises the required inputs and outputs.

```{r rate-table, echo = FALSE}
rate_table <- data.frame(
    Rate = c("Encounter", "FeedingLevel", "EReproAndGrowth", "ERepro",
             "EGrowth", "PredRate", "PredMort", "FMort", "Mort", "RDI",
             "RDD", "ResourceMort", "Rates"),
    Signature = c(
        "function(params, n, n_pp, n_other, t, ...)",
        "function(params, n, n_pp, n_other, t, encounter, ...)",
        "function(params, n, n_pp, n_other, t, encounter, feeding_level, ...)",
        "function(params, n, n_pp, n_other, t, e, ...)",
        "function(params, n, n_pp, n_other, t, e, e_repro, ...)",
        "function(params, n, n_pp, n_other, t, feeding_level, ...)",
        "function(params, n, n_pp, n_other, t, pred_rate, ...)",
        "function(params, n, n_pp, n_other, t, effort, e_growth, pred_mort, ...)",
        "function(params, n, n_pp, n_other, t, f_mort, pred_mort, ...)",
        "function(params, n, n_pp, n_other, t, e_growth, mort, e_repro, ...)",
        "function(rdi, species_params, params, t, ...)",
        "function(params, n, n_pp, n_other, t, pred_rate, ...)",
        "function(params, n, n_pp, n_other, t, effort, rates_fns, ...)"
    ),
    `Return value` = c(
        "numeric matrix, species x size",
        "numeric matrix, species x size",
        "numeric matrix, species x size",
        "numeric matrix, species x size",
        "numeric matrix, species x size",
        "numeric matrix, species x full size grid",
        "numeric matrix, species x size",
        "numeric matrix, species x size",
        "numeric matrix, species x size",
        "numeric vector, one value per species",
        "numeric vector, one value per species",
        "numeric vector, one value per full size bin",
        "named list with all standard rate components"
    ),
    check.names = FALSE
)
knitr::kable(rate_table)
```

## Common arguments

Most custom rate functions use the same core arguments:

- `params`: the `MizerParams` object,
- `n`: species abundance by species and size,
- `n_pp`: the resource abundance on the full size grid,
- `n_other`: a named list of additional component states,
- `t`: the current simulation time.

Some rates also receive prerequisite rates that were already calculated earlier
in the pipeline, for example `encounter`, `feeding_level`, or `pred_mort`.

## Dimensions and dimnames

Most size-resolved rates should return a matrix with the same dimensions as
`initialN(params)`, so a species x size matrix with the same dimnames. The two
important exceptions are:

- `PredRate`, which uses the full prey size grid and therefore returns
  species x `w_full`,
- `ResourceMort`, which returns one value per `w_full` bin rather than one
  value per species.

At registration time `setRateFunction()` calls your function with test inputs
and checks that the returned object has the expected dimensions. This catches
many mistakes immediately.

# Worked example: a custom encounter function

The simplest way to start is to modify one rate while delegating most of the
work back to the built-in mizer function. The example below adds a sinusoidal
seasonal multiplier to the standard encounter rate.

```{r seasonal-encounter-setup}
params <- NS_params
other_params(params) <- list(
    season_amplitude = 0.2,
    season_period = 1
)
```

```{r seasonal-encounter-fun}
seasonalEncounter <- function(params, n, n_pp, n_other, t, ...) {
    p <- other_params(params)
    multiplier <- 1 + p$season_amplitude * sin(2 * pi * t / p$season_period)
    multiplier * mizerEncounter(params, n = n, n_pp = n_pp, n_other = n_other,
                                t = t, ...)
}
```

Register the function and inspect the result:

```{r seasonal-encounter-register}
params2 <- setRateFunction(params, "Encounter", "seasonalEncounter")
```

```{r seasonal-encounter-check}
enc0 <- getEncounter(params2, t = 0)
enc_quarter <- getEncounter(params2, t = 0.25)
range(enc_quarter / enc0, na.rm = TRUE)
```

This pattern is often the easiest way to extend mizer safely:

- fetch custom parameters from `other_params(params)`,
- compute a modifier,
- call the standard mizer rate function,
- return a plain numeric matrix of the same shape.

## Testing a custom rate function

When you develop an extension package, add tests that compare your custom rate
function to the built-in behaviour in a simple case. For the seasonal example a
good first test would check:

- at `t = 0`, the result equals `mizerEncounter()`,
- at `t = 0.25`, the result is scaled by `1 + season_amplitude`,
- the returned object has the same dimensions and dimnames as `initialN(params)`.

# Worked example: adding a detritus-like component

Now suppose you want to add a component that is consumed by fish and slowly
relaxes back to a carrying capacity. We will store the component on the full
resource size grid so that it can be used like an extra prey spectrum.

```{r detritus-functions}
detritusEncounter <- function(params, n, n_pp, n_other, component, ...) {
    params2 <- params
    params2@other_encounter[[component]] <- NULL
    mizerEncounter(params2, n = n, n_pp = n_other[[component]],
                   n_other = n_other, ...)
}

detritusDynamics <- function(params, n_other, rates, dt, component, ...) {
    detritus <- n_other[[component]]
    p <- params@other_params[[component]]
    interaction <- params@species_params$interaction_resource
    mort <- as.vector(interaction %*% rates$pred_rate)
    target <- p$rate * p$capacity / (p$rate + mort)
    target - (target - detritus) * exp(-(p$rate + mort) * dt)
}
```

```{r detritus-component}
detritus_params <- list(
    capacity = initialNResource(params),
    rate = params@rr_pp
)

params3 <- setComponent(
    params,
    component = "detritus",
    initial_value = initialNResource(params) / 2,
    dynamics_fun = "detritusDynamics",
    encounter_fun = "detritusEncounter",
    component_params = detritus_params
)
```

Once the component has been added:

- its initial state is available via `initialNOther(params3)$detritus`,
- its settings can be inspected with `getComponent(params3, "detritus")`,
- its contribution is included automatically in `getEncounter()` and `getDiet()`.

For example:

```{r detritus-diet}
diet <- getDiet(params3, proportion = FALSE)
tail(dimnames(diet)$prey, 3)
```

If you also want the component to contribute to mortality, add a `mort_fun`
when calling `setComponent()`.

# Practical advice for extension authors

## Start from a built-in function

If possible, copy the relevant built-in mizer function, make one change, and
register the modified version. This makes it much easier to stay compatible
with the rest of the rate pipeline.

## Keep model-wide and component-specific parameters separate

Use:

- `other_params(params)` for parameters used by custom rate functions across
  the whole model,
- `component_params` in `setComponent()` for parameters belonging to one
  component only.

This keeps the structure of `params@other_params` readable and makes debugging
much easier.

## Return plain numeric objects

If you replace a built-in rate function, return a plain matrix, vector, or list
with the expected dimensions. Do not return `ArraySpeciesBySize` objects from
custom rate functions that are called through `setRateFunction()`.

## Preserve dimnames where possible

Many bugs in extension code come from returning the right numeric values with
the wrong shape. A good habit is to build outputs from existing mizer arrays so
that dimensions and dimnames are inherited automatically.

## Test registration and use

Test both of these:

1. that `setRateFunction()` or `setComponent()` succeeds,
2. that the downstream function you care about, such as `getEncounter()`,
   `getDiet()`, or `project()`, uses your extension as intended.

# When to move beyond a local script

If your extension becomes useful across several projects, consider turning it
into a small package. That gives you:

- a stable namespace so mizer can find your functions by name,
- a natural place for unit tests and documentation,
- versioning that can be recorded in `params@extensions`.

For larger changes you may also want to discuss your design on the
[mizer issue tracker](https://github.com/sizespectrum/mizer/issues) before
committing to an interface.

# See also

- `vignette("developer_vignette")` for a broader guide to working with the
  mizer code base,
- `?setRateFunction`,
- `?setComponent`,
- `?customFunction`,
- `?getRates`.
