---
title: "Installing Mizer"
output:
  html_document:
    number_sections: yes
    toc: yes
    fig_width: 5
    fig_height: 5
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Installing R and RStudio

Mizer is compatible with R versions 3.1 and later. You can install R on your
computer by following the instructions at [https://cran.r-project.org/] for your
particular platform.

This guide assumes that you will be using RStudio to work with R. There is
really no reason not to use RStudio and it makes a lot of things much easier.
RStudio develops rapidly and adds useful features all the time and so it pays 
to upgrade to the [latest version](https://www.rstudio.com/products/rstudio/download/preview/) 
frequently. This guide was written with version 1.2.1268.

## Installing the mizer package

The easiest way to install the mizer package (assuming you have an active internet
connection) is to start an session and then type:

```{r, eval=FALSE}
install.packages("mizer")
```

After installing mizer, to actually use it you need to load the package using
the `library()` function. Note that whilst you only need to install the
package once, it will need to be loaded every time you start a new R session.

```{r, eval=FALSE}
library(mizer)
```

## Installing development version
The source code for mizer is hosted on
[Github.](https://github.com/sizespectrum/mizer) If you are feeling brave and
wish to try out a development version of mizer you can install the package from
here using the R package devtools (which was used extensively in putting
together mizer). If you have not yet installed devtools, do
```{r, eval=FALSE}
install.packages("devtools")
```
Then you can install the latest version from GitHub using
```{r, eval=FALSE}
devtools::install_github("sizespectrum/mizer")
```
Using the same `devtools::install_github()` function you can also install code
from forked mizer repositories or from other branches on the official
repository.


## Getting help in R

As you probably know, to access documentation in R you can use the `help()`
function. This can be used to access package documentation in a range of ways.
For example:

```{r, eval=FALSE}
? mizer
help(package="mizer")
? project
```

The first command gives a brief introduction to mizer.
The second command gives a list of all available functions.
The third gives the documentation page for the function `project()`.

Some methods are associated with several R classes. For example, the
`plot()` method is generic and can be used on a wide range of R objects.
The `plot()` method has been overloaded in mizer to plot MizerSim objects. 
To select the help page for the appropriate plotting method you can either 
use `?? plot` and then select from a menu of packages or you can supply
the package name in the `help()` command:

```{r, eval=FALSE}
help(plot, package="mizer")
```
To access the the help page of a particular function you can use:
`? function-name`. For example, to access the help page of the
`getFeedingLevel()` function you can use:

```{r, eval=FALSE}
? getFeedingLevel
```

Slightly different syntax is used to get information on a class. For example,
to get the help page on the MizerParams class you can use:

```{r, eval=FALSE}
class ? MizerParams
```