---
title: "The General Mizer Size-spectrum Model"
output:
  html_document:
    number_sections: yes
    toc: yes
    fig_width: 5
    fig_height: 5
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

On this page we will revisit the presentation of the mizer model given
in chapter 3 of the main [mizer vignette](https://cran.r-project.org/web/packages/mizer/vignettes/mizer_vignette.pdf)
but taking care to separate the essential features of the model that are
hard-coded and the various possible specialisations for which mizer provides
setup functions. We will provide links to the functions that can be used to set
or change the various model parameters (all collected together in `setParams()`)
as well as to the functions that calculate the various ecological rates in the
model (all collected together in `getRates()`) because the help pages of these
functions will provide useful additional details.

We will not go into detail of how this model is 
realised in code. Such detail will be provided in the [developer guide](developer_vignette.html). However some details are hidden on this page
and you can see them by clicking on links like the following:
<details>
And clicking again will hide the details again.
</details>



## Consumer size spectra
The model framework builds on two central assumption and a number of lesser 
standard assumption.  

The first central assumption is that an individual can be characterized by its 
weight $w$ and its species number $i$ only. The aim of the model is to calculate 
the size spectrum $N_i(w)$, which is the *density* of individuals of
species $i$ such that $\int_w^{w+dw}N_i(w)dw$ is the *number* of individuals of
species $i$ in the size interval $[w,w+dw]$. In other word: the number of 
individuals in a size range is the area under the number density $N_i(w)$.

<details>
Here is a plot of an example size spectrum for two species with
$N_i(w)$ on the vertical axis for $i=1,2$ and $w$ on the horizontal axis.
```{r, message=FALSE}
library(mizer)
params <- set_scaling_model(no_sp = 2, min_egg = 1e-3)
plotSpectra(params, plankton = FALSE, power = 0)
```

To represent this continuous size spectrum in the computer, the size
variable $w$ is discretized into a vector `w` of discrete weights,
providing a grid of sizes spanning the range from the smallest egg size
to the largest asymptotic size. These grid values divide the full size
range into a finite number of size bins. The size bins should be chosen
small enough to avoid the discretisation errors from becoming too big.

The weight grid is set up to be logarithmically spaced, so that
`w[j]=w[1]*exp(j*dx)` for some fixed `dx`. 
This grid is set up automatically when creating a MizerParams object.

In the code the size spectrum is stored as an array such that `n[i, a]`
holds the density $N_i(w_a)$ at weights $w_a=$`w[a]`, or, if time
dependence is included, an array such that `n[i, a, u]`
holds $N_i(w_a,t_u)$.

Note that, contrary to what one might have expected,
`n[i, a]` is not the *number* of individuals in a size bin
but the *density* at a grid point.
The number of individuals in the size bin between `w[a]` and
`w[a+1]=w[a]+dw[a]` is only approximately given as `n[i, a]*dw[a]`,
where `dw[a]= w[a+1]-w[a]`.

</details>

The time evolution of the size spectrum is described by the 
McKendrik-von Foerster equation, which is a continuity equation with loss:

\begin{equation}
  \label{eq:MvF}
  \frac{\partial N_i(w)}{\partial t} + \frac{\partial g_i(w) N_i(w)}{\partial w} 
  = -\mu_i(w) N_i(w),
\end{equation}

where individual growth $g_i(w)$ and mortality $\mu_i(w)$ will be described
below.

<details>
This McKendrik-von Foerster equation is approximated in mizer by a
finite-difference method (to be described in section ...). This allows the
`project()` function in mizer to project the size
spectrum forwards in time: Given the spectrum at one time the `project()`
function calculates it at a set of later times.

</details>

The McKendrik-von Foerster equation is supplemented by a boundary condition at
the egg weight $w_0$ where the flux of individuals (numbers per time) 
$g_i(w_0)N_i(w_0)$ is determined by the rate $R_i$ of production of offspring by
mature individuals in the population:
\begin{equation}
  g_i(w_0)N_i(w_0) = R_i.
\end{equation}

<details>

In the code this boundary condition is actually implemented as an equation for
the rate of change of the number of individuals in the smallest size bin:
\begin{equation}
\frac{d N_i(w_0)}{dt}=...
\end{equation}

</details>

## Resource size spectrum

Besides the fish spectrum there is also a resource spectrum $N_R(w)$,
representing for example the phytoplankton. This spectrum starts at a smaller
size than the fish spectrum, in order to provide food also for the smallest
individuals (larvae) of the fish spectrum.

By default the time evolution of the resource spectrum is described by a 
semi-chemostat equation:
\begin{equation}
  \label{eq:nb}
  \frac{\partial N_R(w,t)}{\partial t} 
  = r_p(w) \Big[ c_p (w) - N_R(w,t) \Big] - \mu_p(w) N_R(w,t).
\end{equation}
Here $r_p(w)$ is the plankton regeneration rate and $c_p(w)$ is the carrying
capacity in the absence of predation. These parameters are changed with
`setPlankton()`. By default mizer assumes allometric forms
\[r_p(w)= r_p\, w^{n-1}.\]
\[c_p(w)=\kappa\, w^{-\lambda}.\]
It is also possible to implement other plankton dynamics, as
described in the help page for `setPlankton()`. The death $\mu_p(w)$ is
described in the subsection [Plankton mortality].

<details>
Because the plankton spectrum spans a different range of sizes these sizes
are discretized into a different vector of weights
`w_full`. The last entries of `w_full` have to coincide with the entries
of `w`. The resource spectrum is then stored in a vector `n_pp` such that
`n_pp[c]` =$N_R($`w_full[c]`$)$.

The plankton regeneration rate is stored as a vector
`params@rr_pp[c]`$=r_p($`w_full[c]`$)$, and the carrying capacity
is stored as a vector
`params@cc_pp[c]`$=c_p($`w_full[c]`$)$.

</details>



## Predator-prey encounter rate {#sec:pref}

The rate at which a predator of species $i$ and weight $w$ encounters food 
(mass per time) is
determined by summing over all prey species and the resource spectrum and
integrating over all prey sizes $w_p$, weighted by the selectivity factors:
\begin{equation}
  \label{eq:1}
  E_{i}(w) = \gamma_i(w) \int \left(\sum_{j} \theta_{ij} N_j(w_p) +
   \theta_{ip} N_R(w_p) + \right) 
  \phi_i(w,w_p) w_p \, dw_p.
\end{equation}
This is calculated by `getEncounter()`.
The overall prefactor $\gamma_i(w)$ sets the predation power of the predator. It
could be interpreted as a search volume. It is set by `setSearchVolume()`. By 
default it is assumed to scale allometrically as
$\gamma_i(w) = \gamma_i\, w^q.$

The $\theta$ matrix sets the interaction strength between predators and the
various prey species and plankton. It is changed with `setInteraction()`.

The size selectivity is encoded in the predation kernel $\phi_i(w,w_p)$. This is
changed with `setPredKernel()`.

<details>
An important simplification occurs when the predation kernel $\phi_i(w,w_p)$
depends on the size of the prey **only** through the predator/prey size ratio
$w_p/w$,
\[\phi_i(w, w_p)=\tilde{\phi}_i(w/w_p).\]
This is assumed by default but can be overruled.
The default for the predation kernel is the truncated log-normal function
\[
  \label{eq:4}
  \tilde{\phi}_i(x) = \begin{cases}
  \exp \left[ \dfrac{-(\ln(x / \beta_i))^2}{2\sigma_i^2} \right]
  &\text{ if }x\in\left[0,\beta_i\exp(3\sigma_i)\right]\\
  0&\text{ otherwise,}
  \end{cases}
\]
where $\beta_i$ is the preferred predator-prey mass ratio and $\sigma_i$ sets
the width of the predation kernel.

The integral in the expression for the encounter rate is approximated by a 
Riemann sum over all weight brackets:
\[
{\tt encounter}[i,a] = {\tt search\_vol}[i,a]\sum_{k}
\left( n_{pp}[k] + \sum_{j} \theta[i,j] n[j,k] \right) 
  \phi_i\left(w[a],w[k]\right) w[k]\, dw[k].
\]
In the case of a predation kernel that depends on $w/w_p$ only, this becomes
a convolution sum and can be evaluated efficiently via fast Fourier transform.

</details>


## Consumption

The encountered food is consumed subjected to a standard Holling functional 
response type II to represent satiation. This determines the 
*feeding level* $f_i(w)$, which is a dimensionless number between 0 
(no food) and 1 (fully satiated) so that $1-f_i(w)$ is the proportion of the
encountered food that is consumed. The feeding level is given by

\begin{equation}
  \label{eq:f}
  f_i(w) = \frac{E_{i}(w)}{E_{e.i}(w) + h_i(w)},
\end{equation}

where $h_i(w)$ is the maximum consumption rate. This is changed with
`setIntakeMax()`. By default mizer assumes an allometric form
$h_i(w) = h_i\, w^n.$
The feeding level is calculated with the function `getFeedingLevel()`.

The rate at which food is consumed is then 
\begin{equation}
(1-f_i(w))E_{i}(w)=f_i(w)\, h_i(w).
\end{equation}

## Metabolic losses

Some of the consumed food is used to fuel the needs for metabolism and 
activity and movement, at a rate ${\tt metab}_i(w)$. By default 
this is made up out of standard metabolism, scaling with exponent $p$, and
loss due to activity and movement, scaling with exponent $1$:
\[{\tt metab}_i(w) = k_{s.i}\,w^p + k_i\,w.\]
See the help page for `setMetab()`. 

The remaining rate, if any, is assimilated with an efficiency $\alpha_i$ and 
is then available for growth and reproduction. So the rate at which energy
becomes available for growth and reproduction is
\begin{equation}
  \label{eq:Er}
  E_{r.i}(w) = \max(0, \alpha_i f_i(w)\, h_i(w) - {\tt metab}_i(w))
\end{equation}
This is calculated with the `getEReproAndGrowth()` function.


## Investment into reproduction  {#sec:repro}
A proportion $\psi_i(w)$ of the energy available for growth and reproduction is
used for reproduction. This proportion should change from zero below the weight
$w_{m.i}$ of maturation to one at the asymptotic weight $w_{\infty.i}$, where
all available energy is used for reproduction. This function is changed with
`setReproduction()`. Mizer provides a default form for the function which you
can however overrule.

The total production rate of egg production $R_{p.i}$ (numbers per year) is
found by integrating the contribution from all individuals of species $i$:
\begin{equation}
  \label{eq:Rp}
  R_{p.i} = \frac{\epsilon}{2 w_0} \int N_i(w)  E_{r.i}(w) \psi_i(w) \, dw,
\end{equation}
where the individual contribution is obtained by multiplying the rate at which the
individual allocates energy to reproduction by an efficiency factor $\epsilon$ 
and then dividing by the egg weight $w_0$ to convert the energy into number of eggs.
The result is multiplied by a factor $1/2$ to take into account that only 
females reproduce.


## Recruitment
In mizer, density dependence is modelled as a compensation on the egg production. This can be considered as the stock-recruitment relationship (SRR).
The default functional form is such that the recruitment flux $R_i$ (numbers per time) approaches a maximum recruitment as the egg production increases, modelled mathematically analogous to the Holling type II function response as a *Beverton-Holt* type of SRR:

\begin{equation}
  \label{eq:R}
  R_i = R_{\max.i} \frac{R_{p.i}}{R_{p.i} + R_{\max.i}},
\end{equation}
where $R_{\max.i}$ is the maximum recruitment flux of each trait class (Figure~\ref{fig:recruitment}). 

The *Beverton-Holt* type of SRR is not the only density dependence model that mizer can use. Users are able to write their own model so it is possible to set a range of SRRs, e.g. fixed recruitment (as used in the community-type model) or *hockey-stick.*


## Growth
What is left over after metabolism and reproduction is taken into account
is invested in somatic growth. Thus the growth rate is
\begin{equation}
  \label{eq:growth}
  g_i(w) = E_{r.i}(w)\left(1-\psi_i(w)\right).
\end{equation}
It is calculated by the `getEGrowth()` function.

When food supply does not cover the requirements of metabolism and activity, 
growth and reproduction stops, i.e. there is no negative growth.
The individual should then be subjected to a starvation mortality, but starvation
mortality is not implemented in mizer at the moment.


## Mortality
The mortality rate of an individual $\mu_i(w)$ has three sources: 
predation mortality $\mu_{p.i}(w)$, background mortality $\mu_{b.i}(w)$ and 
fishing mortality $\mu_{f.i}(w)$. 

Predation mortality is calculated such that all that is eaten translates into 
corresponding predation mortalities on the ingested prey individuals. 
Recalling that $1-f_j(w)$ is the proportion of the food encountered by a 
predator of species $j$ and weight $w$ that is actually consumed, the
rate at which all predators of species $j$ consume prey of size $w_p$ is
\begin{equation}
  \label{eq:pred_rated}
  {\tt pred\_rate}_j(w_p) = \int \phi_j(w,w_p) (1-f_j(w))
  \gamma_j(w) N_j(w) \, dw.
\end{equation}
This predation rate is calculated by the function `getPredRate()`.

<details>
The integral is approximated by a Riemann sum over all fish weight brackets.
\[
{\tt pred\_rate}[j,c] = \sum_{a}
  {\tt pred_kernel}[j,a,c]\,(1-{\tt feeding_level}[j,a])\,
  \gamma[j,a]\,n[j,a]\,dw[a].
\]

</details>

The mortality rate due to predation is then obtained as
\begin{equation}
  \label{eq:mup}
  \mu_{p.i}(w_p) = \sum_j {\tt pred\_rate}_j(w_p)\, \theta_{ji}.
\end{equation}
This predation mortality rate is calculated by the function `getPredMort()`.

Background mortality $\mu_{b .i}(w)$ is independent of the abundances and is
changed with `setBMort()`. By default mizer assumes an allometric form
\[\mu_{b.i}(w) = \mu_b w_{\infty.i}^{1-n},\]
where $w_{\infty.i}$ is the asymptotic size of species $i$.

Fishing mortality $\mu_{f.i}(w)$ will be discussed in section [Fishing].

The total mortality rate
\[\mu_i(w)=\mu_{p.i}(w)+\mu_{b.i}(w)+\mu_{f.i}(w)\]
is calculated with the function `getMort()`.


## Plankton Mortality

The predation mortality rate on plankton is given by a similar expression
as the predation mortality on fish:
\begin{equation}
  \label{eq:mupp}
  \mu_{p}(w_p) = \sum_j {\tt pred\_rate}_j(w_p)\, \theta_{jp}.
\end{equation}
This is the only mortality on plankton currently implemented in mizer.
It is calculated with the function `getPlanktonMort()`.


## Fishing
This section still to be written.


