---
title: "The Multi Species Model"
output:
  html_document:
    toc: yes
    fig_width: 5
    fig_height: 5
---

```{r include=FALSE}
library(mizer)
```


The previous sections have used wrapper functions to set up MizerParams
objects that are appropriate for community and trait-based simulations. We now
turn our attention to multispecies, or species-specific, models. These are
potentially more complicated than the community and trait-based models and use
the full power of the `mizer` package.

In multispecies type models multiple species are resolved. However, unlike in
the trait-based model which also resolves multiple species, the species
parameters will be those of real-world species. There are several advantages to
this approach. As well as investigating the community as a whole (as was done
for the community and trait-based models), we are able to investigate the
dynamics of individual species. This means that species specific management
rules can be tested and species specific metrics, such as yield, can be compared
to reference levels.

A multispecies model can take more effort to set up. For example, each species
will have different life-history parameters; there may be multiple gear types
with different selectivities targeting different groups of species; the fishing
effort of each gear may change with time instead of just being constant (which
has been the case in the simulations we have looked at so far); the interactions
between the species needs to be considered.

In later sections we build up a multispecies model for the North Sea. To
effectively use `mizer` for a multispecies model we are going to have to take a
closer look at the `MizerParams` class and the `project()` function. This will
all be done in the context of examples so hopefully everything will be clear.

We also take a closer look at some of the summary plots and analyses that can be
performed, for example, calculating a range of size-based indicators.

# Setting up a multispecies model {#sec:MizerParams}

## Overview
The MizerParams class is used for storing model parameters. We have already
met the MizerParams class when we looked at community and trait-based models.
However, to set up a multispecies model we will need to specify many more
parameters.This is probably the most complicated part of using the mizer
package, so we will take it slowly.

A `MizerParams` object stores the:

* life-history parameters of the species in the community, such as asymptotic
  size $w_{\infty}$;
* size-based biological parameters for the species, such as the search volume;
* density-dependent reproduction functions and parameters of each species;
* an interaction matrix to describe the spatial overlap of pairs of species;
* parameters relating to the growth and dynamics of the resource spectrum;
* fishing gear parameters: selectivity and catchability.

Note that the `MizerParams` class does not store any parameters that can vary
through time, such as fishing effort or population abundance. These are stored
in the `MizerSim` class which we will come to later in 
[the section on running a simulation.](running_a_simulation.html#sec:projection)

Although the `MizerParams` class contains a lot of information, it is relatively
straightforward to set up and use. Objects of class `MizerParams` are created
using the constructor method `newMultispeciesParams()` (this constructor method
was called MizerParams() in previous version of mizer). This constructor method
can take many arguments. However, creation is simplified because many of the
arguments have default values.

In the rest of this section we look at the main arguments to the
`newMultispeciesParams()` function. To help understand how the constructor is
used and how the `MizerParams` class relates to the equations given in 
[the model description section,](model_description.html) there is an example section
where we create example parameter objects using data that comes with the `mizer`
package.

## The species parameters {#sec:species_parameters_dataframe}

Although many of the arguments used when creating a `MizerParams` object are
optional, there is one argument that must be supplied by the user: the *species
specific parameters*.  These are stored in a single `data.frame` object. The
`data.frame` is arranged species by parameter, so each column is a parameter and
each row has the parameters for one of the species in the model. Although it is
possible to create the data.frame by hand in R, it is probably easier to create
the data externally as a .csv file (perhaps using a suitable open source
spreadsheet such as LibreOffice) and then read the data into R.

For each species in the model community there are certain parameters that are
essential and that do not have default values. The user must provide values for
these parameters. There are also some essential parameters that have default
values, such as the selectivity function parameters, and some that are
calculated internally using default relationships if not explicitly provided.
These defaults are used if the parameters are not found in the data.frame.

The essential columns of the species parameters data.frame that have no default 
values are: `species`, the names of the species in the community and `w_inf`, 
the asymptotic mass of the species.


## The gear parameters {#sec:fishing_gear}

In `mizer`, fishing mortality is imposed on species by fishing gears. The total
fishing mortality is obtained by summing over the mortality from all gears,
\begin{equation}
%     {#eq:muf}
    \mu_{f.i}(w) = \sum_g F_{g,i}(w),
\end{equation}
where the fishing mortality $F_{g,i}(w)$ imposed by gear $g$ on species $i$ at
size $w$ is calculated as:
\begin{equation}
%     {#eq:sel}
    F_{g,i}(w) = S_{g,i}(w) Q_{g,i} E_{g}
\end{equation}
where $S$ is the selectivity by species, gear and size, $Q$ is the catchability
by species and gear and $E$ is the fishing effort by gear. The selectivity at
size has a range between 0 (not selected at that size) to 1 (fully selected at
that size). Catchability is used as an additional scalar to make the link
between gear selectivity, fishing effort and fishing mortality. For example, it
can be set so that an effort of 1 gives a desired fishing mortality. In this way
effort can then be specified relative to a 'base effort', e.g. the effort in a
particular year.

Selectivity and catchability are stored as arrays in the MizerParams object.
However the user does not have to create these arrays by hand if they provide a
data frame with the necessary information. In particular the selectivity can be
calculate by specifying functions for the selectivity curves. Mizer provides a
range of such selectivity functions and the user just needs to specify their
parameters for each gear and each species in the `gear_params` data frame. All
the details can be found on the help page for `setFishing()`.

Fishing effort is not stored in the MizerParams object. Instead, effort is set
when the simulation is run and can vary through time (see 
[the section on running a simulation](running_a_simulation.html)).


## Example of making `MizerParams` objects {#sec:params_example}

As mentioned in the preceding sections, an object of `MizerParams` is created by
using the `newMultispeciesParams()` constructor method.

The first step is to prepare the species specific parameter data.frame. As
mentioned above, one way of doing this is to use a spreadsheet and save it as a
.csv file. We will use this approach here. An example .csv file has been
included in the package. This contains the species parameters for a multispecies
North Sea model. This file is placed in the `doc` folder of the package
installation. The location of the file can be found by running
```{r label = get_location_for_ns_params,eval=FALSE}
system.file("doc/NS_species_params.csv",package="mizer")
```

This file can be opened with most spreadsheets or a text editor for you to
inspect. This can be loaded into R using the following code (after you have told
R to look in the right directory):
```{r }
species_params <- read.csv("NS_species_params.csv")
```
This reads the .csv file into R in the form of a data.frame.
You can check this with the `class`:
```{r }
class(species_params)
```

Let's have a look at the data frame:

```{r label=show_simple_species_params}
species_params
```

You can see that there are $`r nrow(species_params)`$ species and $`r
ncol(species_params)`$ columns of parameters: `species`,
`w_inf`,`w_mat`,`beta`,`sigma`,`R_max` and `k_vb`.

Of these parameters, `species` and `w_inf` are essential and have no default
values (as described in 
[the section on species parameters](multispecies_model.html#sec:species_parameters_dataframe)). 
`w_inf`
is the asymptotic size of the species, `w_mat` is its maturity size, and `beta`
and `sigma` are parameters of the predation kernel (by default mizer uses a
log-normal predation kernel). `R_max` is a parameter introducing additional
density dependence into reproduction parameter using a Beverton-Holt type
function (see `setReproduction()` for details). The final column, `k_vb`, will
be used to calculate values for `h` and then `gamma`. This column is only
essential here because the `h` and `gamma` are not included in the data.frame.
It would also have been possible to include `h` and `gamma` columns in the
data.frame and not include the `k_vb` column.

The values of the non-essential species specific parameters, like for example
`alpha`, `k`, `ks`, `z0`, `w_min` and `erepro`, are not included in the
data.frame. This means that the default values will be automatically used when
we create the `MizerParams` object.

For this example we will not set up gear parameters. There are no columns
describing the fishing selectivity. There is no `sel_func` column to determine
the selectivity function. This means that the default selectivity function,
`knife_edge`, will be used. As mentioned in 
[the section on fishing gears](multispecies_model.html#sec:fishing_gear), 
this function also needs
another argument, `knife_edge_size`. This is not present in the data.frame and
so it will be set to the default value of `w_mat`. Also, there is no
`catchability` column so a default value for `catchability` of 1 will be used
for all gears and species.

To create the `MizerParams` object we pass the species parameter data.frame into
the `newMultispeciesParams()` constructor method:

```{r label=first_MP, results='hide'}
params <- newMultispeciesParams(species_params)
```

We have just created a `MizerParams` object:

```{r label=class_MP}
class(params)
```

The MizerParams object also stores a copy of the species parameter data frame
that we provided. We can look at it with `species_params()`:

```{r label=looking_at_params_slot}
species_params(params)
```

We can see that this returns the original species data.frame (with `w_inf` and
so on), plus any default values that may not have been included in the original
data.frame. For example, we can see that there are now columns for `alpha` and
`h` and `gamma` etc.

Also note how the default fishing gears have been set up. Even though we did
not provide a gear parameter data frame, the MizerParams object has one that we
can access with 
```{r}
gear_params(params)
```

All species are caught by a gear called "knife_edge_gear".  The selectivity
function for each fishing gear has been set in the `sel_func` column to the
default function, `knife_edge()`. A `catchability` column has been added with a
default value of 1 for each of the species that the gear catches. An example of
setting the catchability by hand can be seen in 
[the section on the North Sea.](a_multispecies_model_of_the_north_sea.html)

There is a `summary()` method for `MizerParams` objects which prints a useful
summary of the model parameters:

```{r label=summary_params}
summary(params)
```

As well as giving a summary of the species in the model and what gear is fishing
what species, it gives a summary of the size structure of the community. For
example there are $`r length(w(params))`$ size classes in the community, ranging
from $`r signif(min(w(params)), 3)`$ to $`r signif(max(w(params)), 3)`$ . These
values are controlled by the arguments `no_w`, `min_w` and `max_w` respectively.
For example, if we wanted 200 size classes in the model we would use:

```{r label=MP_200, eval=FALSE}
params200 <- newMultispeciesParams(species_params, no_w = 200)
summary(params200)
```

## Setting the interaction matrix

So far we have created a `MizerParams` object by passing in only the species
parameter data.frame argument. We did not specify an interaction matrix. The
interaction matrix \eqn{\theta_{ij}} describes the interaction of each pair of
species in the model. This can be viewed as a proxy for spatial interaction e.g.
to model predator-prey interaction that is not size based. The values in the
interaction matrix are used to scale the encountered food in [getEncounter()]
and the predation mortality rate in [getPredMort()] (see 
[the section on predator-prey encounter rate](model_description.html#sec:pref) 
and on [predation mortality](model_description.html#mortality)).

The entries of the interaction matrix are dimensionless numbers taking values
are between 0 (species do not overlap and therefore do not interact with each
other) to 1 (species overlap perfectly). By default mizer sets all values to 1,
implying that all species fully interact with each other, i.e. the species are
spread homogeneously across the model area. 
```{r}
getInteraction(params)
```

For the North Sea this is not the
case and so the model would be improved by also including an interaction matrix
which describes the spatial overlap between species.

An example interaction matrix for the North Sea has been included in `mizer` as
a .csv file.  The location of the file can be found by running:

```{r eval=FALSE}
system.file("doc/inter.csv",package="mizer")
```

Take a look at it in a spreadsheet if you want.  As mentioned above, to read
this file into R we can make use of the `read.csv()` function. However, this
time we want the first column of the .csv file to be the row names. We therefore
use an additional argument to the `read.csv()` function: `row.names`.

We can set the interaction matrix in our existing MizerParams object `params`
with the `setInteraction()` function:
```{r}
params <- setInteraction(params, interaction = inter)
```

Alternatively, instead of changing the interaction matrix in the existing
MizerParams object, we could have created a new object from scratch with our
interaction matrix by passing it to `newMultispeciesParams()`:

```{r results='hide'}
params_new <- newMultispeciesParams(species_params, interaction = inter)
```
Note that the first argument must be the species parameters data.frame.
The remaining arguments can be in any order but should be named.
We are using the default values for all other parameters.

Both methods lead to identical MizerParams objects
```{r}
identical(params, params_new)
```

We now have all we need to start running projections. Before we get to that
though, we'll take a quick look at how different fishing gears can be set up.




## Setting different gears {#sec:setting_different_gears}

In the above example, each species is caught by the same gear (named
"knife_edge_gear"). This is the default when no gear information is provided.
```{r}
gear_params(params)
```


Here, we look at an example where we set up four different gears: Industrial,
Pelagic, Beam and Otter trawl, that catch different combinations of species.
We can achieve that by only changing the `gear` column in the `gear_params`
data frame.
```{r }
gear_params(params)$gear <- c("Industrial","Industrial","Industrial",
                              "Pelagic","Beam","Otter",
                              "Beam","Otter","Beam",
                              "Otter","Otter","Otter")
```
You can see the result by calling `summary()` on the `params` object.
```{r}
summary(params)
```

In this example the same gear now catches multiple stocks. For example, the
*Industrial* gear catches Sprat, Sandeel and Norway Pout. Why would we want to
set up the gears like this? In the next section on 
[running a multispecies model](running_a_simulation.html) we will see that to
project the model through time you can specify the fishing effort for each gear
through time. By setting the gears up in this way you can run different
management scenarios of changing the efforts of the fishing gears rather than on
individual species. It also means that after a simulation has been run you can
examine the catches by gear.


## Setting to steady state

Once the `MizerParams` object has been properly set up, it may be the case that
one wishes put the system in steady state. Sometimes this can be done simply by
running the model using `project()` until it reaches steady state. However, this
method is not guaranteed to work, and there is a function called `steady()` that is
more reliable. The function `steady()` must be supplied with a MizerParams
object. It takes that MizerParams object, looks at the initial system state,
computes the levels of reproduction of the different species, hold them fixed,
and evolves the system until a steady state is reached (or more precisely, until
the amount that the population abundances change during a time-step is below
some small tolerance level). After this, the reproductive efficiency of each
species is altered so that when the reproduction dynamics are turned back on
(i.e., when we stop holding recruitment levels fixed), the values of the
reproduction levels which we held the system fixed at will be realized. The
steady function is not sure to converge, and the way it re-tunes the
reproductive efficiency values may not be realistic, but the idea is to alter
the other parameters in the system until `steady()` does arrive at a steady
state with sensible reproductive efficiency values.

Now that we know how to create a multispecies model we shall discuss how to 
[run a multispecies model.](running_a_simulation.html)
