---
title: "The Multi Species Model"
output:
  html_document:
    number_sections: yes
    toc: yes
    fig_width: 5
    fig_height: 5
---

```{r include=FALSE}
library(mizer)
```


The previous sections have used wrapper functions to set up `MizerParams`
objects that are appropriate for community and trait-based simulations. We now
turn our attention to multispecies, or species-specific, models. These are
potentially more complicated than the community and trait-based models and use
the full power of the `mizer` package.

In multispecies type models multiple species are resolved. However, unlike in
the trait-based model which also resolves multiple species, explicit species are
represented. There are several advantages to this approach. As well as
investigating the community as a whole (as was done for the community and
trait-based models), we are able to investigate the dynamics of individual
species. This means that species specific management rules can be tested and
species specific metrics, such as yield, can be compared to reference levels.

A multispecies model can take more effort to set up. For example, each species
will have different life-history parameters; there may be multiple gear types
with different selectivities targeting different groups of species; the fishing
effort of each gear may change with time instead of just being constant (which
has been the case in the simulations we have looked at so far); the interactions
between the species needs to be considered.

In later sections we build up a multispecies model for the North Sea. To
effectively use `mizer` for a multispecies model we are going to have to take a
closer look at the `MizerParams` class and the `project()` function. This will
all be done in the context of examples so hopefully everything will be clear.

We also take a closer look at some of the summary plots and analyses that can be
performed, for example, calculating a range of size-based indicators.

# Setting up a multispecies model {#sec:MizerParams}

## Overview
The `MizerParams` class is used for storing model parameters. We have already
met the `MizerParams` class when we looked at community and trait-based models.
However, direct handling of the class was largely hidden by the use of the
wrapper functions.

A `MizerParams` object stores the:

* life-history parameters of the species in the community, such as asymptotic
  size $w_{\infty}$;
* size-based biological parameters for the species, such as the search volume;
* density-dependent reproduction functions and parameters of each species;
* an interaction matrix to describe the spatial overlap of pairs of species;
* parameters relating to the growth and dynamics of the plankton spectrum;
* fishing gear parameters: selectivity and catchability.


Note that the `MizerParams` class does not store any parameters that can vary
through time, such as fishing effort or population abundance. These are stored
in the `MizerSim` class which we will come to later in [the section on running a
simulation.](running_a_simulation.html#sec:projection)

Although the `MizerParams` class seems complicated, it is relatively
straightforward to set up and use. Objects of class `MizerParams` are created
using the constructor method `newMultispeciesParams()` (this constructor method
was called MizerParams() in previous version of mizer). This constructor method
can take many arguments. However, creation is simplified because many of the
arguments have default values.

In the rest of this section we look at the main arguments to the
`newMultispeciesParams()` function. To help understand how the constructor is
used and how the `MizerParams` class relates to the equations given in [the
model description section,](model_description.html) there is an example section
where we create example parameter objects using data that comes with the `mizer`
package.

## The species parameters {#sec:species_parameters_dataframe}

Although many of the arguments used when creating a `MizerParams` object are
optional, there is one argument that must be supplied by the user: the *species
specific parameters*.  These are stored in a single `data.frame` object. The
`data.frame` is arranged species by parameter, so each column is a parameter and
each row has the parameters for one of the species in the model. Although it is
possible to create the data.frame by hand in R, it is probably easier to create
the data externally as a .csv file (perhaps using a suitable open source
spreadsheet such as LibreOffice) and then read the data into R.

For each species in the model community there are certain parameters that are essential and that do not have default values. The user must provide values for these parameters.
There are also some essential parameters that have default values, such as the selectivity function parameters, and some that are calculated internally using default relationships if not explicitly provided. These defaults are used if the parameters are not found in the data.frame. The species parameter information is stored as a data.frame which hold the information given in [the MizerParams help page.](../reference/newMultispeciesParams.html#note)


<!-- 

A description of the columns of the species parameter data.frame and any default values can be seen Table~\@ref(tab:species_params).

-->


The essential columns of the species parameters data.frame that have no default 
values are: `species`, the names of the species in the community and `w_inf`, 
the asymptotic mass of the species.



<!-- need to check these are still the defaults -->

Essential columns that have default values are: `k`, the activity coefficient (default = 0); `alpha`, the assimilation efficiency (default = 0.6); `erepro`, the reproductive efficiency (default value = 1); `w_min`, the size of recruits (default value is the smallest size of the community size spectrum); `sel_func`, the name of the fishing selectivity function (default value = "knife_edge"); `gear`, the name of the fishing gear that catches the species (default value is the name of the species in the `species` column); `catchability`, the catchability of the fishing gear on that species (default value = 1). Additionally, columns that contain the selectivity function parameters are needed. As mentioned, the default selectivity function is a *knife_edge* function. This has an argument `knife_edge_size` that determines the knife-edge position and has a default value of `w_min`.
If any of these columns are not included in the species parameter data.frame, the default values are used.

As mentioned above, there are some columns that are essential but if they are not provided, values for them are estimated using the values in the other columns. These columns are:
`h`, the maximum food intake; `gamma`, the volumetric search rate; `ks`, the coefficient for standard metabolism; `z0`, the mortality from other sources, $\mu_{b,i}.$ These parameters can be included as columns in the species parameters `data.frame` if they are available. If they are not provided then the `newMultispeciesParams()` construction method will try to calculate them.

The `h` column is calculated as:

\begin{equation}
    h = \frac{3 k_{vb}}{\alpha f_0} w_\infty^{1/3}
\end{equation}

where $k_{vb}$ is the von Bertalanffy $K$ parameter and $f_0$ is the feeding level of small individuals feeding mainly on the plankton.
This mean that if an `h` column is not included in the species parameter data.frame, a column for `k_vb` is necessary. If it is not included then the `newMultispeciesParams()` function will fail with an error message. The calculation also requires a value of the feeding level of small individuals. This can be passed as an additional argument, `f0`, to the `newMultispeciesParams()` constructor and it has a default value of 0.6.



The `gamma` column is calculated using [the equation for volumetric search rate.](model_description.html#eq:gamma)
This calculation requires that the `h` column is available, either included in the species parameter data.frame, or calculated internally using the `k_vb` column as described above. This means that if you include a `k_vb` column in the data.frame, `h` and `gamma` will be calculated from it.


The `z0` column (mortality from other sources) is calculated using the equation ${\tt z0}_i(w) = {\tt z0pre }\  w_{\infty.i}^{\tt{z0exp}}.$ The values `z0pre` (equivalent to $\mu_0$) and `z0exp` (the power that $w_\infty$ is raised to) can be passed as arguments to the `newMultispeciesParams()` constructor and have default values of 0.6 and `n-1` (where `n` has a default value of 2/3) respectively.

The `ks` column is calculated as 20\% of `h`, i.e. the standard metabolism coefficient is 20\% of the maximum consumption.
Standard metabolism is used in calculation of the growth of individuals as $k_{s.i}  w_i^p$ (see [the section on allometric rates](model_description.html#sec:allo)). The species parameter information is stored as a data.frame which hold the information described in [the MizerParams help page.](../reference/newMultispeciesParams.html#note)
 
<!-- The old generator for table 2 used to be here, but I am commenting it out --> 
 

<!--
 
%\begin{sidewaystable}[ht]
\begin{table}[ht]
    \begin{center}
	%\begin{tabular}{r|p{7cm}|p{7cm}}
	%\begin{tabularx}{r|X|X}
	    %\begin{tabularx}{\linewidth}{r|p{7cm}|X}
	    \begin{tabularx}{\linewidth}{r|X|X}
	    \hline
	    Column name & Description & Default value \\
	    \hline
	    \multicolumn{3}{c}{Life history parameters} \\
	    \hline
	    species & Name of the species & Compulsory (no default)\\
	    w_inf & The asymptotic mass of the species & Compulsory (no default)\\
	    w_mat & Maturation mass. Used to calculate values for $\psi$. WHAT IS WMAT IN TERMS OF BIOLOGY? Is it the mass at first maturity? & Compulsory (no default)\\
	    beta & Preferred predator prey mass ratio & Compulsory (no default)\\
	    sigma & Width of prey size preference & Compulsory (no default)\\
        h & Maximum food intake rate. If this is not provided, it is calculated using the `k_vb` column. Therefore, either `h` or `k_vb` must be provided. & Optional (no default) \\
        k_vb & The von Bertalanffy $K$ parameter. Only used to calculate `h` if that column is not provided & Optional (no default)\\
	    gamma & Volumetric search rate. If this is not provided, it is calculated using the `h` column and other parameters. & Optional (no default)\\
        ks & Standard metabolism coefficient & `h` * 0.2\\
        z0 & Background mortality (constant for all sizes). If this is not provided then `z0` is calculated as $z0pre * w_{\infty} ^{z0exp}$. `z0pre` and `z0exp` have default values of 0.6 and -1/3 respectively.  & Optional (no default)\\
	    k & Activity coefficient & 0.0\\
	    alpha & Assimilation efficiency & 0.6 \\
	    erepro & Reproductive efficiency & 1 \\
	    w_min & The size class that recruits are placed in. & smallest size class of the species size spectrum \\
	    \hline
	    \multicolumn{3}{c}{Fishing gear parameters (see Section~\@ref(sec:fishing_gear) for more details).} \\
	    \hline
	    sel_func & The name of the selectivity function to be used.  & "knife_edge".\\
	    gear & The name of the fishing gear that selects the species. At the moment a species can only be selected by one gear. & Name of the species \\
	    catchability & The catchability of the fishing gear. & 1\\
	    %a & Length-weight conversion coefficient & Only necessary if length-weight conversion is used (e.g. if specifying length ranges when calculating indicators, or in the fishing selectivity function) \\
	    %b & Length-weight conversion power & See notes for column $a$\\
	    other columns & Other parameters used by the selectivity function. For example, if the default "knife_edge" function is used then the parameters "knife_edge_size" must also be specified as columns (see Section~\@ref(sec:fishing_gear)). & \\ 
	    \hline
	    \multicolumn{3}{c}{Density-dependent recruitment parameters (see Section~\@ref(sec:srr) for more details).} \\
	    \hline
	    other columns & Any arguments that appear in the stock-recruitment function must also have a column of values (see Section~\@ref(sec:srr)) & \\
	%\end{tabular}
	\end{tabularx}
	\caption{Columns of the species parameters data.frame}
	 {#tab:species_params}
    \end{center}
%\end{sidewaystable}
\end{table}



You can see in Table~\@ref(tab:species_params) that most of the species specific
parameters relate to the life history of the species.
The others relate to the gear selectivity function and the stock-recruitment relationship.
These are explained further in Sections~\@ref(sec:fishing_gear) and \@ref(sec:srr) respectively. 

-->

## Fishing gears and selectivity {#sec:fishing_gear}


In this section we take a look at how fishing is implemented and how fishing gears are set up within
`mizer`.

In `mizer`, fishing mortality is imposed on species by fishing gears. The total fishing
mortality is obtained by summing over the mortality from all gears,
\begin{equation}
%     {#eq:muf}
    \mu_{f.i}(w) = \sum_g F_{g,i}(w),
\end{equation}
where the fishing mortality $F_{g,i}(w)$ imposed by gear $g$ on species $i$ at size $w$ is calculated as:
\begin{equation}
%     {#eq:sel}
    F_{g,i}(w) = S_{g,i}(w) Q_{g,i} E_{g}
\end{equation}
where $S$ is the selectivity by species, gear and size, $Q$ is the catchability by species
and gear and $E$ is the fishing effort by gear.
The selectivity at size has a range between 0 (not selected at that size) to 1 (fully selected at
that size).
Catchability is used as an additional scalar to make the link between gear selectivity, fishing
effort and fishing mortality. 
For example, it can be set so that an effort of 1 gives a desired fishing mortality.
In this way effort can then be specified relative to a 'base effort', e.g. the effort
in a particular year.
Fishing effort is not stored in the `MizerParams` object.
Instead, effort is set when the simulation is run and can vary through time (see [the section on running a simulation](running_a_simulation.html)).


At the moment a species can only be selected by one fishing gear, although each gear can select more than one species (this is a limitation with the current package that will be developed in future releases).

The selectivity at size of each gear is given by a selectivity function. Some selectivity
functions are included in the package. New functions can be defined by the user. 
Each gear has the same selectivity function for all the species it selects, but the parameter
values for each species may be different, e.g. the lengths of species that a gear selects may be different.

The name of the selectivity function is given by the `sel_func` column in the species parameters data.frame.
Each selectivity function has a range of arguments. Values for these arguments
must be included as columns in the species parameters data.frame. The names of the columns must exactly
match the names of the arguments. 
For example, the default selectivity function is `knife_edge` which has sudden change of selectivity from 0 to 1 at a certain size.
The arguments for this selectivity function can be seen in the help page
for this function. To see them enter:

```{r eval=FALSE}
?knife_edge
```

It can be seen that the `knife_edge()` function has arguments `w` and `knife_edge_size`
The first argument, `w`, is size (the function calculates selectivity at size).
All selectivity functions must have `w` as the first argument.
The values for the other arguments must be found in the species parameters data.frame.
So for the `knife_edge()` function there should be a `knife_edge_size` column (but note that because `knife_edge()` is the default selectivity function, the `knife_edge_size` argument actually has a default value = `w_mat`).
This can be seen in the example in [the section with examples of making a MizerParams object.](multispecies_model.html#sec:params_example)
If the columns of the selectivity function arguments are not in the species parameter data.frame, an error is thrown when the `MizerParams` object is created.



Users are able to write their own size based selectivity function. The first argument to the
function must be `w` and the function must return a vector of the selectivity
(between 0 and 1) at size.

The name of the fishing gear is given in the `gear` column of the species parameter
data.frame. If the `gear` column is not specified, the default gear name is simply
the name of the species. This implies that each species is fished by a different gear.
This approach can be used to explore the impacts of changing fishing mortality on individual species.

## The stock-recruitment relationship {#sec:srr}


In size spectrum modelling recruitment refers to the flux of individuals that enter the size-spectrum at the smallest size group of that species (given by the parameter `w_min` in the species parameter data.frame).
As can be seen in [the section on reproduction](model_description.html#sec:repro), calculating the recruitment flux involves calculating the density independent recruitment, $R_{p.i}$.
The $R_{p.i}$ is then modified by a stock-recruitment relationship (SRR) to impose some form of density-dependence. This then results in the density-dependent recruitment, $R_i$ (see [the section on the stock-recruitment relationship](model_description.html#sec:recruitment)).
Without this density dependence, the realised recruitment flux to the smallest size class is determined only by $R_{p.i}$.
The default SRR is a Beverton-Holt type function.

Similar to the fishing selectivity functions, any parameter used in the stock-recruitment function, other than $R_{p.i}$, must be in the species parameter data.frame and the column must have the same name as the function argument.
For example, the default stock-recruitment function has a second argument to it, `R_max`. Therefore the species parameter data.frame must have an `R_max` column.

Users are able to write their own stock-recruitment function. The first argument
to the function must be `rdi`, which is the density independent recruitment
$R_{p.i}$, the second argument must be `species_params`, the dataframe with the
species parameters. The name of this function should then be assigned to the
`srr` slot of the MizerParams object.



## The interaction matrix {#sec:interaction_matrix}

The interaction matrix describes the interaction of each pair of species
in the model. This can be viewed as a proxy for spatial interaction
e.g. to model predator-prey interaction that is not size based.
The values in the interaction matrix are used to scale the encountered food and predation mortality (see [the section on predator-prey preference](model_description.html#sec:pref)).
The matrix is square with every element being the interaction between a pair of species. The dimensions, nrows and ncolumns, therefore equal the number of species.
The values are between 0 (species do not overlap and therefore do not interact with each other) to 1 (species overlap perfectly).
If all the values in the interaction matrix are set to 1 then predator-prey interactions are determined entirely by size-preference.

The interaction matrix must be of type `array` or `matrix`. 
One way of creating your own is to enter the data using a spreadsheet (such as LibreOffice) and saving it as a .csv file. 
The data can be read into R using the command `read.csv()`. This reads in the data as a data.frame. We then need
to convert this to a matrix using the `as()` function. An example of how to do this is given in [the section with examples of making MizerParams objects.](multispecies_model.html#sec:params_example)





It should be noted that the order of species in the interaction matrix has to be the same as the order in the species parameters data.frame.
Although you can specify the dimnames of the interaction matrix, these names are overwritten by the species names from the species parameters data.frame
inside the `MizerParams` constructor.

If an interaction matrix is not specified to the `newMultispeciesParams()` constructor the default interaction matrix is used. This has all values set to 1. 



## The other arguments {#sec:other_mizerparams_arguments}

As well as the essential species parameters data.frame and the interaction
matrix, there are several other arguments to the `MizerParams` constructor.
These have default values. 
The arguments can be seen in the `newMultispeciesParams()` help page.

Some of these parameters may be used to calculate the species specific parameters if they are not provided. For example, if there is no `gamma` column in the species parameter data.frame, then it is calculated using `kappa` and `f0`.
This means that depending on which columns have been provided in the species parameters data.frame, some of the parameters from the the `newMultispeciesParams()` help page may not be used. For example, if the column `z0` (the mortality from other sources) has been included, then the arguments `z0pre` and `z0exp` are not used.

Determining a value for the `kappa` argument can be difficult and may need to be estimated through some kind calibration process.
The default value `kappa` is for the North Sea model.


## Examples of making `MizerParams` objects {#sec:params_example}

As mentioned in the preceding sections, an object of `MizerParams` is created by using the `newMultispeciesParams()`
constructor method. You can see the help page for the constructor:

```{r label=help_MP_constructor, eval=FALSE}
help(MizerParams)
```



This shows that the constructor takes the following arguments:

`object` The species parameter data.frame (see [the section on species parameters](multispecies_model.html#sec:species_parameters_dataframe)). This is compulsory with no default value.

`inter` The interaction matrix (see [the section on predator-prey preference](model_description.html#sec:pref)). The default is a matrix of 1s.
    
`...` Other model parameters.


In the rest of this section we demonstrate how to pull these elements together to make `MizerParams` objects. 

The first step is to prepare the species specific parameter data.frame.
As mentioned above, one way of doing this is to use a spreadsheet and save it as a .csv file. We will use this approach here.
An example .csv file has been included in the package. This contains the species parameters for a multispecies North Sea model.
This file is placed in the `doc` folder of the package installation.
The location of the file can be found by running:

```{r label = get_location_for_ns_params,eval=FALSE}
system.file("doc/NS_species_params.csv",package="mizer")
```

This file can be opened with most spreadsheets or a text editor for you to inspect.
This can be loaded into R using the following code (after you
have told R to look in the right directory):

```{r }
params_data <- read.csv("NS_species_params.csv")
```


This reads the .csv file into R in the form of a data.frame.
You can check this with the `class`:

```{r }
class(params_data)
```

The example data.frame can be inspected by entering the name of the object.

```{r label=show_simple_params_data}
params_data
```

You can see that there are $`r nrow(params_data)`$ species and $`r ncol(params_data)`$ columns of parameters: `species`, `w_inf`,`w_mat`,`beta`,`sigma`,`R_max` and `k_vb`.


Of these parameters, `species`, `w_inf`, `w_mat`, `beta` and `sigma` are essential and have no default values (as described in [the section on species parameters](multispecies_model.html#sec:species_parameters_dataframe)).
`R_max` is a SRR parameter. We are going to use the default Beverton-Holt type SRR which has `R_max` as an argument (see [the section on the stock-recruitment relationship](model_description.html#sec:recruitment)), making this column also essential.
The final column, `k_vb`, will be used to calculate values for `h` and then `gamma`. This column is only essential here because the `h` and `gamma` are not included in the data.frame. It would also have been possible to include `h` and `gamma` columns in the data.frame and not include the `k_vb` column.

The values of the non-essential species specific parameters `alpha`, `k`, `ks`, `z0`, `w_min` and `erepro` are not included in the data.frame. This means that the default values will be automatically used when we create the `MizerParams` object.



Note that there are no columns describing the fishing selectivity. There is no `sel_func` column to determine the selectivity function.
This means that the default selectivity function, `knife_edge`, will be used. As mentioned in [the section on fishing gears](multispecies_model.html#sec:fishing_gear),
this function also needs another argument, `knife_edge_size`. This is not present in the data.frame and so it will be set to the default value of `w_mat`. Also, there is no `catchability` column so a default value for `catchability` of 1 will be used for all gears and species.

This species parameter data.frame is the minimum we need to create a `MizerParams` object as it contains only essential columns. To create the `MizerParams` object we pass the data.frame into the `newMultispeciesParams()` constructor method:

```{r label=first_MP, results='hide'}
params <- newMultispeciesParams(params_data)
```

We have just created a `MizerParams` object:

```{r label=class_MP}
class(params)
```


As has been mentioned in [the section on the community
model](community_model.html) and [the section on the trait based
model](trait_model.html), a `MizerParams` object holds all the information
needed to define a mizer model.

The `params` object contains among other things the species parameters 
data.frame that was passed in to the constructor. We can inspect this with:

```{r label=looking_at_params_slot}
species_params(params)
```

We can see that this returns the original species data.frame (with `w_inf` and
so on), plus any default values that may not have been included in the original
data.frame. For example, we can see that there are now columns for `alpha` and
`h` and `gamma` etc.

Also note how the default fishing gears have been set up. Because we did not
specify gear names in the original species parameter data.frame, each species is
fished by a unique gear named after the species.  This can be seen in the new
`gear` column which holds the names of the fishing gears. Also, the selectivity
function for each fishing gear has been set in the `sel_func` column to the
default function, `knife_edge()`. A `catchability` column has been added with a
default value of 1 for each of the species that the gear catches. An example of
setting the catchability by hand can be seen in [the section on the North
Sea.](a_multispecies_model_of_the_north_sea.html)

There is a `summary()` method for `MizerParams` objects which prints a useful
summary of the model parameters:

```{r label=summary_params, results='hide'}
summary(params)
```

As well as giving a summary of the species in the model and what gear is fishing
what species, it gives a summary of the size structure of the community. For
example there are $`r length(w(params))`$ size classes in the community, ranging
from $`r signif(min(w(params))), 3)`$ to $`r signif(max(w(params)), 3)`$ . These
values are controlled by the arguments `no_w`, `min_w` and `max_w` respectively.
For example, if we wanted 200 size classes in the model we would use:

```{r label=MP_200, eval=FALSE}
params200 <- newMultispeciesParams(params_data, no_w = 200)
summary(params200)
```


So far we have created a `MizerParams` object by passing in only the species parameter data.frame argument.
This means the interaction matrix will be set to the default value (see [the section on predator/prey preference.](model_description.html#sec:pref)) This is a matrix of 1s, implying that all species fully interact with each other, i.e. the species are spread homogeneously across the model area.
For the North Sea this is not the case and so the model would be improved by also including an interaction matrix which describes the spatial overlap between species.

An example interaction matrix for the North Sea has been included in `mizer` as a .csv file.  The location of the file can be found by running:

```{r eval=FALSE}
system.file("doc/inter.csv",package="mizer")
```

Take a look at it in a spreadsheet if you want.  As mentioned above, to read this file into R we can make use of the `read.csv()` function. However, this time we want the first column of the .csv file to be the row names. We therefore use an additional argument to the `read.csv()` function: `row.names`.

```{r }
inter <- read.csv("inter.csv", row.names=1)
```

The `read.csv()` function reads the data into a data.frame.  We want the interaction matrix to be of class `matrix` so we need to make use of the `as()` function.

```{r label=inter_df_to_matrix}
inter <- as(inter, "matrix")
```

We now have an interaction matrix that can be passed to the `MizerParams` constructor along with the species parameter data.frame.
To make the `MizerClass` object you just call the constructor method and pass in the arguments.
We will use default values for the remainder of the arguments to the `newMultispeciesParams()` method.

<!--
(see Section~\@ref(sec:other_mizerparams_arguments) and Table~\@ref(tab:other_params)).
-->

This means that we only need to pass in two arguments to the constructor:


```{r results='hide'}
params <- newMultispeciesParams(params_data, interaction = inter)
```

Note that the first argument must be the species parameters data.frame.
The remaining arguments can be in any order but should be named.
If we didn't want to use default values for the other arguments we would pass them in to the constructor by name.

We now have all we need to start running projections. Before we get to that though, we'll take a quick look at how different fishing gears can be set up.




## Setting different gears {#sec:setting_different_gears}

In the above example, each species is caught by a different gear (named after the species it catches). This is the default when there is no `gear` column in the species parameter data.frame.

Here, we look at an example where we do specify the fishing gears.  We take the original `params_data` species parameter data.frame that was read in above and bind an additional column, `gear`, to it. This `gear` column contains the name of the gear that catches the species in that row.
Here we set up four different gears: Industrial, Pelagic, Beam and Otter trawl, that catch different combinations of species.

```{r }
params_data_gears <- params_data
params_data_gears$gear <- c("Industrial","Industrial","Industrial",
		      "Pelagic","Beam","Otter",
		      "Beam","Otter","Beam",
		      "Otter","Otter","Otter")
```

If you inspect the `params_data` object you will see a new column, `gear`, has been added to it.  
We then make a new `MizerParams` object as before:

```{r label=set_params_gears, results='hide'}
params_gears <- newMultispeciesParams(params_data_gears, interaction = inter)
```

You can see the result by calling `summary()` on the `params_gears` object.

In this example the same gear now catches multiple stocks. For example, the *Industrial* gear catches Sprat, Sandeel and Norway Pout.
Why would we want to set up the gears like this?
In the next section we will see that to project the model through time you can specify the fishing effort for each gear through time.
By setting the gears up in this way you can run different management scenarios of changing the efforts of the fishing gears rather than on
individual species.
It also means that after a simulation has been run you can examine the catches by gear.

## Reaching steady state

Once the `MizerParams` object has been properly setup, it may be the case that one wishes put the system in steady state. Sometimes this can be done simply by running the model using `project` until it reaches steady state. However, this method is 
not sure to work, and there is a function called `steady()` that is more reliable. The function `steady()` must be supplied with a MizerParams object. It takes that MizerParams object, looks at the initial system state, computes the levels of reproduction of the different species, hold them fixed, and evolves the system until a steady state is reached (or more precisely, until the amount that the population abundances change during a time-step is below some small tolerance level). After this, the reproductive efficiency of each species is altered so that when the reproduction dynamics are turned back on (i.e., when we stop holding recruitment levels fixed), the values of the reproduction levels which we held the system fixed at will be realized. The steady function is not sure to converge, and the way it re-tunes the reproductive efficiency values may not be realistic, but the idea is to alter the other parameters in the system until `steady()` does arrive at a steady state with sensible reproductive efficiency values.

Now that we know how to create a multispecies model we shall discuss how to [run a multispecies model.](running_a_simulation.html)
