---
title: "Using ggplot2 and plotly with mizer"
output:
  html_document:
    toc: yes
    fig_width: 5
    fig_height: 5
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE)
```

## Introduction

In this tutorial we will use the 
[ggplot2 package](https://ggplot2.tidyverse.org/) and the  
[plotly package for R](https://plot.ly/r/) 
to visualise the results from mizer simulations.
```{r}
library(mizer)
library(ggplot2)
library(plotly)
```

Mizer provides several functions for calculating summaries of the mizer
simulation results, see `?summary_functions`. Many of these functions have
corresponding plotting functions, see `?plotting_functions`. However it is easy
to produce customised plots directly using `ggplot()` or `plot_ly()`. This gives
more flexibility than the built-in plotting functions. Also, you will
occasionally want to look at different quantities for which perhaps there is not
built-in plotting function. In those cases the examples you see below will
provide a useful blueprint.

Both ggplot2 and plotly works with data frames, and the convenient way to
manipulate data frames is the [dplyr package](https://dplyr.tidyverse.org/).
```{r}
library(dplyr)
```

We create a simple simulation that we will use for our examples below.
```{r}
params <- newMultispeciesParams(NS_species_params)
sim <- project(params, t_max = 10, t_save = 0.5, effort = 0)
```

## From arrays to data frames

Mizer likes to work with arrays indexed by species and time or size. For example
the built-in summary functions return such arrays. These arrays need to be
converted to data frames before they can be conveniently plotted with either
`ggplot()` or `plot_ly`. This conversion is achieved by the function `melt()`.

For example, the function `getBiomass()` returns a two-dimensional array
(matrix) with one dimension corresponding to the time and the second dimension
to the species.
```{r}
biomass <- getBiomass(sim)
str(biomass)
```

This array can be converted with the `melt()` function to a data frame that
contains one row for each entry in the array.
```{r}
biomass_frame <- melt(biomass)
str(biomass_frame)
```

## ggplot2 or plotly

In this form the information can be handed to `plot_ly()` and converted to a
line plot:
```{r warning = FALSE}
pp <- plot_ly(biomass_frame) %>% 
    add_lines(x = ~time, y = ~value, color = ~sp)
pp
```

We have specified that the time is plotted along the x axis, the value along the
y axis and that the different species are represented by the colours of the 
lines. Note the American spelling "color" required by plotly.

Alternatively we can do the same thing with `ggplot()`:
```{r}
pg <- ggplot(biomass_frame) +
    geom_line(aes(x = time, y = value, colour = sp))
pg
```

Notice the different syntax for the ggplot2 and for plotly packages. The
underlying ideas are similar: they are both implementations of the grammar of
graphics. I recommend learning ggplot2 first, and switching to plotly only when
it has clear advantages (in particular for animations, see below).

The main advantage of plotly, namely the interactivity of the resulting figures,
can be obtained also with the ggplot syntax by running the result of `ggplot()`
through the function `ggplotly()`:
```{r}
ggplotly(pg)
```
 
 Below we will always for each plot first give the ggplot code and then the 
 plotly code.
 
 
## Adding labels
We may want to add labels to the figure and to each of the axes. In ggplot this
is done with `labs()`.
```{r}
pg + labs(title = "Biomass plot",
          x = "Time [years]",
          y = "Biomass [g]")
```


In plotly we use the `layout()` function.
```{r warning=FALSE}
pp %>% layout(
    title = "Biomass plot",
    xaxis = list(
        title = "Time [years]"
    ),
    yaxis = list(
        title = "Biomass [g]"
    )
)
```

## Filtering out data

We can use the filter function to filter out some of the data. For example we
could select only the data for specific species:
```{r warning=FALSE}
two_species_biomass <- filter(biomass_frame, sp %in% c("Gurnard", "Herring"))
```
Now if we plot this reduced data frame we get
```{r}
ggplot(two_species_biomass) +
    geom_line(aes(x = time, y = value, color = sp))
```

In the above we first converted the array to a data frame with `melt()` and then
selected the data of interest with `filter()`. We could alternatively have
first selected only the desired entries in the array and then created the
data frame with `melt()` from the resulting smaller array:
```{r warning=FALSE}
nfr <- melt(getBiomass(sim)[, c("Gurnard", "Herring")])
ggplot(nfr) +
    geom_line(aes(x = time, y = value, color = sp))
```

The result looks almost identical, except that the colours associated to the
species have changed. 

## Specifying line colours

If we want to make sure the same species always has the
same colour, we can use the colours specified by the MizerParams object
```{r}
getColours(params)
```
To use these colours in ggplot we add `scale_colour_manual()`:
```{r}
ggplot(biomass_frame) +
    geom_line(aes(x = time, y = value, color = sp)) + 
    scale_colour_manual(values = getColours(params))
```

In plotly we add `colors = getColours(params)r` to the `add_lines()` command:
```{r}
plot_ly(biomass_frame) %>% 
    add_lines(x = ~time, y = ~value, color = ~sp,
              colors = getColours(params))
```

Again note the American spelling "colors" required by plotly.


## Plotting spectra

Of course mizer has the `plotSpectra()` function for plotting size spectra.
Again it is instructional to create such plots by hand.

We can access the abundance spectra of the species via `N(sim)`. This is a
three-dimensional array (time x species x size). Let us first look at the
abundance at the final time.
```{r}
final_n <- N(sim)[idxFinalT(sim), , , drop = FALSE]
```
The `drop = FALSE` means that the result will again be a 3 dimensional array.
```{r}
str(final_n)
```
We use the `melt()` function to convert this array into a data frame.
```{r}
nf <- melt(final_n)
```
This has created a data frame with 4 variables and one observation for each
of the `r nrow(nf)` entries in the `1` x `r dim(final_n)[2]` x
`rdim(final_n)[3]` matrix `final_n`.
```{r}
str(nf)
```
The first three variables take their values from the dimension names of the
array. Of course the time variable is the same for all observations, because we
selected these before creating the data frame. The fourth variable called
`value` is the value of the entry of the array, so the abundance density in our
case.

There are a lot of entries with value 0, which we are not really interested in,
so it makes sense to remove them:
```{r}
nf <- filter(nf, value > 0)
```
This leaves a data frame with only `r nrow(nf)` observations. 

We can send this data frame to ggplot and add a line for the spectrum
for each species, with a different colour for each, and specify that we want
both the x axis and the y axis to be on a logarithmic scale.
```{r warning=FALSE}
pg <- ggplot(nf) +
    geom_line(aes(x = w, y = value, color = sp)) +
    scale_x_log10() +
    scale_y_log10()
pg
```

The corresponding syntax for plotly is
```{r warning=FALSE}
p <- plot_ly(nf) %>%
    add_lines(x = ~w, y = ~value, color = ~sp) %>% 
    layout(xaxis = list(type = "log", exponentformat = "power"),
           yaxis = list(type = "log", exponentformat = "power"))
p
```
In the above we used the pipe operator `%>%` which feeds the return value of one
function into the first argument of the next function. 


## Including resource spectrum

We can include additional lines in the plot by merging several data frames. For
example, we can include another line for the resource spectrum. We first
convert also the resource abundance at the final time into a data frame and
filter out the zero values
```{r}
nf_pp <- melt(NResource(sim)[idxFinalT(sim), , drop = FALSE]) %>% 
    filter(value > 0)
```
This data frame only contains three variables, because it does not have the 
`sp` column specifying the species. We add this column with the value "Resource"
```{r}
nf_pp$sp <- "Resource"
```
Now this new data frame has the same columns as the data frame `nf` and the two
can be bound together
```{r}
nf <- rbind(nf, nf_pp)
```
Using this extended data frame gives the following plot:
```{r}
p <- ggplot(nf) +
    geom_line(aes(x = w, y = value, color = sp)) + 
    scale_colour_manual(values = getColours(params)) +
    scale_x_log10() +
    scale_y_log10()
p
```

Of course we could use the same data frame also with plotly.


## Limiting the axes

We might want to zoom in on the part that includes the fish. There are three
ways to achieve this. The first is to use `filter()` to filter out all the
rows in the data frame that have small w and then plot the resulting data frame
as usual:
```{r warning=FALSE}
nf %>%
    filter(w > 10^-4) %>% 
    ggplot() +
    geom_line(aes(x = w, y = value, color = sp)) + 
    scale_colour_manual(values = getColours(params)) +
    scale_x_log10() +
    scale_y_log10()
```

The second method is to specify limits for the axes. In ggplot this is done
by adding `limits` to the axis scales:
```{r warning=FALSE}
ggplot(nf) +
    geom_line(aes(x = w, y = value, color = sp)) + 
    scale_colour_manual(values = getColours(params)) +
    scale_x_log10(limits = c(10^-4, NA)) +
    scale_y_log10(limits = c(NA, 10^20))
```

The `NA` means that the existing limits are kept.

In plotly we specify the `range` as follows:
```{r warning=FALSE}
plot_ly(nf) %>%
    add_lines(x = ~w, y = ~value, color = ~sp,
              colours = getColours(params)) %>% 
    layout(xaxis = list(type = "log", exponentformat = "power",
                        range = c(-4, 4)),
           yaxis = list(type = "log", exponentformat = "power",
                        range = c(-14, 20)))
```

Note how in plotly the range is specified by giving the logarithm to base 10 of the limits.


## Animating spectra

Instead of picking out a specific time we can ask plotly to make an animation
showing the changing spectra over time. So we melt the entire `N(sim)` array and
use the time variable to specify the frames with the `frame = ~time` argument
to `add_lines()`:
```{r}
melt(N(sim)) %>% 
    filter(value > 0) %>% 
    plot_ly() %>% 
    add_lines(x = ~w, y = ~value, 
              color = ~sp, colors = getColours(params),
              frame = ~time,
              line = list(simplify = FALSE)) %>% 
    layout(xaxis = list(type = "log", exponentformat = "power"),
           yaxis = list(type = "log", exponentformat = "power"))
```

Note how this produces a smooth animation in spite of the fact that we saved
the abundances only once a year. That interpolation is facilitated by the
`line = list(simplify = FALSE)` argument.

The ggplot package does not provide a similarly convenient way of creating
animations. There is the gganimate package, but it is not nearly as convenient.


## Comparing simulations

We may also want to make plots contrasting the results of two different
simulations, for example with different fishing policies. To illustrate this
we create two simulations with different fishing effort:
```{r}
sim1 <- project(params, t_max = 10, t_save = 0.2, effort = 2)
sim2 <- project(params, t_max = 10, t_save = 0.2, effort = 4)
```

Let us look at a plot of the fishing yield against time. This is calculated
by the `getYield()` function, which returns
an array (time x species) that we can convert to a data frame
```{r}
yield1 <- melt(getYield(sim1))
yield2 <- melt(getYield(sim2))
```

Let's look at the plot of the yield from the first simulation:
```{r}
ggplot(yield1) +
    geom_line(aes(x = time, y = value, colour = sp))
```

To make the plot less cluttered, we keep only the 4 most important species
```{r}
yield1 <- filter(yield1, sp %in% c("Saithe", "Cod", "Haddock", "N.pout"))
yield2 <- filter(yield2, sp %in% c("Saithe", "Cod", "Haddock", "N.pout"))
```
and plot again
```{r warning=FALSE}
p1 <- ggplot(yield1) +
    geom_line(aes(x = time, y = value, colour = sp))
p1
```

For simulation 2 the plot looks like this:
```{r warning=FALSE}
p2 <- ggplot(yield2) +
    geom_line(aes(x = time, y = value, colour = sp))
p2
```

Comparison will be easier if we combine these two plots.
For that we add an extra variable to the data frames that allow us to 
distinguish them and then we merge them together.
```{r warning=FALSE}
yield1$effort = as.factor(2)
yield2$effort = as.factor(4)
yield = rbind(yield1, yield2)
```

In ggplot we can now use `facet_grid()` to put the plot for each value of
`effort` side-by-side:
```{r}
ggplot(yield) +
    geom_line(aes(x = time, y = value, 
                  colour = sp)) +
    facet_grid(cols = vars(effort))
```

Or we can use the `linetype` aesthetic to represent the different `effort`
values by different line types:
```{r}
ggplot(yield) +
    geom_line(aes(x = time, y = value, 
                  colour = sp,
                  linetype = effort))
```

An alternative is to use the interactive comparison widgets from the
vdiffr package. For that purpose we need to create the two plots with the
same limits on the axes.
```{r warning=FALSE}
p1 <- ggplot(yield1) +
    geom_line(aes(x = time, y = value, colour = sp)) +
    ylim(0,1.75e12)
p2 <- ggplot(yield2) +
    geom_line(aes(x = time, y = value, colour = sp)) +
    ylim(0,1.75e12)
```
We can then create a widget that allows the user to toggle between the two
plots by clicking on it,
```{r}
vdiffr::widget_toggle(p1, p2)
```

or by sliding across it:
```{r}
vdiffr::widget_slide(p1, p2)
```

plotly is less good at faceting, but it can put arbitrary plots
side-by-side (or arrange them into a grid) with `subplot()`
```{r warning=FALSE}
subplot(p1, p2, shareX = TRUE, shareY = TRUE)
```

Note how the `shareY = TRUE` argument to `suplot()` makes sure the two plots
use the same scale on the y axis, and similarly for `shareX = TRUE`.


Also in plotly we can now tie the `effort` variable to the line type.
```{r warning=FALSE}
plot_ly(yield) %>% 
    add_lines(x = ~time, y = ~value, 
              color = ~sp, linetype = ~effort)
```

Arguably, ggplot2 does a nicer job in this case.
