---
title: "Comparing the First- and Second-Order Consumer Time Steppers"
output:
  html_document:
    toc: yes
    fig_width: 6
    fig_height: 4
vignette: >
  %\VignetteIndexEntry{Comparing the First- and Second-Order Consumer Time Steppers}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(mizer)
```

This vignette compares the current first-order consumer density update,
`project_n()`, with the predictor-corrector update, `project_n_2()`, on the
North Sea example model `NS_params`.

The comparison is intentionally modest. It is meant as a reproducible smoke
test for speed and time-step sensitivity, not as a comprehensive benchmark.

# Projection Helpers

The public `project()` function currently uses `project_n()`. To compare the
two consumer updates directly, we define a small projection loop modelled on
`project_simple()`, with the consumer update as the only switch.

For this comparison we call `project_n()` for the old method even when
diffusion is zero. This isolates the time-stepping comparison with
`project_n_2()`; the production `project()` function may use the separate
`project_n_no_diffusion()` shortcut when diffusion is absent.

```{r}
project_with_consumer_step <- function(params, dt, t_max,
                                       method = c("euler", "predictor_corrector"),
                                       effort = params@initial_effort) {
    method <- match.arg(method)
    steps <- as.integer(round(t_max / dt))
    dt <- t_max / steps

    n <- params@initial_n
    n_pp <- params@initial_n_pp
    n_other <- params@initial_n_other
    t <- 0

    no_sp <- nrow(params@species_params)
    no_w <- length(params@w)
    idx <- 2:no_w
    w_min_idx_array_ref <- (params@w_min_idx - 1) * no_sp + seq_len(no_sp)

    a <- matrix(0, nrow = no_sp, ncol = no_w)
    b <- matrix(0, nrow = no_sp, ncol = no_w)
    c <- matrix(0, nrow = no_sp, ncol = no_w)
    S <- matrix(0, nrow = no_sp, ncol = no_w)

    resource_dynamics_fn <- get(params@resource_dynamics)
    other_dynamics_fns <- lapply(params@other_dynamics, get)
    rates_fns <- lapply(params@rates_funcs, get)

    for (step in seq_len(steps)) {
        rates <- rates_fns$Rates(
            params,
            n = n, n_pp = n_pp, n_other = n_other,
            t = t, effort = effort, rates_fns = rates_fns
        )

        n_other_new <- list()
        for (component in names(params@other_dynamics)) {
            n_other_new[[component]] <- other_dynamics_fns[[component]](
                params,
                n = n, n_pp = n_pp, n_other = n_other,
                rates = rates, t = t, dt = dt, component = component
            )
        }

        n_pp_new <- resource_dynamics_fn(
            params,
            n = n, n_pp = n_pp, n_other = n_other,
            rates = rates, t = t, dt = dt,
            resource_rate = params@rr_pp,
            resource_capacity = params@cc_pp
        )

        if (method == "euler") {
            n_new <- project_n(
                params, rates, n, dt, a, b, c, S, idx,
                w_min_idx_array_ref, no_sp, no_w
            )
        } else {
            n_new <- project_n_2(
                params, rates, n, dt, a, b, c, S, idx,
                w_min_idx_array_ref, no_sp, no_w,
                rates_fns = rates_fns,
                n_pp = n_pp_new,
                n_other = n_other_new,
                t = t,
                effort = effort
            )
        }

        n <- n_new
        n_pp <- n_pp_new
        n_other <- n_other_new
        t <- t + dt
    }

    list(n = n, n_pp = n_pp, n_other = n_other, t = t)
}

relative_l2_error <- function(x, reference) {
    sqrt(sum((x - reference)^2)) / sqrt(sum(reference^2))
}
```

# Accuracy

We compare final consumer spectra against a smaller-time-step reference
solution from the predictor-corrector method. Because both methods use the same
spatial discretisation, this test is only about the time discretisation error
on the fixed `NS_params` weight grid.

```{r accuracy}
t_max <- 16
dt_values <- c(0.4, 0.2, 0.1, 0.05, 0.025)
reference_dt <- 0.2/2^7

params <- NS_params
initial_effort(params) <- 4

reference <- project_with_consumer_step(
    params,
    dt = reference_dt,
    t_max = t_max,
    method = "predictor_corrector"
)

accuracy <- do.call(rbind, lapply(dt_values, function(dt) {
    euler <- project_with_consumer_step(params, dt, t_max, method = "euler")
    pc <- project_with_consumer_step(params, dt, t_max,
                                     method = "predictor_corrector")

    data.frame(
        dt = dt,
        method = c("project_n", "project_n_2"),
        relative_error_n = c(
            relative_l2_error(euler$n, reference$n),
            relative_l2_error(pc$n, reference$n)
        ),
        relative_error_n_pp = c(
            relative_l2_error(euler$n_pp, reference$n_pp),
            relative_l2_error(pc$n_pp, reference$n_pp)
        )
    )
}))

accuracy
```

The consumer spectrum error is the primary comparison here. The resource error
is included as a useful diagnostic, but the resource update is not changed by
`project_n_2()`.

```{r}
plot(
    relative_error_n ~ dt,
    data = accuracy[accuracy$method == "project_n", ],
    log = "xy",
    type = "b",
    pch = 16,
    xlab = expression(Delta * t),
    ylab = "relative error in final consumer spectrum",
    ylim = range(accuracy$relative_error_n)
)
lines(
    relative_error_n ~ dt,
    data = accuracy[accuracy$method == "project_n_2", ],
    type = "b",
    pch = 17,
    col = 2
)
legend(
    "bottomright",
    legend = c("project_n", "project_n_2"),
    pch = c(16, 17),
    col = c(1, 2),
    lty = 1,
    bty = "n"
)
```

# Speed

The second-order method does more work per time step. In the full
predictor-corrector form it performs one predictor solve and then recalculates
the rates before doing the corrector solve.

The timings below use base R `system.time()`. They are not a replacement for a
proper benchmarking package, but they are enough to show the order of magnitude
of the overhead in this model.

```{r}
timing_reps <- 0

time_projection <- function(method, dt) {
    elapsed <- replicate(
        timing_reps,
        system.time(
            project_with_consumer_step(params, dt, t_max, method = method)
        )[["elapsed"]]
    )
    median(elapsed)
}

timings <- do.call(rbind, lapply(dt_values, function(dt) {
    data.frame(
        dt = dt,
        method = c("project_n", "project_n_2"),
        elapsed_seconds = c(
            time_projection("euler", dt),
            time_projection("predictor_corrector", dt)
        )
    )
}))

timings
```

```{r}
speed_summary <- merge(
    timings[timings$method == "project_n_2", c("dt", "elapsed_seconds")],
    timings[timings$method == "project_n", c("dt", "elapsed_seconds")],
    by = "dt",
    suffixes = c("_project_n_2", "_project_n")
)
speed_summary$slowdown <- with(
    speed_summary,
    elapsed_seconds_project_n_2 / elapsed_seconds_project_n
)
speed_summary
```

# Interpretation

On short runs the predictor-corrector method may not always show a clean
second-order convergence slope because the comparison is affected by the fixed
spatial grid, the first-order resource update, nonlinear rate feedbacks and
ordinary timing noise. The useful things to check are:

* whether `project_n_2()` reduces the consumer-spectrum error for a given
  `dt`;
* how much extra time the predictor and rate recalculation cost;
* whether the results remain finite and non-negative.

```{r}
final_check <- project_with_consumer_step(
    params,
    dt = min(dt_values),
    t_max = t_max,
    method = "predictor_corrector"
)

c(
    all_finite = all(is.finite(final_check$n)),
    all_non_negative = all(final_check$n >= 0)
)
```
