---
title: "Comparing the First- and Second-Order Consumer Time Steppers"
output:
  html_document:
    toc: yes
    fig_width: 6
    fig_height: 4
vignette: >
  %\VignetteIndexEntry{Comparing the First- and Second-Order Consumer Time Steppers}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
library(mizer)
```

This vignette compares the current first-order consumer density update,
selected with `method = "euler"`, with the predictor-corrector update selected
with `method = "predictor_corrector"`, on the North Sea example model
`NS_params`.

The comparison is intentionally modest. It is meant as a reproducible smoke
test for speed and time-step sensitivity, not as a comprehensive benchmark.

# Accuracy

We compare final consumer spectra against a smaller-time-step reference
solution from the predictor-corrector method. Because both methods use the same
spatial discretisation, this test is only about the time discretisation error
on the fixed `NS_params` weight grid.

```{r accuracy}
#| code-fold: true
#| output: true
t_max <- 8
dt_values <- c(0.4, 0.2, 0.1, 0.05, 0.025)
reference_dt <- 0.4 / 2^6

params <- NS_params
initial_effort(params) <- 4

relative_l2_error <- function(x, reference) {
    x_final <- finalN(x)
    reference_final <- finalN(reference)
    sqrt(sum((x_final - reference_final)^2)) / sqrt(sum(reference_final^2))
}

reference <- project(
    params,
    dt = reference_dt,
    t_max = t_max,
    method = "predictor_corrector"
)

accuracy <- do.call(rbind, lapply(dt_values, function(dt) {
    euler <- project(params, dt = dt, t_max = t_max, t_save = t_max,
                     method = "euler")
    pc <- project(params, dt = dt, t_max = t_max, t_save = t_max,
                  method = "predictor_corrector")

    data.frame(
        dt = dt,
        euler_error = relative_l2_error(euler, reference),
        predictor_corrector_error = relative_l2_error(pc, reference)
    )
}))

accuracy
```

The consumer spectrum error is the primary comparison here. The resource error
is included as a useful diagnostic, but the resource update is not changed by
the consumer time-stepper.

```{r accuracy-plot}
#| code-fold: true
#| output: true
plot(
    euler_error ~ dt,
    data = accuracy,
    log = "xy",
    type = "b",
    pch = 16,
    xlab = expression(Delta * t),
    ylab = "relative error in final consumer spectrum",
    ylim = range(accuracy$euler_error, accuracy$predictor_corrector_error)
)
lines(
    predictor_corrector_error ~ dt,
    data = accuracy,
    type = "b",
    pch = 17,
    col = 2
)
legend(
    "bottomright",
    legend = c('method = "euler"', 'method = "predictor_corrector"'),
    pch = c(16, 17),
    col = c(1, 2),
    lty = 1,
    bty = "n"
)
```

# Speed

The second-order method does roughly twice the work per time step. In the full
predictor-corrector form it performs one predictor solve and then recalculates
the rates before doing the corrector solve.

# Interpretation

On short runs the predictor-corrector method may not always show a clean
second-order convergence slope because the comparison is affected by the fixed
spatial grid, the first-order resource update, nonlinear rate feedbacks and
ordinary timing noise.

Because the predictor-corrector method is more expensive per time step, the
relevant comparison is between the Euler method at a given `dt` and the
predictor-corrector method at twice the `dt` so that both have similar run
times.
