---
title: "Running a Simulation"
output:
  html_document:
    toc: yes
    fig_width: 5
    fig_height: 5
---

```{r include=FALSE}
library(mizer)
```

## Introduction {#sec:projection}

In the [community](community_model.html) and [trait-based](trait_model.html)
models, we used the `project()` function to perform simple simulations where the
fishing effort was held constant throughout the duration of the simulation. In
the trait-based model example, we also looked at how the effort for different
gears could be specified. In this section we take a detailed look at how the
`project()` function works and the different ways in which effort and time can
be set up. This will allow us to run more general systems like the 
[multispecies model.](multispecies_model.html)

In `mizer`, simulations are performed using the `project()` function. This
function takes a `MizerParams` object and projects it forward through time,
starting from an initial population abundance and with a pre-determined fishing
effort pattern.

Running a projection with `project()` requires various arguments:

* A `MizerParams` object - The model parameters (see previous
	section);

* Fishing effort - The fishing effort of each gear through time;

* Time arguments - Arguments to control the time of the simulation,
	including the simulation time step, the length of the simulation
	and how frequently the output is stored.

* Initial population - The initial abundances of the stocks and the
	resource spectrum;

The help page for `project()` describes the arguments in more detail.

The `MizerParams` class was explored in the previous section. In this section we
will look at the other arguments and use examples to perform some simple
projections.



## The time arguments {#sec:time_arguments}

There are four arguments that control time in the `project()` function:
`dt`, `t_max` and `t_save` and `t_start`. All of them have default values.

* `t_max` determines the maximum time of the simulation (in years), i.e. how
  long the projection is run for. The default value for `t_max` is 100.
* `dt` is the time step used by the numerical solver in `project()`. The smaller
  the value, the longer the model will take to run. Sometimes it is necessary to
  use a small value to avoid numerical instabilities. The default value is 0.1.
* `t_save` sets the interval (in years) at which `project()` stores the state of
  the model in the resulting `MizerSim` object. For example, if `t_save` = 2, the
  state of the model is stored at t = 0, 2, 4... etc. `t_save` must be a multiple
  of `dt`. The default value of `t_save` is 1.
* `t_start` is the year in which the simulation starts. This is only needed if
  you want to label the simulation results by the actual year rather than by the
  number of years that have elapsed since the start of the simulation.

## Setting the fishing effort

The fishing effort argument describes the effort of the fishing gears in the
model through time. We have already seen that information on the fishing gears
and their selectivities and catchabilities is stored in the `MizerParams`
object.

There are three ways of setting the fishing effort. Examples of all three can be
seen in 
[the section on projection examples.](running_a_simulation.html#sec:projection_examples)

The simplest way is by passing the `effort` argument as a single number. This
value is then used as the fishing effort by all of the gears at each time step
of the projection, i.e. fishing effort is constant throughout the simulation and
is the same for all gears. We have seen this method in the community and
trait-based model sections above. The length of the simulation is determined by
the `t_max` argument.

The second method for setting the fishing effort is to use a numeric vector that
has the same length as the number of gears. The values in the vector are used as
the fishing effort of each gear at each time step, i.e. again, the fishing
effort is constant through time but now each gear can have a different constant
effort. The effort vector must be named and the names must be the same as the
gears in the `MizerParams` object. Again, the length of the simulation is
determined by the `t_max` argument.

Finally, the most sophisticated way of setting the fishing effort is to use a
two-dimensional array or matrix of values, set up as time step by gear. Each row
of the array has the effort values of each fishing gear at a particular time
time. The array must have dimension names. The names of the first dimension (the
row names) are the times. The steps between the times can be greater than the
`dt` argument but the times must be increasing. The names of the second
dimension (the column names) must match the names of the gears in the
`MizerParams` object used in the projection.

It is not necessary to supply a `t_max` argument when the effort is specified as
an array because the maximum time of the simulation is taken from the dimension
names. If a value for `t_max` is also supplied it is ignored.

## Setting the initial population abundance {#sec:setting_initial_abundances}

When running a simulation with the `project()` function, the initial populations
of the species and the resource spectrum need to be specified. Default values
for the initial values are automatically set when creating a MizerParams object.
However you can change these by assigning new values to `initialN()` and
`initialNResource()`. `initialN()` should be set to a matrix (with dimensions
species x size) that contains the initial abundances of each species at size
(the sizes must match those in the species size spectrum). `initialNResource` to
a vector of the same length as the the length of the full spectrum.


## Projection examples {#sec:projection_examples}

In this section we'll look at how to run simulations with the `project()`
function. The examples will focus on how fishing effort can be specified in
different ways. The results of the simulations will not be explored in detail.
We will leave that for
[the section on exploring the simulation results.](exploring_the_simulation_results.html#).


\subsection{Summary functions for MizerSim objects}
\label{sec:summaries}

Remember that the fishing mortality by size on a species is the product of the
selectivity, the catchability and the effort of the gear that caught it. We have
not specified any catchability values in the species parameter data.frame so the
default value of 1 is used. The selectivity ranges between 0 and 1. This means
that in these examples the fishing mortality of a fully selected species is
given by the effort of the gear that catches it.

### Projections with single, simple constant effort

When we use a single value for the `effort` argument, the value is used as a
constant effort for all the gears. This method can be particularly useful for
quickly projecting forward without fishing (you just set the `effort` argument
to 0).

We will use a MizerParams object with four gears similar to the one that we created in 
[the section with examples of making a MizerParams objects](multispecies_model.html#sec:params_example)
This example MizerParams object is called `NS_params` and is shipped with the mizer package.
Let's have a look at its summary:
```{r}
summary(NS_params)
```


As well as thinking about the `effort` argument we also need to consider the
time parameters. We will project the populations forward until time equals 10
(`t_max` = 10), with a time step of 0.1 (`dt` = 0.1), saving the output every
time step (`t_save` = 1). We use a constant effort value of 1.0.

```{r label=siple_projection_single_effort}
sim <- project(NS_params, effort = 1, t_max = 10, dt = 0.1, t_save = 1)
```


The resulting `sim` object is of class `MizerSim`. At this point we won't
explore how the results can be investigated in detail. However, we will use the
basic summary plot that you have seen before:

```{r label=plot_basic_ms_sim}
plot(sim)
```

The big difference between this multispecies model and the trait-based model can
be seen in the range of predation mortality and feeding level values. With the
trait-based model all the *species* had the same predation mortality and feeding
level patterns. Here the species all have different patterns, driven by their
differing life history characteristics and the heterogeneous interaction matrix.

You can also see in the above figure that each species has different fishing
selectivity (see the fishing mortality panel). Remember that the default setting
for the fishing gears is a knife-edge gear where the knife-edge is positioned at
the species `w_mat` parameter.

The effort through time can be inspected with `effort()` (we use the `head()`
function to just show the first few lines).
```{r label=show_effort}
head(effort(sim))
```
The `effort` slot shows the effort by time and gear. In this example, we
specified the `effort` argument as a single numeric of value 1. As you can see
this results in the same effort being used for all gears for all time steps.

A `summary()` method is also available for objects of type `MizerSim`. This is
essentially the same as the summary for `MizerParams` objects, but includes
information on the simulation time parameters.

```{r summary_mizersim,eval=FALSE}
summary(sim)
```

If we decrease `t_save` but keep `t_max` the same then we can see that the time
dimension of the `effort` array changes accordingly. This will also be true of
the simulation results returned by `N()` and `NResource()`. Here we reduce `t_save`
to 0.5, meaning that the effort and abundance information is stored at t = 1.0,
1.5, 2.0 etc.

```{r label=simple_proj_tsave05}
sim <- project(NS_params, effort = 1, t_max = 10, dt = 0.1, t_save = 0.5)
head(effort(sim))
```

### Setting constant effort for different gears {#sec:proj_diff_gears_constant_effort}

As mentioned above, we can also set the effort values for each gear separately
using a vector of effort values. This still keeps the efforts constant through
time but it means that each gear can have a different constant effort.

We need to create a named vector of effort, where the names match the gears. For
example, here we want to switch off the industrial gear (i.e. effort = 0), keep
the pelagic gear effort at 1, set the effort of the beam trawl gears to 0.3 and
the effort of the otter trawl gear to 0.7. We set the `effort` like this:

```{r label=set_effort_vector}
effort <- c(Industrial = 0, Pelagic = 1, Beam = 0.3, Otter = 0.7)
```

We then call `project()` with this effort and inspect the resulting `effort`
matrix (again we use the `head()` function to just show the first few lines):

```{r label=project_effort_vector}
sim <- project(NS_params, effort = effort, t_max = 10, dt = 1, t_save = 1)
head(effort(sim))
```

You can see that the effort for each gear is constant but each gear has the
effort that was specified in the `effort` argument.

This impact of this can be seen plotting the fishing mortality. There is a
dedicated plot, `plotFMort()`, that shows the fishing mortality at size for each
species at a particular time step (the default is the final time step). The
fishing mortality on each of the species is determined by the effort of the gear
that caught it.

```{r label=plotFMort, fig.keep='last'}
plotFMort(sim)
```


### An example of changing effort through time

In this example we set up a more complicated fishing effort structure that
allows the fishing effort of each gear to change through time. As mentioned
above, to do this, effort must be supplied as a two dimensional array or matrix.
The first dimension is time and the second dimension is gear. The dimensions
must be named. The gear names must match the gears in the `MizerParams` object.
Also, as mentioned above, if effort is passed in as an array then the length of
the simulation is determined by the time dimension names and the argument
`t_max` is not used. Instead the simulation runs from the earliest time in the
effort array to the latest.

In this example, we will project forward from time $t=1$ to time $t=10$. The
effort of the industrial gear is held constant at 0.5, the effort of the pelagic
gear is increased linearly from 1 to 2, the effort of the beam trawl decreases
linearly from 1 to 0, whilst the effort of the otter trawl decreases linearly
from 1 to 0.5.

First we create the empty `effort` array:
```{r label=set_empty_effort_array}
gear_names <- c("Industrial","Pelagic","Beam","Otter")
times <- seq(from = 1, to = 10, by = 1)
effort_array <- array(NA, dim = c(length(times), length(gear_names)),
    dimnames = list(time = times, gear = gear_names))
```

Then we fill it up, one gear at a time, making heavy use of the `seq()` function to create a sequence:

```{r label=fill_effort_array}
effort_array[,"Industrial"] <- 0.5
effort_array[,"Pelagic"] <- seq(from = 1, to = 2, length = length(times))
effort_array[,"Beam"] <- seq(from = 1, to = 0, length = length(times))
effort_array[,"Otter"] <- seq(from = 1, to = 0.5, length = length(times))
```

The first few rows of the effort array are shown as an illustration:
```{r label=show_effort_array}
head(effort_array)
```
The first row gives the effort between times 1 and 2, the second between times 2
and 3, and so on. The time 10 row in the array is not actually used, except to
set the final time for the simulation.

Now we can use this effort array in the projection:

```{r label=project_effort_array}
sim <- project(NS_params, effort = effort_array, dt = 0.1, t_save = 1)
head(sim@effort)
```

As you can see, it can be quite fiddly to set up a complicated effort array so
it may be easier to prepare it in advance as a .csv file and read it in, similar
to how we read in the interaction matrix in 
[the section with an examples of making a MizerParams object.](multispecies_model.html#sec:params_example) We
give an example of this in 
[the section on the North Sea model.](a_multispecies_model_of_the_north_sea.html)

Note that in this example we set up the effort array so that the effort was set
every whole time step (e.g. time = 1, 2, etc). This does not have to be the case
and it is possible to set the effort more frequently than that, e.g. at time =
1.0, 1.5, 2.0, 2.5 etc. The only restriction is that the difference between time
dimension names must be at least as big as the `dt` argument.

Now that we know how to run a simulation, the we are ready to learn how to 
[explore the simulation results.](exploring_the_simulation_results.html)
