---
title: "The Trait-Based Model"
output:
  html_document:
    toc: yes
    fig_width: 5
    fig_height: 5
---

```{r include=FALSE}
library(mizer)
```


## Introduction
The trait-based size spectrum model can be derived as a simplification of the
general model outline in the [model description](model_description.html)
section. It is more complicated than a [community model](community_model.html)
and the most significant difference between the two is that while the community
model aggregates all species into a single spectrum, the trait-based model resolves many
species. 

In a trait-based model the asymptotic size is considered to be the most
important *trait* characterizing a species. Most of the species-specific
parameters, such as $\beta$ and $\sigma$, are the same for all species. Other
model parameters are determined by the asymptotic size. For example, the weight
at maturation is a fixed fraction of the asymptotic size. The asymptotic sizes
of the species are spread evenly on a logarithmic scale. The number of species
is not important and does not affect the general dynamics of the model.

## Setting up a trait-based model {#sec:set_up_trait_model}

To help set up a trait-based model, there is a wrapper function,
`newTraitParams()`. Like the `newCommunityParams()` function described in the
section on the [community model](community_model.html), this function can take
many arguments. Most of them have default values so you don't need to worry
about them for the moment. See the help page for `newCommunityParams()` for more
details.

One of the key differences between the community type model and
the trait-based model is that reproduction and egg production are considered. In
the community model, reproduction is constant and there is no relationship
between the abundance in the community and egg production. In the trait-based
model, the egg production is dependent on mature individuals investing part
of their energy income into reproduction. The relationship between the energy
invested into reproduction and the actual rate of egg production is modelled 
using a *Beverton-Holt* type function (the
default in `mizer`, see [the section on density-dependence in reproduction](model_description.html#recruitment)) where the reproduction rate $R_i$ (numbers per time) approaches a maximum as the energy invested increases.

Here we set up the model to have 10 species, with asymptotic sizes ranging from 
10 g to 100 kg. All the other parameters have default values. 

```{r label=demo_trait_model_params}
params <- newTraitParams(no_sp = 10, min_w_inf = 10, max_w_inf = 1e5)
```

The output reminds the user that $\gamma$ was not specified, and so will be
calculated using other parameters.

This function returns an object of type `MizerParams` that holds all the model
information, including species parameters. This object can therefore be
interrogated in the same way as described in 
[the section on the community model](community_model.html).

```{r label=summary_comm_params2}
summary(params)
```


The summary shows us that now we have 10 species in the model, with asymptotic
sizes ranging from $`r min(species_params(params)$w_inf)`$ to $`r
max(species_params(params)$w_inf)`$. The rather strange-looking values for the
sizes is due to the fact that the size classes are equally spaced on a
logarithmic scale.

The size at maturity (`w_mat`) is linearly related to the asymptotic size. Each
species has the same preferred predator-prey mass ratio parameter values (`beta`
and `sigma`, see 
[the section on predator/prey mass ratio](model_description.html#sec:ppmr)). 
There are $`r length(w(params))`$ size
bins in the community and $`r length(w_full(params))`$ size bins including the
resource spectrum. Ignore the summary section on fishing gear for the moment.
This is explained later.

## Running the trait-based model {#sec:running_trait_model}

As with the [community model](community_model.html), we can project the
trait-based model through time using the `project()` function. Here we project
the model for 75 years without any fishing (the `effort` argument is set to 0).
We use the default initial population abundances so there is no need to pass in
any initial population values (see 
[the section on setting the initial abundances](running_a_simulation.html#sec:setting_initial_abundances)).


```{r label=trait_project_no_fishing}
sim <- project(params, t_max = 75, effort = 0)
```

This results in a `MizerSim` object that contains the abundances of the
community and resource spectra through time, as well as the original
`MizerParams` object. As with the community model, we can get a quick overview
of the results of the simulation by calling the `plot()` method:

```{r plot_comm_sim_2}
plot(sim)
```


The summary plot has the same panels as the one generated by the community
model, but here you can see that all the species in the community are plotted.
The panels show the situation in the final time step of the simulation, apart
from the biomass through time plot. As this is a trait-based model where all
species fully interact with each other, the predation mortality and feeding
level by size is the same for each species. In this simulation we turned fishing
off so the fishing mortality is 0. The size-spectra show the abundances at size
to be evenly spaced by log of asymptotic size.

## Example of a trophic cascade with the trait-based model {#sec:trophic_cascade_trait_model}

As with the community model, it is possible to use the trait-based model to
simulate a trophic cascade. Again, we perform two simulations, one with fishing
and one without. We therefore need to consider how fishing gears and selectivity
have been set up by the `newTraitParams()` function.

The default fishing selectivity function is a knife-edge function, which only
selects individuals larger than 1000 g. There is also only one fishing gear in
operation, and this selects all of the species. You can see this if you call the
`summary()` method on the `params` argument we set up above. At the bottom of
the summary there is a section on *Fishing gear details*. You can see that there
is only one gear, called `knife_edge_gear` and that it selects species 1 to 10.
To control the size at which individuals are selected there is a
`knife_edge_size` argument to the `newTraitParams()` function. This has a
default value of 1000 g.

In `mizer` it is possible to include more than one fishing gear in the model and
for different species to be caught by different gears. We will ignore this for
now, but will explore it further below when we introduce an industrial fishery
to the trait-based model.

To set up the trait-based model to have fishing we set up the `MizerParams`
object in exactly the same way as we did before but here the `knife_edge_size`
argument is explicitly passed in for clarity:

```{r label=set_trait_fishing}
params_knife <- newTraitParams(no_sp = 10, min_w_inf = 10, max_w_inf = 1e5,
    knife_edge_size = 1000)
```

First we perform a simulation without fishing in the same way we did above by
setting the `effort` argument to 0:

```{r label=sim_trait_no_fish}
sim0 <- project(params_knife, effort = 0, t_max = 75)
```

Now we simulate with fishing. Here, we use an effort of 0.75. As mentioned in
[the section on trophic cascades in the community model](community_model.html#sec:trophic_cascade_comm_model), 
the fishing
mortality on a species is calculated as the product of effort, catchability and
selectivity (see 
[the section on fishing gears](multispecies_model.html#sec:fishing_gear) 
for more details). Selectivity
ranges between 0 (not selected) and 1 (fully selected). The default value of
catchability is 1. Therefore, in this simulation the fishing mortality of a
fully selected individual is simply equal to the effort. This effort is constant
throughout the duration of the simulation (however, mizer does allow variable
effort).

```{r label=sim_trait_with_fish}
sim1 <- project(params_knife, effort = 0.75, t_max = 75)
```

Again, we can plot the summary of the fished community using the default
`plot()` function. The knife-edge selectivity at 1000 g can be clearly seen in
the fishing mortality panel:

```{r label=plot_trait_fmort}
plot(sim1)
```

The trophic cascade can be explored by comparing the total abundances of all
species at size when the community is fished and unfished. As mentioned above,
we obtain the abundances with  `N(sim)`, which returns a three dimensional array
with dimensions time x species x size. Here we have 76 time steps (75 from the
simulation plus one which stores the initial population), 10 species and 100
sizes:

```{r label=trait_dim_n}
dim(N(sim0))
```

As with the community model, we are interested in the relative total abundances
by size in the final time step so we use the `finalN()` function. This gives us
a matrix with one row per species and one column per size bin. We sum in each
column to get a vector with the total abundance per size bin:

```{r label = total_abund_trait}
total_abund0 <- colSums(finalN(sim0))
total_abund1 <- colSums(finalN(sim1))
```

We can then use these vectors to calculate the relative abundances:

```{r label='relative_comm_abundance2'}
relative_abundance <- total_abund1 / total_abund0
```

This can be plotted using the commands below:

```{r label=plot_relative_comm_abund2}
plot(x = w(params), y = relative_abundance, log = "xy", type = "n", 
     xlab = "Size (g)", ylab = "Relative abundance", ylim = c(0.1, 10))
lines(x = w(params), y = relative_abundance)
lines(x = c(min(w(params)), max(w(params))), y = c(1, 1), lty = 2)
```

The impact of fishing on species larger than 1000 g can be clearly seen. The
fishing pressure lowers the abundance of large fish ($> 1000$ g). This then
relieves the predation pressure on their smaller prey (the preferred
predator-prey size ratio is given by the $\beta$ parameter, which is set to 100
by default), leading to an increase in their abundance. This in turn increases
the predation mortality on their smaller prey, which reduces their abundance and
so on.

This impact can also be seen by looking at the predation mortality by size. The
predation mortalities are retrieved using the `getPredMort()` function. As
mentioned above, for the trait based model the predation mortality by size is
the same for each species. Therefore we only look at the predation mortality of
the first species.

```{r label='get_M2_trait'}
m2_no_fishing <- getPredMort(params, finalN(sim0))[1, ]
m2_with_fishing <- getPredMort(params, finalN(sim1))[1, ]
```

The predation mortalities can then be plotted.

```{r label=plot_relative_trait_m2, fig.cap="Predation mortalities from the unfished (dashed line) and fished (solid line) trait-based model."}
plot(x = w(params), y = m2_with_fishing, log = "x", type = "n", 
     xlab = "Size [g]", ylab = "Predation Mortality [1/year]")
lines(x = w(params), y = m2_no_fishing, lty = 2)
lines(x = w(params), y = m2_with_fishing)
```


## Setting up an industrial fishing gear {#sec:trait_additional_gear}

In this section we want to operate an *industrial* fishery. Industrial
fishing targets the small zooplanktivorous species that are typically used for
fishmeal production.

In the previous simulations we had only one fishing gear and it targeted all the
species in the community. This gear had a knife-edge selectivity that only
selected species larger than 1 kg. We can see that when we look at the gear
parameters
```{r}
gear_params(params)
```

We will expand the model to include multiple fishing gears.
This requires us to look more closely at how fishing gears are handled in
`mizer`. In `mizer` it is possible for a fishing gear to catch only a subset of
the species in the model. This is useful because when running a simulation with
`project()` you can specify the effort per gear and so you can turn gears on or
off as you want. Each gear has a selectivity curve for each species. 

We will set up the model to include two fishing gears: an `industrial` gear that
only catches species with an asymptotic size less than or equal to 500g, and a
second gear, `other`, that catches everything else. The position of the
knife-edge for both gears will occur at 0.05 x the asymptotic size i.e. the
selectivity parameters will be different for each species and will depend on the
asymptotic size.

For this we will need to change the `gear_params` data frame. If we want to
keep the original model, we should first make a copy before making 
modifications.
```{r}
params_multi_gear <- params
```

To start with we need to know what the asymptotic sizes of the species in the
model are so we can determine the knife-edge positions for each species. These
are stored in the `w_inf` column of the `species_params` data frame inside the
`params` object. We can then use these asymptotic sizes to set a vector of
knife edge sizes that are 0.05 times the asymptotic size:

```{r label=calc_knife_edge_trait}
gear_params(params_multi_gear)$knife_edge_size <- 
  species_params(params)$w_inf * 0.05
```

Now we want to assign each species to either the *industrial* or *other* gear.

```{r label=trait_industrial_gear_names}
no_sp <- 10
gear <- rep("Industrial", no_sp)
gear[species_params(params)$w_inf > 500] <- "Other"
gear_params(params_multi_gear)$gear <- gear
```
To check what has just happened let us look at the new gear parameter data
frame:
```{r label=looking_at_trait_species_params}
gear_params(params_multi_gear)
```

Having created our `MizerParams` object with multiple gears, we can now turn our
attention to running a projection with multiple gears. In our previous examples
of calling `project()` we have specified the fishing effort with the `effort`
argument using a single value. This fixes the fishing effort for all gears in
the model, for all time steps. We can do this with our multi-gear parameter
object:

```{r label=project_multi_gear_trait_single_effort}
sim_multi_gear <- project(params_multi_gear, t_max = 75, effort = 0.5)
```

By plotting this you can see that the fishing mortality for each species now has a different selectivity pattern, and that the position of the selectivity knife-edge is given by the asymptotic size of the species.

```{r label=plot_multi_gear_trait_single_effort}
plot(sim_multi_gear)
```

For the industrial fishery we said that we only wanted species with an
asymptotic size of 500 g or less to be fished. There are several ways of
specifying the `effort` argument for `project()` . Above we specified a single
value that was used for all gears, for all time steps. It is also possible to
specify a separate effort for each gear that will be used for all time steps. To
do this we pass in effort as a named vector. Here we set the effort for the
*Industrial* gear to 0.75, and the effort of the *Other* gear to 0 (effectively
turning it off).

```{r label=project_trait_multigear}
sim_multi_gear <- project(params_multi_gear, t_max = 75,
    effort = c(Industrial = 0.75, Other = 0))
```

Now you can see that the *Industrial* gear has been operating and that fishing
mortality for species larger than 500 g is 0.

```{r label=plot_multi_gear_trait}
plot(sim_multi_gear)
```


## The impact of industrial fishing

In the previous section we set up and ran a model in which an industrial fishery
was operating that only selected smaller species. We can now answer the
question: what is the impact of such a fishery? We can again compare abundances
of the fished (`sim_industrial1`) and unfished (`sim_industrial0`) cases:

```{r label = industrial_fishery_simulation}
sim_industrial0 <- project(params_multi_gear, t_max = 75, effort = 0)
sim_industrial1 <- project(params_multi_gear, t_max = 75,
    effort = c(Industrial = 0.75, Other = 0))
total_abund0 <- apply(finalN(sim_industrial0), 2, sum)
total_abund1 <- apply(finalN(sim_industrial1), 2, sum)
relative_abundance <- total_abund1 / total_abund0
```

And plot the relative abundances:

```{r label=plot_relative_comm_abund_industrial}
plot(x = w(params), y = relative_abundance, log = "xy", type = "n", 
     xlab = "Size [g]", ylab="Relative abundance", ylim = c(0.1, 10))
lines(x = w(params), y = relative_abundance)
lines(x = c(min(w(params)),max(w(params))), y = c(1, 1), lty = 2)
```

This shows another trophic cascade, although this time one driven by fishing the
species at the midrange part of the spectrum, not the largest individuals as
before. This trophic cascade acts in both directions. The cascade upwards is
driven by the lack of food for predators leading to smaller realised maximum
sizes. The cascade downwards has the same mechanism as fishing on large fish, a
combination of predation mortality and food limitation.

The next section explains how to setup the more general 
[multispecies model.](multispecies_model.html)
