---
title: "Upgrading mizer"
output:
  html_document:
    toc: yes
    toc_float: yes
    fig_width: 5
    fig_height: 4
vignette: >
  %\VignetteIndexEntry{Upgrading mizer}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    eval = FALSE
)
library(mizer)
```

This article collects the changes that may require you to update your own code
or models when you upgrade to a new release of mizer. Existing model objects
created with an earlier version continue to load and run — they are upgraded
automatically — so the notes below are about changes in *behaviour* and in the
functions you call, not about stored objects. The changes are grouped by the
release in which they took effect, most recent first.

# Setting resource parameters

Two related changes affect how you modify the resource size spectrum. Together
they make the resource scalars behave like the species parameters: a scalar is
an input, and the size-dependent arrays are computed from it.

## Assigning to `resource_params()` now updates the resource arrays

Previously, assigning to `resource_params()` — or to one of its components, such
as `resource_params(params)$kappa <- ...` — only stored the new scalar values.
The size-dependent carrying capacity (`cc_pp`) and replenishment rate (`rr_pp`)
were left unchanged until you next called `setResource()`.

Now these assignments immediately rebuild the arrays from the scalars, exactly as
`species_params()<-` rebuilds the species rates:

- `kappa`, `lambda` and `w_pp_cutoff` rebuild the carrying capacity;
- `r_pp` and `n` rebuild the replenishment rate.

Arrays that you have set by hand are left untouched (see *Frozen arrays* below).

If your code changed a resource scalar and then called `setResource()` to apply
it, nothing breaks — you can drop the now-redundant `setResource()` call. If you
changed a resource scalar and relied on the arrays *not* changing until later,
review that code.

## Assigning to `resource_params()` does not balance the resource

*Balancing* means adjusting the rate and capacity together so that the resource
replenishes at exactly the rate at which it is consumed, keeping it at its steady
state. Assigning to `resource_params()` rebuilds the arrays from the scalars but
does **not** balance, so the resource steady state generally shifts.

Balancing is now solely a feature of `setResource()`. To change a resource
coefficient *and* keep the resource balanced, call `setResource()` rather than
assigning to `resource_params()`:

```{r}
# Rebuild the capacity from a new coefficient and rebalance the rate,
# so the steady state is preserved:
params <- setResource(params, resource_capacity = new_kappa)

# Likewise, set a new rate coefficient and rebalance the capacity:
params <- setResource(params, resource_rate = new_r_pp)
```

## The resource setters gained a `balance` argument

`resource_rate<-`, `resource_capacity<-`, `resource_level<-` and
`resource_dynamics<-` still balance by default (unchanged behaviour), but they
now accept a `balance` argument so you can switch balancing off:

```{r}
# Set the capacity but leave the rate untouched (do not rebalance):
resource_capacity(params, balance = FALSE) <- my_capacity
```

## Frozen arrays are protected from incidental balancing

When you set `rr_pp` or `cc_pp` by hand (by assigning a full array rather than a
scalar), mizer marks it "set manually" — it is *frozen* and will not be
recomputed from the resource parameters. Previously, an operation that
re-balanced the resource *without* being given a replacement rate or capacity —
for example changing only `resource_dynamics`, or calling `setResource()` with
neither a rate nor a capacity — would silently overwrite such a frozen array. It
is now kept, and a warning is issued instead.

To deliberately recompute a frozen array from the resource parameters, pass
`reset = TRUE`. An *explicit* balance — where you hand `setResource()` a new rate
or capacity — still overrides the freeze; pass `balance = FALSE` if you want the
frozen array kept in that case.
