---
title: "Upgrading mizer"
output:
  html_document:
    toc: yes
    toc_float: yes
    fig_width: 5
    fig_height: 4
vignette: >
  %\VignetteIndexEntry{Upgrading mizer}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    eval = FALSE
)
library(mizer)
```

This article collects the changes that may require you to update your own code
or models when you upgrade to a new release of mizer. Existing model objects
created with an earlier version continue to load and run — they are upgraded
automatically — so the notes below are about changes in *behaviour* and in the
functions you call, not about stored objects. The changes are grouped by the
release in which they took effect, most recent first.

Only changes that can alter the behaviour of *existing* code are listed. The
many purely additive features (new functions, new optional arguments, new
plots) are described in the [changelog](https://sizespectrum.org/mizer/news/index.html)
and are not repeated here.

# Upgrading from mizer 3.1 to 3.2

## Setting resource parameters

Two related changes affect how you modify the resource size spectrum. Together
they make the resource scalars behave like the species parameters: a scalar is
an input, and the size-dependent arrays are computed from it.

### Assigning to `resource_params()` now updates the resource arrays

Previously, assigning to `resource_params()` — or to one of its components, such
as `resource_params(params)$kappa <- ...` — only stored the new scalar values.
The size-dependent carrying capacity (`cc_pp`) and replenishment rate (`rr_pp`)
were left unchanged until you next called `setResource()`.

Now these assignments immediately rebuild the arrays from the scalars, exactly as
`species_params()<-` rebuilds the species rates:

- `kappa`, `lambda` and `w_pp_cutoff` rebuild the carrying capacity;
- `r_pp` and `n` rebuild the replenishment rate.

Arrays that you have set by hand are left untouched (see *Frozen arrays* below).

If your code changed a resource scalar and then called `setResource()` to apply
it, nothing breaks — you can drop the now-redundant `setResource()` call. If you
changed a resource scalar and relied on the arrays *not* changing until later,
review that code.

### Assigning to `resource_params()` does not balance the resource

*Balancing* means adjusting the rate and capacity together so that the resource
replenishes at exactly the rate at which it is consumed, keeping it at its steady
state. Assigning to `resource_params()` rebuilds the arrays from the scalars but
does **not** balance, so the resource steady state generally shifts.

Balancing is now solely a feature of `setResource()`. To change a resource
coefficient *and* keep the resource balanced, call `setResource()` rather than
assigning to `resource_params()`:

```{r}
# Rebuild the capacity from a new coefficient and rebalance the rate,
# so the steady state is preserved:
params <- setResource(params, resource_capacity = new_kappa)

# Likewise, set a new rate coefficient and rebalance the capacity:
params <- setResource(params, resource_rate = new_r_pp)
```

### The resource setters gained a `balance` argument

`resource_rate<-`, `resource_capacity<-`, `resource_level<-` and
`resource_dynamics<-` still balance by default (unchanged behaviour), but they
now accept a `balance` argument so you can switch balancing off:

```{r}
# Set the capacity but leave the rate untouched (do not rebalance):
resource_capacity(params, balance = FALSE) <- my_capacity
```

### Frozen arrays are protected from incidental balancing

When you set the size dependence of the resource capacity or the resource rate
by hand (by assigning a full vector rather than a scalar), mizer marks it "set
manually" — it is *frozen* and will not be recomputed from the resource
parameters. Previously, an operation that re-balanced the resource *without*
being given a replacement rate or capacity — for example changing only
`resource_dynamics`, or calling `setResource()` with neither a rate nor a
capacity — would silently overwrite such a frozen array. It is now kept, and a
warning is issued instead.

To deliberately recompute a frozen array from the resource parameters, pass
`reset = TRUE` to `setResource()`.

## The `species_params` data frame is now an S3 subclass

The `species_params` data frame now has class `c("species_params",
"data.frame")` (and `gear_params` similarly). It behaves like an ordinary data
frame, but subsetting and subassignment go through class-preserving S3 methods
and can trigger reactive re-validation and conversions (for example filling in a
weight from a length). Code that relied on `class(species_params(params))` being
exactly `"data.frame"`, or that stripped attributes with the assumption of a
plain data frame, may need adjusting. When you need a plain frame, coerce
explicitly with `as.data.frame()`.

## Accessing a column with `$` now returns a named vector

Extracting a single column from a `species_params` or `gear_params` object with
`$` now returns a vector named by species (or by `"species, gear"` for
`gear_params`):

```{r}
species_params(params)$w_mat
#>   Sprat  Herring      Cod
#>    ...      ...      ...
```

The values are unchanged, but the names are new. This is convenient for
identifying entries, but code that compared such a vector with `identical()` to
an unnamed vector, or that used it as-is where names matter (for example as
row/column names elsewhere), may behave differently. Strip the names with
`unname()` if you need the old behaviour. The `species` column itself is
returned unnamed.

## Setting `sel_func` adds the required argument columns

Assigning a selectivity function name to a `gear_params` object now
automatically adds the argument columns that the function needs (as `NA`),
ready to be filled in:

```{r}
gp$sel_func <- "sigmoid_length"
# gp now has l25 and l50 columns, both NA
```

Previously these columns had to be added by hand. Code that checks which columns
are present in `gear_params`, or that expected setting `sel_func` to leave the
column set unchanged, will now see the extra columns (#431).

## Passing a data frame to `species_params()` / `given_species_params()` now validates it

Calling `species_params()` or `given_species_params()` on a plain data frame now
runs the same validation and defaults that `validSpeciesParams()` and
`validGivenSpeciesParams()` apply, rather than only checking for misspellings and
converting lengths to weights. `species_params(df)` fills in the default columns
(`w_max`, `alpha`, `n`, `p`, `interaction_resource`, `z_ext`, and the rest), and
`given_species_params(df)` applies the consistency corrections (for example
clamping `w_mat` below `w_inf`), derives `w_inf` from `w_max`/`w_repro_max` when
it is absent, and now stops if the frame has duplicate species rows. Models built
or modified through `newMultispeciesParams()`, `setParams()` and the
`species_params()<-` / `given_species_params()<-` setters are unaffected, because
those already ran this validation. Only code that called the two accessors
directly on a bare data frame will see the extra columns and stricter checks
(#432).

## Printing of mizer array objects shows the values

`print()` on the array objects returned by the rate getters (`ArraySpeciesBySize`,
`ArrayTimeBySpecies`, `ArrayResourceBySize`, `ArrayTimeByResourceBySize` and
`ArrayTimeBySpeciesBySize`, as returned by `getEncounter()`, `getBiomass()`,
`getFMort()`, `NResource()` and similar) now truncates the output instead of
flooding the console with all the array entries. If your code or reports relied
on the old printed format, use `as.data.frame()` to go back to the full output.

## Upper boundary condition at `w_max`

The size-spectrum solver now holds the abundance at zero above each species'
maximum size `w_max`. Without diffusion this happens automatically and results
are unchanged. With diffusion switched on this change stops a small amount of
density leaking to sizes above `w_max`, so results there change slightly. See
`vignette("numerical_details")`.

## Extension packages: dynamic marker classes

If you develop a mizer extension, an installed extension package is now
recognised as a dispatching extension from the S3 methods it registers for its
marker class (for example `getEncounter.mizerMR`), rather than only from a
statically defined S4 marker class. You can now omit the static
`setClass("mizerFoo", contains = "MizerParams")` and let mizer create the marker
class dynamically. This lets two independently developed extensions be chained
in either load order. See `vignette("creating-extension-packages")`.

# Upgrading from mizer 3.0 to 3.1

Version 3.1 leaves default results unchanged from 3.0 unless you opt in to the
new experimental second-order-in-size scheme. The changes below can still affect
existing code in specific situations.

## Maximum-size species parameters clarified

The maximum-size parameters have been given clearer, separate roles (#325):

- `w_inf`, the von Bertalanffy asymptotic size, is now the primary maximum-size
  parameter and is used as the default for `w_repro_max` (the size at which a
  mature individual invests all its energy in reproduction) and for `w_mat`.
- `w_max` is now purely a computational boundary — it sets the size grid and the
  plot range — and defaults to `1.5 * w_inf`.
- The default external mortality parameter `z0` is now computed from `w_inf`
  rather than `w_max`, so the computational boundary `w_max` no longer feeds into
  any model parameter.

Existing models and scripts are unaffected: if `w_inf` is not supplied it is
taken from `w_repro_max` or `w_max`, so old objects behave as before. However,
**new models built from the defaults may differ from 3.0.0**. If you build models
from scratch, check that `w_inf`, `w_max` and `w_repro_max` mean what you intend.

## `getTrophicLevel()` gives the resource a size-dependent trophic level

`getTrophicLevel()` and `getTrophicLevelBySpecies()` now assign the resource a
size-dependent trophic level,
$T_R(w) = \max(1,\, 1 + \log(w / w_R) / \log(\beta_R))$, instead of treating the
resource as trophic level 0. The new `w_R` and `beta_R` arguments control this.
Trophic levels computed with these functions will therefore be higher than
before. Set the arguments explicitly if you need to reproduce old numbers.

## Bug fixes that change results

Several fixes correct earlier behaviour and so change output:

- **`summary()` of a `MizerSim`** now reports the fishing effort that was used
  during the simulation, rather than the model's `initial_effort`. Gears whose
  effort varied over time show the mean, flagged with a note giving the range.
  The printed summary therefore differs for simulations run with time-varying
  effort.
- **`MizerSim` method for `plotDiet()`** introduced in version 3.0 simply
  plotted the diet at the initial time of the simulation. Now `plotDiet()` for
  a `MizerSim` accepts a `time_range` argument. The diet is now computed from
  the *simulated* abundances at the requested times, defaulting to the final
  saved step, rather than the initial one (#357).
- **Other components and `t_save`.** `project()` was advancing the abundances of
  other components (set via `setComponent()`) only once per *saved* time step
  instead of once per `dt` step. They are now integrated with the same `dt` as
  the consumer and resource spectra, so results for models with other components
  no longer depend on `t_save`.
- **Time-varying effort in `getRDI()`, `getRDD()`, `getFlux()`.** On a
  `MizerSim` object these now use the simulated time-varying effort rather than
  the initial effort, so they change for simulations with varying effort (#370).
- **`plotCDF()` / `plotlyCDF()` bin placement.** Each cumulative value is now
  plotted at its bin's *upper* edge, correcting a one-bin offset.
  The curves shift by one bin compared with 3.0 (#383).
- **`distanceMaxRelRDI()`.** Now returns `Inf` instead of `NaN` when a previous
  RDI is zero, so `projectToSteady()` no longer mistakes a `NaN` distance for
  convergence. Convergence behaviour can therefore differ in edge cases.

## Second-order methods advance the resource at the midpoint

If you use `project()` with `method = "predictor_corrector"` (or the new
`method = "tr_bdf2"`), the resource and the other components are now advanced
with midpoint rates rather than the start-of-step value, so that they reach the
same second-order accuracy in time as the consumer spectra. Results from these
methods therefore differ slightly from 3.0. The default `method = "euler"` and
the steady states are unchanged.

## Opting in to the second-order-in-size scheme

3.1 adds an optional, experimental second-order-accurate finite-volume scheme in
the size variable, controlled by the new `second_order_w` slot. It is **off by
default**, so default results are byte-identical to 3.0. If you switch it on
(via `second_order_w()<-` or the `second_order_w` argument of the `new...Params()`
constructors), size-integrated diagnostics and the resource spectrum shift by
$O(\Delta w)$, so a calibrated model may need recalibrating. See `?second_order_w`
and the "Numerical Details" vignette.

# Upgrading from mizer 2.5.4 to 3.0

Version 3.0 is a large release. Most new capabilities are additive and off by
default, but there are several renamed arguments, deprecations and behavioural
changes that can affect existing code.

## Renamed arguments and changed defaults (breaking changes)

- **First argument of `plotBiomass()`, `plotYield()`, `plotYieldGear()`** (and
  their `MizerSim` methods and `plotly*` wrappers) is renamed from `sim` to
  `object`, for consistency with the other plot generics. Calls that passed the
  simulation by name, `plotBiomass(sim = my_sim)`, must become
  `plotBiomass(object = my_sim)`. Positional calls are unaffected.
- **`plotBiomassObservedVsModel()` / `plotlyBiomassObservedVsModel()`** now
  default to `ratio = FALSE` for all object types. Calls that relied on the
  previous ratio plot must set `ratio = TRUE` explicitly.
- **`plotDiet()` no longer accepts a `time_range` argument.** Remove it from your
  calls. (In 3.1 a `time_range` argument returns for the `MizerSim` method — see
  above.)
- **Dimnames of `getMort()` and `getPredRate()`** arrays are now `sp` and `w`
  (matching `getFMort()` and the other rate getters). Code that referred to the
  old dimnames by name must be updated.

## Rate getters return classed array objects

Functions that return arrays of the form (species × size), (time × species) or
(time × species × size) now attach extra attributes and an S3 class
(`ArraySpeciesBySize`, `ArrayTimeBySpecies` or `ArrayTimeBySpeciesBySize`). The
numeric values and ordinary matrix behaviour (arithmetic, subsetting) are
unchanged, but the extra class and attributes mean that a strict comparison such
as `identical(getMort(params), old_value)` can now report a difference where the
numbers agree. Use `unclass()`, or compare with `all.equal()` on the values, if
you need to ignore the class. These objects also carry `print()`, `summary()`,
`plot()` and `as.data.frame()` methods, so printing them looks different from a
bare matrix.

## `setInitialValues()` is deprecated

`setInitialValues()` is deprecated. Replace

```{r}
params <- setInitialValues(params, sim)
```

with

```{r}
params <- finalParams(sim)
```

or, when averaging over a time range, with
`getParams(sim, time_range, geometric_mean)`. This reflects a shift in
interpretation: a `MizerParams` object now represents not just the model
specification but also its current state (the abundances), which can be
extracted from a simulation with `getParams()`, `finalParams()` and
`initialParams()`.

## Growth can no longer be negative

Growth is now forced to be non-negative, preventing unphysical shrinkage. In any
model where the energy available for growth used to go negative (for example a
strongly food-limited large individual), growth is now clamped at zero instead,
so projected size spectra can differ from 2.5.4. No warning is issued when growth
stops at or after the maturity size.

## `project()` timing and effort handling

- **Inherited `dt` and `method`.** When `project()` is called on an existing
  `MizerSim` object, `dt` and `method` now default to the values stored in the
  simulation's new `sim_params` slot. If you pass values that differ from the
  stored ones, a warning is issued. To use different settings deliberately, pass
  them explicitly and expect the warning.
- **`t_max` / `t_save` with an effort array.** These arguments are now respected
  even when an effort array is supplied (#231). With `t_max` the simulation
  extends beyond the times in the effort array using the last known effort; with
  `t_save` the save frequency is controlled independently, interpolating effort
  as needed. Simulations that previously derived their length or save times
  solely from the effort array may now produce a different set of saved steps.
- **State at `t_max` always saved.** `project()` now warns when `t_max` is not a
  multiple of `t_save` and ensures the state at `t_max` is saved even if the
  final interval is shorter than `t_save` (#341). The returned simulation may
  therefore contain one extra saved time step compared with 3.0.

## `plot()` and `summary()` are now S3 methods

The `plot()` and `summary()` methods for `MizerParams`, `MizerSim` and the mizer
array classes are now registered as S3 methods rather than S4 methods, so
`plot()` and `summary()` stay plain S3 generics when mizer is loaded. This avoids
interfering with S4 dispatch in other packages, but code that relied on
`plot`/`summary` being S4 generics (for example via `selectMethod()` or
`getMethod()`) needs adjusting.

## Bug fixes that change results

- **`getMeanMaxWeight()`** now applies the species selector to the denominator as
  well, so its values change when a subset of species is selected.
- **`plotSpectra()` axis limits.** It no longer forces the y-axis lower limit to
  `1e-20` (it auto-scales to the data) and, when `resource = FALSE`, it uses
  `min(params@w)` rather than `min(params@w) / 100` as the default lower size
  limit. Plots therefore look different.
- **`getFMort()` on a `MizerSim`** was silently dropping the component names from
  `n_other`, breaking rate functions that access `n_other` by name (e.g.
  `n_other[["resource"]]`); it now preserves them.
- **`getFMort.MizerSim()`** now passes the time argument `t` to user-defined
  fishing-mortality functions, so a time-dependent fishing function now sees the
  correct time.

## Predation diffusion is available but off by default

3.0 adds a diffusion term to the growth dynamics, controlled by the new
`use_predation_diffusion` slot. It defaults to `FALSE`, preserving the behaviour
of earlier mizer, so existing models are unchanged unless you switch it on with
`use_predation_diffusion(params) <- TRUE`. Likewise the new species parameters
`z_ext`, `d`, `E_ext` and `D_ext` for external mortality, encounter and diffusion
all default to values that leave the model unchanged.
