---
title: "Building a new multispecies model, part 1"
author: "Gustav Delius"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r}
install.packages("devtools")
devtools::install_github("sizespectrum/mizer", ref = "Lysekill")
```

The purpose of this course is for all of us to share our experiences and 
insights into the mizer model and to all contribute to the future plans for
the platform.

My role is to introduce you to a few of the features of the next version of
mizer and in particular to show you how to use them to build mizer models from
scratch more easily than may have been possible in the past. 

The way the practical sessions are organised is that I run through these
R notebooks while you are watching and give my running commentary. Then I hand
over to you and you will do further explorations, possibly building on the
examples I have run through. During that time I hope there will be a lot of
discussion and of helping each other.

The place to find out about the next version of mizer is 
https://sizespectrum.org/mizer/dev/. Note the "dev" in the URL. If it is
missing, you are looking at the old version. You can also identify the pages
of the development version by the red version number in the navigation bar.

As you browse the website, you will frequently come across mistakes or unclear
sections. You would do us all a big favour if you could correct the mistakes
and improve the explanations. Most pages have an "Edit this page" link 
underneath the page title. This will allow you to edit the page straight in
your browser and contribute your changes via a pull request on GitHub.

We now start going through some of the recent changes in mizer. But first let's
make sure `mizer` is loaded. We also load `tidyverse`, because we find it convenient.
```{r}
library("mizer")
library("tidyverse")
```

As you know, the specification of a mizer model is stored in an object of 
class MizerParams, which has many slots holding the various rate functions
and other parameters. In this notebook we will always use the name `params`
for the variable holding our MizerParams object. This object now has slots
`params@initial_n`, `params@initial_n_pp` and `params@initial_n_other` holding the
initial values for the fish abundances, the resource abundances and the
unstructured resource biomasses, respectively. (The unstructured resources are
another new feature, very much in the experimental stage, that we will discuss
later.)

To illustrate this, let us create a scaling model.
```{r}
params <- set_scaling_model(
    n = 2/3,
    q = 3/4,
    no_w = 100
)
```
Many of the mizer plot functions that used to work only for MizerSim objects
(the objects holding the results of a simulation run) now also work with
MizerParams objects, and plot the initial state stored in the params object.
```{r}
plotSpectra(params)
```

We see that the model consists of 11 species, with each species having a size-
distribution that is just a translated version of the others. The axes are on
log scales, so a translation in the graph corresponds to a scaling in the
variables. This motivates the name "scaling model". It is a trait-based model
of the kind that Ken will discuss in his lectures.

The `plotSpectra()` function also has a new parameter `power` that can be used
to look at the size spectra in the different ways already discussed by Ken.
With `power = 2` you get the Sheldon spectrum, because the function then plots
the abundance density multiplied by size to the power 2.
```{r}
plotSpectra(params, power = 2)
```

If you are wondering what motivates the drop-off in the resource abundance and
then the low abundance of the species: let's plot also the total abundance.
```{r}
plotSpectra(params, power = 2, total = TRUE)
```

We see that the abundances were chosen so that the sum of resource and fish
abundances produce the Sheldon power-law spectrum, at least initially. If
there were an infinite number of species continuing to the right, then they
would sum up to the power law everywhere. As it is, the total abundance drops
off.

The larger species also have fewer predators than the smaller ones and the
largest species have no predation mortality at all. The scaling model
compensates for this by putting a larger background mortality on the larger
individuals so that the total mortality follows the power law. Thus these
species experience a power-law of prey and hence a power-law of growth as
well as a power-law death. This is why we can write down an analytic solution
for their steady-state abundances, which is what is used as the initial 
condition.

However, due to the fact that there is only a discrete number of species and
a discrete set of weight classes, there are some discretisation errors and the
analytic solution is not exactly the solution of the numerical mizer model.
To find the later, we can use the function `steady()` that runs the 
model dynamics until the system is close to steady.
```{r}
params <- steady(params)
```
We see that it did not take long to reach the steady state, which is because
we started off so close to it already. 

Note how the `steady()` function again returns a MizerParams object, just with
updated initial values.
```{r}
plotSpectra(params, power = 2)
```

Let's zoom in on the interesting region.
```{r}
plotSpectra(params, power = 2, total = TRUE, 
            ylim = c(1e-4, NA), wlim = c(1e-1, NA))
```

Now over to you. Play with the parameters of the `set_scaling_model()` function,
in particular the exponents `n` and `q`, and see if the effects on the community
conform with your expectations. Does the system always reach a nice steady
state?

In case you have not worked in an R notebook before: you can insert a new code
chunk by clicking the *Insert Chunk* button on the toolbar or by pressing
*Ctrl+Alt+I*. When you click the *Run* button or press *Ctrl+Enter*, the code
in the chunk is executed and the results displayed underneath.

