---
title: "Building a new multispecies model, part 3"
author: "Gustav Delius"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("mizer")
library("tidyverse")
```

We now consider the case where we know some life-history parameters of some
important species in our ecosystem, and want to study these embedded in a full
set of background species. We use a trait-based model for the background species
and then add our foreground species with their specified parameters.

Let's look at hake and mullet for example for our foreground species. Let's
assume we know the following parameters for them:
```{r}
species_params <- tribble(
    ~species, ~w_inf, ~w_mat, ~beta, ~sigma, ~a,      ~b,    ~k_vb, ~l50, ~l25,
    "Hake",   10470,  250,    11,    1.1,    0.00667, 3.035, 0.178, 16.6, 16,
    "Mullet", 250,    16.5,   283,   1.8,    0.00624, 3.15,  0.6,   15.5, 13.2
)
species_params$sel_func <- "sigmoid_length"
```
Here `w_inf` is the asymptotic size, `w_mat` the maturity size, `beta` and
`sigma` the parameters of the log-normal predation kernel, `a` and `b` the
parameter in the length to weight conversion $w = a\, l^b$, `k_vb` the von
Bertalanffy growth curve parameter and `l50` and `l25` the parameters of the
`sigmoid_length()` selectivity function.

In practice you would probably not type them into R as above but create a 
spreadsheet with the parameters and then read that in with `read_csv()`.

We now start by creating a scaling model for the background species.
```{r}
params <- set_scaling_model(
    min_w_pp = 10^-12,
    no_sp = 16, 
    min_w_inf = 1, 
    max_w_inf = 10^6,
    min_egg = 10^-5,
    min_w_mat = 10^-0.6,
    w_pp_cutoff = 2 * 10^-0.6,
    no_w = 200,
    bmort_prop = 0.1) %>% 
    steady(t_max = 100) %>% 
    markBackground()
```
Note how we used the pipe operator `%>%` to pipe the result of `set_scaling_model()` into `steady()` and then the result from that into `markBackground()`. The mizer functions are pipe-friendly in that their first
parameter is always the MizerParams object or the MizerSim object.

The call to `markBackground()` just specifies that these are background species.
This has the effect for example that they are grey when plotted:
```{r}
plotSpectra(params, power = 2, total = TRUE, 
            ylim = c(10^-6, NA), wlim = c(10^-4, NA))
```

and we will see other consequences below.

We now add our foreground species with the parameters specified in the data
frame `species_params`.
```{r}
params <- addSpecies(params, species_params)
```
```{r}
plotSpectra(params, power = 2, total = TRUE, 
            ylim = c(10^-6, NA), wlim = c(10^-4, NA))
```

We see that `addSpecies()` adds the new species at a very low abundance
compared to the background species. This is on purpose, because at that low
abundance there is very little backreaction from the background species, and
therefore we are in a regime where we understand the steady-state size
distribution of the species analytically. Therefore mizer was able to find
initial values for the size distributions that are close to steady state
values. We see that little changes as we run the system to steady state:
```{r}
params <- steady(params)
plotSpectra(params, power = 2, total = TRUE, 
            ylim = c(10^-6, NA), wlim = c(10^-4, NA))
```

We make a copy of this for later purposes.
```{r}
params2 <- params
```

Now we can increase the abundance of the background species.
```{r}
params <- rescaleAbundance(params, factor = 16)
plotSpectra(params, power = 2, total = TRUE, 
            ylim = c(10^-6, NA), wlim = c(10^-4, NA))
```
Note how `rescaleAbundance()` knew to only rescale the foreground species and
leave the background species alone. Note also, that the introduction of the
foreground species has led to a bulge in the total community spectrum. However
the abundance of the background species was originally chosen so that the
community abundance would be close to the Sheldon power-law spectrum. We
therefore retune their abundance so that again the community spectrum is as
close to the Sheldon power-law as possible.
```{r}
params <- retuneBackground(params)
plotSpectra(params, power = 2, total = TRUE, 
            ylim = c(10^-6, NA), wlim = c(10^-4, NA))
```

We see that as we introduce the foreground species, the background species
become less important. 

The above plot does not yet show a steady state. We still need to call
`steady()` to find the steady state.
```{r}
params <- steady(params)
plotSpectra(params, power = 2, total = TRUE, 
            ylim = c(10^-6, NA), wlim = c(10^-4, NA))
```

Oh dear! The steady state looks quite different. The background species have
increased a lot in their abundance. Perhaps the answer is to just retune their
abundances and run to steady state again.
```{r}
params <- params %>% retuneBackground() %>% steady()
plotSpectra(params, power = 2, total = TRUE, 
            ylim = c(10^-6, NA), wlim = c(10^-4, NA))
```

It is getting worse! Clearly this iteration is unstable. This is because we
changed the system by too big a step in one go when we rescaled the abundances
of the foreground species by a factor of 16. We need to do this in smaller 
steps.
```{r}
params <- params2 %>% 
    rescaleAbundance(factor = 2) %>% 
    retuneBackground() %>% 
    steady() %>% 
    rescaleAbundance(factor = 2) %>% 
    retuneBackground() %>% 
    steady() %>% 
    rescaleAbundance(factor = 2) %>% 
    retuneBackground() %>% 
    steady() %>% 
    rescaleAbundance(factor = 2) %>% 
    retuneBackground() %>% 
    steady()
plotSpectra(params, power = 2, total = TRUE, 
            ylim = c(10^-6, NA), wlim = c(10^-4, NA))
```

We can rescale the foreground species to whatever abundance we like. So we need
some observation to choose the right abundance. In our system we have for both
of our foreground species observations of the total biomass of all individuals above a certain cutoff size. We put that information into the species_params data frame.
```{r}
species_params(params)$biomass_observed <- NA
species_params(params)$cutoff_size <- NA
species_params(params)["Hake", "biomass_observed"] <- 0.013
species_params(params)["Hake", "cutoff_size"] <- 0.3
species_params(params)["Mullet", "biomass_observed"] <- 0.0018
species_params(params)["Mullet", "cutoff_size"] <- 0.25
```

Now I want to introduce a very convenient gadget that allows you to tune the
parameters of the system interactively. This is not yet very well 
documented, so I will demonstrate.
```{r message=FALSE, warning=FALSE}
params <- tuneParams(params)
```


